---
title: "Functions"
subtitle: "R for Data Science (2e), Chapter 25"
toc: true
theme: darkly
format: html
---

## Introduction

Functions allow automation of common tasks.

Advantages:

- Code is easier to understand with named functions.

- Updating code in less places, if needed.

- Eliminate chance of making mistakes when copy-and-pasting

- Improved reusability

**Rule of Thumb:**

Whenever block of code is copied and pasted twice - consider writing a function.

Three useful types of functions:

- Vector functions - Input: One or more vectors; Output: Vector

- Data frame functions - Input: Data frame; Output: Data frame

- Plot functions - Input: Data frame; Output: Plot

### Prerequisites

```{r}
#| label: Prereq

library(tidyverse)
library(nycflights13)
```

## Vector functions

**Input: One or more vectors; Output: Vector**

```{r}
#| label: vector functions intro
set.seed(1950)

df <- tibble(
  a = rnorm(5),
  b = rnorm(5),
  c = rnorm(5),
  d = rnorm(5),
)

df |>
  mutate(
    a = (a - min(a, na.rm = TRUE)) / (max(a, na.rm = TRUE) - min(a, na.rm = TRUE)),
    b = (b - min(b, na.rm = TRUE)) / (max(b, na.rm = TRUE) - min(b, na.rm = TRUE)),
    c = (c - min(c, na.rm = TRUE)) / (max(c, na.rm = TRUE) - min(c, na.rm = TRUE)),
    d = (d - min(d, na.rm = TRUE)) / (max(d, na.rm = TRUE) - min(d, na.rm = TRUE))
  )

# Fuction for rescaling values between 0 and 1 (Idea) 
rescale01 <- function(df){
  
  ncol_df = ncol(df)
  
  for(i in 1:ncol_df){
    df[i] = (df[i] - min(df[i], na.rm = TRUE)) / (max(df[i], na.rm = TRUE) - min(df[i], na.rm = TRUE))
  }
  
  return(df)
}

rescale01(df)

```

### Writing a function

Analyze code - what parts are constant and what parts vary. Above code:

```         
(a - min(a, na.rm = TRUE)) / (max(a, na.rm = TRUE) - min(a, na.rm = TRUE))
(b - min(b, na.rm = TRUE)) / (max(b, na.rm = TRUE) - min(b, na.rm = TRUE))
(c - min(c, na.rm = TRUE)) / (max(c, na.rm = TRUE) - min(c, na.rm = TRUE))
(d - min(d, na.rm = TRUE)) / (max(d, na.rm = TRUE) - min(d, na.rm = TRUE)) 
```

Replacing the parts with █:

```         
(█ - min(█, na.rm = TRUE)) / (max(█, na.rm = TRUE) - min(█, na.rm = TRUE))
```

Need three things:

1.  **Name** - here we'll use `rescale01`.
2.  **Arguments** - varying across calls, here jsut one - call it `x`.
3.  **Body** - code that's repeated across all calls.

Template for creating function:

```         
name <- function(arguments){
   body
}
```

Can write function as:

```{r}
#| label: function for rescaling book solution

rescale01 <- function(x){
   (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
} 

# Then can use mutate():
df <- tibble(
  a = rnorm(5),
  b = rnorm(5),
  c = rnorm(5),
  d = rnorm(5),
)

df |>
  mutate(
    a = rescale01(a),
    b = rescale01(b),
    c = rescale01(c),
    d = rescale01(d)
  )

```

See chapter 26 for `across()` - reducing duplication: `df |> mutate(across(a:d, rescale01))`.

### Improving function

Can improve previous function by using `range()`:

(see ?range: `range` returns vector containing minomum and maximum of all given arguments.)

```{r}
#| label: improving functions
set.seed(1950)


df <- tibble(
  a = rnorm(5),
  b = rnorm(5),
  c = rnorm(5),
  d = rnorm(5),
)

# Idea: Input = df, Output = df

rescale0_1 <- function(df){
  
  ncol_df <- ncol(df)
  
  for(i in 1:ncol_df){
    
    rng <- range(df[i], na.rm = TRUE, finite = TRUE) # ignoring infinite values
    
    df[i] = (df[i] - rng[1]) / (rng[2] - rng[1])
    
  }
  
  return(df)
  
}

rescale0_1(df)

# Idea: Single function for column-wise operation

rescale01 <- function(x){
  
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  
  x = (x - rng[1]) / (rng[2] - rng[1])
  
  return(x)
}

df |>
  mutate(
    a = rescale01(a),
    b = rescale01(b),
    c = rescale01(c),
    d = rescale01(d)
  )
    
  
  
```

### Mutate functions

"Mutate functions" - functions that work well inside `mutate()` and `filter()` - return output same length as input.

```{r}
#| label: Mutating functions

# Function computing z-score
z_score <- function(x){
  
  z = (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
  
  return(z)
  
}

# Function ensuring values of vector lie between minimum and maximum:
clamp <- function(x, min, max){
  
  case_when(
    x < min ~ min,
    x > max ~ max,
    .default = x
  )
  
}

clamp(1:10, min = 3, max = 7)
```

Can also du repeated string manipulation:

```{r}
#| label: Mutating functions - repeated string manipulation

first_upper <- function(x){
  
  str_sub(x, start = 1, end = 1) <- str_to_upper(str_sub(x, start = 1, end = 1))
  
  return(x)
  
}

first_upper("hello")

# str_to_sentence() works for single strings as well
```

Or removing percentege sign, commas, dollar sign, before converting to number:

```{r}
#| label: Mutating functions - Stripping metacharacters

clean_number <- function(x){
  
  is_pct <- str_detect(x, pattern = "%")
  
  num <- x |>
    str_remove_all(pattern = "%") |>
    str_remove_all(pattern = ",") |>
    str_remove_all(pattern = fixed("$")) |>
    as.numeric()
  
  if_else(is_pct, num / 100, num)
  
}

clean_number("$12,300")
clean_number("45%") # difference: function checks for percentages!
# Alternatively use readr::parse_number()

parse_number("$12,300")
parse_number("45%")
```

Example of function cleaning `NA`values:

```{r}
#| label: Function finding specific NA values

fix_na <- function(x){
  
  if_else(x %in% c(997, 998, 999), NA, x)
  
}
```

### Summary functions

Another family of vector functions is summary functions - functions that return single value for use in `summarize()`:

```{r}
#| label: Summary functions

# Example 1
commas <- function(x){
  
  str_flatten(string = x, collapse = ",",
              last = " and ")
  
}

commas(c("cat", "dog", "pigeon"))

# Alternatively use str_flatten_comma()
str_flatten_comma(string = c("cat", "dog", "pigeon"), last = " and ")

# Example 2
cv <- function(x, na.rm = TRUE){
  
  sd(x, na.rm = na.rm) / mean(x, na.rm = na.rm)
  
}

cv(runif(100, min = 0, max = 50))
cv(runif(100, min = 0, max = 500))

# Example 3
n_missing <- function(x){
  
  sum(is.na(x))
  
}

# Example 4 - Multiple vectors as input
mape <- function(actual, predicted){
  
  sum(abs((actual - predicted) / actual)) / length(actual)
  
}
```

### Exercises

```{r}
#| label: Exercises - vector functions

# Ex1.1
x <- c(1, 5, 2, NA)
y <- c(NA, 3, NA, 9)
z <- c(0, 7, 7, 7)

mean(is.na(x))
mean(is.na(y))
mean(is.na(z))

mean_n_missing <- function(x){
  
  mean(is.na(x))
  
}

mean_n_missing(x)
mean_n_missing(y)
mean_n_missing(z)

# Alternatively: Function returning vector of mean("number of NA")
df <- bind_cols(x = x, y = y, z = z)

mean_nmiss <- function(df){
  
  vec = vector(length = ncol(df))
  
  ncol = ncol(df)
  
  for(i in 1:ncol){
  
    vec[i] = mean(is.na(df[i]))  
  
  }
  return(vec)
}
mean_nmiss(df)


# Ex1.2 
x / sum(x, na.rm = TRUE)
y / sum(y, na.rm = TRUE)
z / sum(z, na.rm = TRUE)

prop <- function(x){
  
x / sum(x, na.rm = TRUE)  
  
}

prop(x)
prop(y)
prop(z)

# Alterntively:
df <- bind_cols(x = x, y = y, z = z)

prop_df <- function(df){
  
  matrix = matrix(nrow = nrow(df), ncol = ncol(df))
  
  
  for(i in 1:ncol(df)){
    
    vec = df[i] / sum(df[i], na.rm = TRUE)
    
    for(j in 1:nrow(vec)){
    
      matrix[j, i] <- vec[j, ]  
      
    }
    
  }
  
  return(matrix)
}

prop_df(df)

# Alterntively


prop_df1 <- function(df){
  
  m <- as_tibble(matrix(nrow = nrow(df), ncol = ncol(df)))
  
  
  for(i in 1:ncol(df)){
    
    vec <-  df[[i]] / sum(df[[i]], na.rm = TRUE)
    m[, i] <- vec
  }
  
  return(m)
}

prop_df1(df)


# Ex2
x <- c(1:10, Inf, -Inf)
x

rescale01a <- function(x){
  
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  
  z <- (x - rng[1]) / (rng[2] - rng[1])
  
  return(
    case_when(
    z == Inf ~ 1,
    z == -Inf ~ 0,
    .default = z
  )
)
}

rescale01a(x)

# Ex3
x <- c("2004-03-26", "1998-06-10", "1976-03-22", "1973-12-01")


ages <- function(v){
  
  age_string <- tibble(age = as.duration(today() - as_date(v)))
  
  separate_wider_regex(data = age_string, cols = age,
                       patterns = c(
                         ".*\\(",
                       age = "\\~[0-9]+\\.[0-9]+ [a-z]+",
                       "\\)"))
  
}

ages(x)

# Alternatively

library(lubridate)

ages <- function(x) {
  floor(time_length(interval(as_date(x), today()), "years"))
}

ages(x)

# Ex4
x <- runif(100)

# variance function
variance <- function(x){
  
  sum((x - mean(x, na.rm = TRUE))^2) / (length(x) - 1)
  
}

var(x)
variance(x)

# skewness function
skew <- function(x){
  
  n <- sum((x - mean(x, na.rm = TRUE))^3) / length(x)
  d <- (sum((x - mean(x, na.rm = TRUE))^2) / length(x))^(3/2)
  
  sk <- n/d
  return(sk)
}

skew(x)
DistributionUtils::skewness(x)


# Ex5
x <- c(1, 2, NA)
y <- c(1, NA, NA)

both_na <- function(x, y){
  
  posx <- which(is.na(x))
  posy <- which(is.na(y))
  
  length(intersect(posx, posy))
  
}

both_na(x, y)


# Ex6
# Returns a data frame with information about the file
# The function then extracts only the isdir column
is_directory <- function(x) {
  file.info(x)$isdir
}

# file.access() checks file permissions
# second argument specifies what permission to check
is_readable <- function(x) {
  file.access(x, 4) == 0
}

```

## Data frame functions

Useful for code repeated within dplyr verb - BUT will often repeat verbs themselves.

Copying and pasting multiple verbs, multiple times - useful to write data frame function - Input: Data frame; Output: Data frame (or vector).

### Indirection and tidy evaluation

`grouped_mean()` - computes the mean of `mean_var` grouped by `group_var`:

```{r}
#| label: Indirection and tidy evaluation
#| eval: false

grouped_mean <- function(df, group_var, mean_var){
  
  df |>
    group_by(group_var) |>
    summarize(mean(mean_var))
  
}

diamonds |>
  grouped_mean(cut, carat)

# Use made up data frame:

df <- tibble(
  mean_var = 1,
  group_var = "g",
  group = 1,
  x = 10,
  y = 100
)

df

df |> grouped_mean(group, x)
df |> grouped_mean(group, y)
# Issue: Does df |> group_by(group_var) |> summarize(mean(mean_var))
df |> group_by(group_var) |> summarize(mean(mean_var))
# Instead of: df |> group_by(group) |> summarize(mean(x))
df |> group_by(group) |> summarize(mean(x))
```

**What causes indirection?**

Dplyr uses **tidy evaluation** allowing to refer to names of variables inside data frame without any special treatment.

Need wy to tell `group_by()` and `summarize()` not to treat `group_var` and `mean_var` as the name of the variables, but instead look inside them for the variable we actually want to use.

**Solution** of tidy evaluation is called **embracing** - i.e., wrapping var in braces `{{ var }}` - tells dplyr to use variables stored inside the argument, not the argument as the literal variable name - makes dplyr look inside `var` rather than looking for variable called `var`.

```{r}
#| label: Embracing

df <- tibble(
  mean_var = 1,
  group_var = "g",
  group = 1,
  x = 10,
  y = 100
)

grouped_mean <- function(df, group_var, mean_var){
  df |>
    group_by({{ group_var }}) |>
    summarize({{ mean_var }})
}

df |> 
  grouped_mean(group, x)
```

### When to embrace?

**Key** - figuring out which arguments to "embrace"

$\rightarrow$ documentation.

Two most common sub-types of tidy evaluation:

- **Data-masking** - used in `arrange()`, `filter()`, `summarize()` - compute with variables

- **Tidy-selection** - used for `select()`, `relocate()`, `rename()`

E.g., see ?arrange(), the "..." argument.

### Common use cases

Performing same set of summaries in initial data exploration:

```{r}
#| label: Embracing common functions

summary6 <- function(data, var){
  
  data |>
    summarize(
      min = min({{ var }}, na.rm = TRUE),
      mean = mean({{ var }}, na.rm = TRUE),
      median = median({{ var }}, na.rm = TRUE),
      max = max({{ var }}, na.rm = TRUE),
      n = n(),
      n_miss = sum(is.na({{ var }})),
      .groups = "drop" # set "drop" to avoid message/leaving data in ungrouped stage
    )
  
}

diamonds |>
  summary6(carat)

# Compare to base R summary() function
summary(diamonds$carat)


# Can now use wrapped summarize() on grouped data as well:
diamonds |>
  group_by(cut) |>
  summary6(carat)

# Note: Arguments to summarize89 are data-masking, so is 'var' arg to summary6() - can also summarize computed variables:
diamonds |>
  group_by(cut) |>
  summary6(log10(carat))

```


```{r}
#| label: Embracing - example1
# Another popular summarize() helper function is version of count(), that computes proportions
count_prop <- function(df, var, sort = TRUE){ # default value for sort!
  df |>
    count({{ var }}, sort = sort) |>
    mutate(prop = n / sum(n))
}

diamonds |>
  count_prop(clarity)

```

Summarizing multiple variables requires `across()` (See chapter 26)

```{r}
#| label: Embracing - example2
# Function for finding sorted unique values of a variable for a subset of data:
unique_where <-  function(df, condition, var){
  
  df |> 
    filter({{ condition }}) |>
    distinct({{ var }}) |>
    arrange({{ var }})
  
}

flights |>
  unique_where(month == 12, dest)

# Or if always working with same data:
subset_flights <- function(rows, cols) {
  flights |> 
    filter({{ rows }}) |> 
    select(time_hour, carrier, flight, {{ cols }})
}

```


### Data- masking vs. tidy-selection

:::{.callout-important}
**Data-masking**

Functions - `group_by()`, `mutate()`, `filter()`, `summarise()`
$\rightarrow$ Every argument is interpreted as an expression referring to a column.

E.g., `group_by(year, month, day)` - Find column called `year`, then `month`, then `day`.


**Tidy-select**

Functions - `select()`, `across()`, `pick()`
$\rightarrow$ Don't evaluate expressions into vectors.

E.g., `pick(c(year, month, day))` - select these three columns.
:::

```{r}
#| label: Data-masking vs. tidy-selection
#| eval: false

count_missing <- function(df, group_vars, x_var){
  df |>
  group_by({{ group_vars }}) |>
  summarize(
    n_miss = sum(is.na({{ x_var}})),
    .groups = "drop"
  )
}

 flights |>
   count_missing(c(year, month, day), dep_time)
# Issue
 df |> group_by(c(year, month, day)) # Error: arg must be size 1
# See ?group_by also

# Solution pick(), see ?pick() -> using select semantics while inside a "data-masking" function like mutate() or summarize().
count_missing <- function(df, group_vars, x_var){
  
  df |>
    group_by(pick({{ group_vars }})) |>
    summarize(
      n_miss = sum(is.na({{ x_var}})),
      .groups = "drop"
    )
  
}

flights |>
  count_missing(c(year, month, day), dep_time)
  
```

Another useful use of `pick()` - making 2d table of counts:
```{r}
#| label: Useful function with pick()

count_wide <- function(data, rows, cols) {
  data |> 
    count(pick(c({{ rows }}, {{ cols }}))) |> 
    pivot_wider(
      names_from = {{ cols }}, 
      values_from = n,
      names_sort = TRUE,
      values_fill = 0
    )
}

diamonds |>
  count_wide(c(clarity, color), cut)

```

### Exercises
```{r}
#| label: Exercises

# Ex1.1. Finds all flights that were cancelled (i.e. is.na(arr_time)) or delayed by more than an hour.
filter_severe <- function(df, arr_time, dep_delay){
  
  df |>
    filter(is.na(arr_time) | dep_delay > 60)
  
}

flights |> filter_severe()

# Ex1.2: Counts the number of cancelled flights and the number of flights delayed by more than an hour.

summarize_severe <- function(df) {
  df |>
    summarize(
      n_cancelled = sum(is.na(arr_time)),
      n_delayed = sum(dep_delay > 60, na.rm = TRUE),
      .groups = "drop"
    )
}

flights |> group_by(dest) |> summarize_severe()

# Ex1.3: Finds all flights that were cancelled or delayed by more than a user supplied number of hours
filter_severe <- function(df, hours){
  
  df |>
    filter(is.na(arr_time) | dep_delay > hours)
  
}

flights |> filter_severe(hours = 2)

# Ex1.4: Summarizes the weather to compute the minimum, mean, and maximum, of a user supplied variable
summarize_weather <- function(df, temp){
  df |>
    summarize(
  min = min({{ temp }}, na.rm = TRUE),
  mean = mean({{ temp }}, na.rm = TRUE),
  max = max({{ temp }}, na.rm = TRUE)
    )
}

weather |>
  summarize_weather(humid)

# Ex1.5: Converts the user supplied variable that uses clock time (e.g., dep_time, arr_time, etc.) into a decimal time (i.e. hours + (minutes / 60))

standardize_time <- function(df, var_time){
  
  df |>
    mutate(dec_time = {{ var_time }} %/% 100 + ({{ var_time }} %% 100) / 60)
  
}

flights |> standardize_time(sched_dep_time)


# Ex2
# distinct() -> data-masking (...)
# count() -> data-masking(...; wt)
# group_by() -> data-masking(...)
# rename_with() -> tidy-select(...; .cols)
# slice_min() -> tidy-select(.by), data-masking(...; order_by; weight_by)
# slice_sample() -> tidy-select(.by), data-masking(...; order_by; weight_by)

# Ex3: Generalize the following function so that you can supply any number of variables to count:
count_prop <- function(df, var, sort = FALSE) {
  df |>
    count({{ var }}, sort = sort) |>
    mutate(prop = n / sum(n))
}
# use pick()
count_prop <- function(df, ..., sort = FALSE) {
  df |>
    count(..., sort = sort) |>
    mutate(prop = n / sum(n))
}

flights |> count_prop(dest, origin)
```


## Plot functions

Might want to return plot - `aes()` is a data-masking function:
```{r}
#| label: Functions for plotting

diamonds |>
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.1)

diamonds |>
  ggplot(mapping = aes(x = carat)) +
  geom_histogram(binwidt = 0.05)
```

Can compute a function - knowing that `aes()` is a data-masking function:
```{r}
#| label: Function for plotting - 2

histogram <- function(df, var, binwidth = NULL){
  
  df |>
    ggplot(
      mapping = aes(x = {{ var }})
    ) +
    geom_histogram(binwidth = binwidth)
  
}

diamonds |> histogram(carat, 0.1) +
  labs(x = "Size (in carats)", y = "Number of diamonds") # can still add additional components
```

### More variables

```{r}
#| label: Function for plotting w multiple variables

# Example
linearity_check <- function(df, x, y){
  df |>
  ggplot(
    mapping = aes(x = {{ x }}, y = {{ y }})) +
      geom_point() +
      geom_smooth(method = "loess", color = "red", se = FALSE) +
      geom_smooth(method = "lm", se = FALSE)
}

starwars |>
  filter(mass < 1000) |>
  linearity_check(mass, height)
```

Or alternative colored scatterplot for very large datasets where overplotting is a problem:
```{r}
#| label: Function for plots - overplotting

hex_plot <- function(df, x, y, z, bins = 20, fun = "mean") {
  df |> 
    ggplot(aes(x = {{ x }}, y = {{ y }}, z = {{ z }})) + 
    stat_summary_hex(
      aes(color = after_scale(fill)), # make border same color as fill
      bins = bins, 
      fun = fun,
    )
}

diamonds |> hex_plot(carat, price, depth)
```

### Combining with other tidyverse

```{r}
#| label: Useful helpers

# Example:  vertical bar chart - automatically sorting the bars in frequency order using fct_infreq()

sorted_bars <- function(df, var){
  df |>
    mutate({{ var }} := fct_rev(fct_infreq({{ var }}))) |> # ':=' evaluate left-hand side first, then use the resulting name
    ggplot(aes(y = {{ var }})) +
    geom_bar()
}

diamonds |> sorted_bars(clarity)


# Example: Making it easy to draw a bar plot just for a subset of the data
conditional_bars <- function(df, condition, var){
  df |>
    filter({{ condition }}) |>
    ggplot(aes(x = {{ var }})) + 
    geom_bar()
}

diamonds |> conditional_bars(cut == "Good", clarity)
```

See also https://gist.github.com/GShotwell/b19ef520b6d56f61a830fabb3454965b for more applications.

### Labeling

`rlang::englue()`  - works similar to `str_glue()` - any value wrapped in `{ }` will be inserted into the string - also understands `{{ }}`, which automatically inserts appropriate variable name:
```{r}
#| label: Labeling inside functions

histogram <- function(df, var, binwidth){
  label <- rlang::englue("A histogram of {{var}} with binwidth {binwidth}")
  
  df |>
    ggplot(aes(x = {{ var }})) +
    geom_histogram(binwidth = binwidth) +
    labs(title = label)
}

diamonds |> histogram(carat, 0.1)
```

