---
title: "Hierarchical data"
subtitle: "R for Data Science (2e), Chapter 23"
toc: true
theme: darkly
format: html
---

## Introduction

**Data rectangling** - taking data that is hierarchical, or tree-like, and converting it into rectangular data frame made up from rows and columns.

Two crucial functions:

- `tidyr::unnest_longer` and `tidyr::unnest_wider`

### Prerequisites

```{r}
#| label: Prereq

library(tidyverse)
library(repurrrsive) # Example datasets for reactangling practice
library(jsonlite) # reading JSON files into R lists
```

## Lists

simple vectors like integers, numbers, characters, date-times, and factors - homogeneous, i.e., every element is of the same data type.

Lists $\rightarrow$ stores elements of different types in the same vector - created by `list()`.

```{r}
#| label: Creating lists

x1 <- list(1:4, "a", TRUE)
x1

# Can name the components of a list, either by names(x) <- c("name") or:
x2 <- list(a = 1:2, b = 1:3, c = 1:4)
x2

# Useful alternative to display list - inspecting trough str():
str(x1)
str(x2)
```

## Hierarchy

```{r}
#| label: List within a nest
#| eval: false
x3 <- list(list(1, 2), list(3, 4))
x3
str(x3)

# Different to using c() - which generates flat vector:
x4 <- c(list(1, 2), list(3, 4))
x4
str(x4)

# str() allows us - with more complex lists - to understand the hierarchy:
x5 <- list(1, list(2, list(4, list(4, list(5)))))
x5
str(x5)

# With list getting more complex - str() fails - use View():
View(x5)
```

### List-columns

Lists can live within tibbles - **list-columns**.

USeful when placing objects in tibbles usually not belonging in there - used in tidymodels storing things liek model output or resamples in data frame.

```{r}
#| label: List-columns

df <- tibble(
  x = 1:2,
  y = c("a", "b"),
  z = list(list(1, 2 ), list(3, 4, 5))
)

df

# Column-lists still behave like any other column:
df |>
  filter(x == 1)
```

## Unnesting

Turning column-lists back into regular rows and columns - usually two basic forms of list-columns: named and unnamed

```{r}
#| label: named and unnamed column-lists

# Named list-columns:
# Usually same name in every row
# Unnesting - each named element becomes new named column
df1 <- tribble(
  ~x, ~y,
  1, list(a = 11, b = 12),
  2, list(a = 21, b = 22),
  3, list(a = 31, b = 32),
)

# Unnamed list-columns:
# Number of elements tend to vary from row-to-row
# Unnesting - unnest into rows - one row per observation
df2 <- tribble(
  ~x, ~y,
  1, list(11, 12, 13),
  2, list(21),
  3, list(31, 32),
)
```

Two functions for these two cases: `unnest_wider()` and `unnest_longer()`

### unnest_wider()

Each row has same number of elements with the same names - put each component into own column with `unnest_wider()`:

- names of new columns come from names of the list elements - can also use `names_sep` argument to distinguish column x from y:

```{r}
#| label: unnest_wider()

df1 |>
  unnest_wider(col = y)

# Using names_sep():
df1 |>
  unnest_wider(col = y, names_sep = "_") # avoiding ambiguating by repeating names
```

### unnest_longer

Each row contains an unnamed list - almost natural to put elements into its own row with `unnest_longer()`:

```{r}
#| label: unnest_longer()

df2 |>
  unnest_longer(col = y)
```

Note, `x` is duplicated for each element inside of `y` - one row of output for each element inside the list-column:

```{r}
#| label: unnest_longer() with empty element

df6 <- tribble(
  ~x, ~y,
  "a", list(1, 2),
  "b", list(3),
  "c", list()
)

df6 |>
  unnest_longer(col = y) # Row disappears when element is empty

# Can keep empty element by keep_empty = TRUE
df6 |>
  unnest_longer(col = y, keep_empty = TRUE)
```

### Inconsistent types

Unnesting list-columns containing different types of vector:

```{r}
#| label: Unnesting lists w different types

df4 <- tribble(
  ~x, ~y,
  "a", list(1),
  "b", list("a", TRUE, 5)
)

df4 |>
  unnest_longer(y)

```

Output contains a list-column - every element of the list-column contains a single element - `unnest_longer()` can't find common typeof vector, thus keeps original in a list-column. Note: Every element in this column is a list, even though contents of different types. See chapter 26 for dealing with inconsistent types.

### Other functions

- `unnest_auto()`: automatically picks between `unnest_longer()` and `unnest_wider()`

- `unnest()`: unnests both rows and columns - useful with list-column containing 2d structure like a data frame.

### Exercise

```{r}
#| label: Exercises

# Ex1: What happens when you use unnest_wider() with unnamed list-columns like df2? 
# What argument is now necessary?
df2 |>
  unnest_wider(y, names_sep = "_")

# Ex2: What happens when you use unnest_longer() with named list-columns like df1?
# What additional information do you get in the output?
# How can you suppress that extra detail?
df1 |>
  unnest_longer(y) # get extra column y_id containing of elements inside each list

df1 |>
  unnest_longer(y, indices_include = FALSE)

df1 |>
  unnest_longer(y) |>
  pivot_wider(names_from = "y_id", values_from = "y")

# Ex3: 
df4 <- tribble(
  ~x, ~y, ~z,
  "a", list("y-a-1", "y-a-2"), list("z-a-1", "z-a-2"),
  "b", list("y-b-1", "y-b-2", "y-b-3"), list("z-b-1", "z-b-2", "z-b-3")
)

df4 |> unnest_longer(y) |> unnest_longer(z)
# How to preserve relationship:
df4 |> unnest_longer(col = c(y, z))
```

## JSON

JSON (**JavaScript Object Notation**) is the standard format used by most web APIs to exchange data. It maps closely to R objects, although there is no perfect 1-to-1 correspondence.

### JSON data types

- `null` → `NA`
- **String** → character
- **Number** → numeric/integer (`Inf`, `-Inf`, and `NaN` are not supported)
- **Boolean** → `TRUE` / `FALSE` (`true` / `false` in JSON)
- **Array** (`[]`) → **unnamed list** in R
- **Object** (`{}`) → **named list** in R

::: callout-note
JSON has **no native date or datetime type**. Dates are usually stored as character strings and should be converted using `readr::parse_date()` or `readr::parse_datetime()`.
:::

### `jsonlite`

The **jsonlite** package provides two main functions for working with JSON:

- `read_json()` – reads a JSON file from disk.
- `parse_json()` – parses a JSON string into R objects.

`fromJSON()` is another commonly used function, but it automatically simplifies nested structures (`simplifyVector = TRUE`). For learning how to work with nested data, `parse_json()` is preferred because it preserves the original structure.

```{r}
#| label: JSON examples

library(jsonlite)

# Scalar
str(parse_json("1"))

# Array -> unnamed list
str(parse_json("[1, 2, 3]"))

# Object -> named list
str(parse_json('{"x": [1, 2, 3]}'))
```

## Top-level JSON structures

### 1. Top-level array

Each element represents an observation.

```{r}
#| label: Top-level array

json <- '[
  {"name": "John", "age": 34},
  {"name": "Susan", "age": 27}
]'

tibble(json = parse_json(json))
```

Result:

- 2 rows
- each row is one observation stored as a named list

### 2. Top-level object

The entire JSON is one object that contains other fields.

Since `parse_json()` returns a **single named list**, wrap it in `list()` before creating a tibble:

```{r}
#| label: Top-level array contd

json <- '{
  "status": "OK", 
  "results": [
    {"name": "John", "age": 34},
    {"name": "Susan", "age": 27}
 ]
}
'

tibble(json = list(parse_json(json)))
```

This creates a tibble with **one row**, where the JSON object is stored in a single list-column.

## Rectangling nested JSON

A typical workflow is:

1.  Parse the JSON using `parse_json()`.
2.  Determine whether the top level is an **array** or an **object**.
3.  Explore the structure using `unnest_longer()` and `unnest_wider()`.
4.  Once the path to the desired component is known, use `hoist()` to extract it directly.
5.  Convert character variables (e.g., dates or numbers) to their appropriate R types if necessary.

### Common tidyr functions

| Function | Purpose |
|--------------------------------------|----------------------------------|
| `unnest_longer()` | Turns list elements into **rows** (for arrays). |
| `unnest_wider()` | Turns named list elements into **columns** (for objects). |
| `hoist()` | Extracts specific nested components without unnesting everything. |

::: callout-tip
**Rule of thumb**

- **Array (`[]`)** → use `unnest_longer()`
- **Object (`{}`)** → use `unnest_wider()`
- **Known path to a nested value** → use `hoist()`
:::

## Starting the rectangling process

When working with JSON, the first step depends on the **top-level structure**.

### Case 1: Top-level array (`[]`)

A top-level array usually represents **multiple observations**. After parsing, each element becomes one row of a tibble.

```{r}
#| label: Top-level array contd2


json <- '[
  {"name": "John", "age": 34},
  {"name": "Susan", "age": 27}
]'

df <- tibble(json = parse_json(json))

df
```

Expand each named list into columns:

```{r}
#| label: Top-level array - contd
df |>
  unnest_wider(json)
```

Result:

- Each JSON object becomes **one row**.
- `unnest_wider()` extracts the fields (`name`, `age`) into columns.

------------------------------------------------------------------------

### Case 2: Top-level object (`{}`)

Sometimes the entire JSON is a **single object** containing metadata and the actual data.

```{r}
#| label: Top-level array - "single" object


json <- '{
  "status": "OK",
  "results": [
    {"name": "John", "age": 34},
    {"name": "Susan", "age": 27}
  ]
}'

df <- tibble(json = list(parse_json(json)))

df
```

Since the parsed JSON is **one named list**, wrap it in `list()` before creating the tibble.

Extract the nested observations:

```{r}
#| label: Top-level array - "single" object contd

json <- '{
  "status": "OK",
  "results": [
    {"name": "John", "age": 34},
    {"name": "Susan", "age": 27}
  ]
}'


df |>
  unnest_wider(json) |>
  unnest_longer(results) |>
  unnest_wider(results)
```

Result:

- `status` is repeated for each observation.
- `results` is an array, so `unnest_longer()` creates one row per element.
- Each result is a named list, so `unnest_wider()` converts its fields into columns.

------------------------------------------------------------------------

### Extract only the component you need

If you're only interested in a nested component, you can access it directly before creating the tibble:

```{r}
#| label: Extract only the component you need

json <- '{
  "status": "OK",
  "results": [
    {"name": "John", "age": 34},
    {"name": "Susan", "age": 27}
  ]
}'


df <- tibble(results = parse_json(json)$results)

df |>
  unnest_wider(results)
```

This avoids carrying along unnecessary metadata (e.g., `status`) and is often the cleanest approach.

::: callout-tip
**Rule of thumb**

- Top-level **array (`[]`)** → `tibble(json = parse_json(json))`
- Top-level **object (`{}`)** → `tibble(json = list(parse_json(json)))`
- If you only need one nested component, extract it first (e.g. `parse_json(json)$results`).
:::


### Exercises
```{r}
#| label: Rectangling exercise

json_col <- parse_json('
  {
    "x": ["a", "x", "z"],
    "y": [10, null, 3]
  }
')
json_row <- parse_json('
  [
    {"x": "a", "y": 10},
    {"x": "x", "y": null},
    {"x": "z", "y": 3}
  ]
')

df_col <- tibble(json = list(json_col)) 
df_row <- tibble(json = json_row)

df_col |>
  unnest_wider(json) |>
  unnest_longer(c(x, y))

df_row |>
  unnest_wider(json) 
```

