---
title: "Basics"
format: html
---

::: callout-important
Always write and comment code in such a way that it is clear and that, even five years from now, you can quickly understand what your code does, how and why it works, and how to use it correctly.
:::

### 2.3 Comments and assignments - "\#", "\<-"

```{r}
#| label: Assignemnt types

# assignment types

# left assignment - recommended
a <- 2 # Defining parameter (comments next to code allowed!)
a

# right assignment - not recommended
2 -> a
a

# assignment via "=" - unususal in R
a = 2
a
```

### 2.4 R as calculator

| Operator | Meaning        |
|:--------:|----------------|
|   `+`    | Addition       |
|   `-`    | Subtraction    |
|   `*`    | Multiplication |
|   `/`    | Division       |
|   `^`    | Power          |
| `srqt()` | Square root    |

: Arithmetic operations

```{r}
#| label: Arithmetic operations

# arithmetic operations
# power
3 ^ 2
3 ^ 3

# square root
sqrt(9)
# 3rd square root
27 ^ (1 / 3)

# pay attention to brackets
9 ^ (1 / 2)
# not same as:
9 ^ 1 / 2

# ex: golden ratio
gamma <- (sqrt(5) - 1) / 2
gamma

```

### 2.5 Useful shortcuts

| Shortcut | Function |
|:--------------------------------------|:--------------------------------|
| `CTRL+L` | Clears the R console. |
| `CTRL+TAB` | Switches to the next window (e.g., from the console to the script window). |
| `CTRL+A` | Selects all code in the script window. |
| `CTRL+R` | Runs the selected code or the current line in the R console (`CTRL+ENTER` in RStudio). |
| `ARROW KEYS` | Navigate to the previous (`↑`) or next (`↓`) line of code in the R console. |
| `ESC` | Cancels the current command. |

: Useful Keyboard Shortcuts for the R Console

### 2.7 Installing and loading packages

```         
Installing and loading packages

# install package
install.packages("pkg_name") - once

# load package
library(pkg_name) - done every R session
```

#### 2.8.3 Exercises

```{r}
#| label: 2.8.3 Exercises

# Ex1: BMI = kg / m^2; Person: 65kg, 1.76m 
# a) create variables
kg <- 65
m <- 1.76
m2 <- m^2
# b) compute BMI
bmi <- kg / m2
bmi

# Ex2: Closed form fibonacci sequence
n <- 8:10
x <- 1/5
y <- ((1 + sqrt(5)) / 2)^n
z <- ((1 - sqrt(5)) / 2)^n

cf_fib <- x * (y - z)
cf_fib
```

## 3. Vectors and Logical Queries

### 3.1 Generating vectors

#### 3.1.1 Concatenating Elements into Vectors – c()

| Index   | 1   | 2   | 3   | 4   | 5   | 6   |
|:--------|:----|:----|:----|:----|:----|:----|
| Element | 4   | 7   | 6   | 2   | 3   | 5   |

: Tbl showing shuffled deck of cards

```{r}
#| label: Concatenating elements into vector

# Concatenating elements into vector
stack <- c(4, 7, 6, 2, 3, 5) # can combine arbitrary number of elmts into vec
stack
```

#### 3.1.2 Indexing and length of vectors - length()

In a vector, R starts indexing elements with 1 - last elements has index `length(vec)`:

```{r}
#| label: indexing and length of vector

# indexing and length of vector
length(stack)

# can access the indexes with seq_along; See ?seq_along
# seq_along(stack)[1]

```

#### 3.1.3 Simple sequences - ":"

Simple sequences with step size 1 - computed by `:`:

```{r}
#| label: Simple sequences

# Simple sequences
1:6

# other way around
6:1

# negative numbers
-2:3

-(2:3)

# works with decimal numbers
1.5:6.5

# incomplet seq
1:4.5

# Can generate index vector for stack this way
1:length(stack) # see ?seq_along
```

### 3.2 Selecting elements: subsetting - "\[ \]"

Use subsetting via `[ ]` to access elements of vectors. Two methods: + Subsetting via indices (3.2.1) + Subsetting via boolean values (3.2.2)

#### 3.2.1 Subsetting via indices - "-"

```{r}
#| label: Subsetting via indices

# Subsetting via indices
stack
# top card
stack[1]
# second and fourth card
stack[c(2, 4)]
# first and last card
stack[c(1, length(stack))] # use length() for flexibility of code

# subsetting created a new vector
stack[3:5] # gets new indices

# assinging elements to temp
temp <- stack[3:5]
temp # "temporary"

# accessing first card in temp:
temp[1]
stack[3:5][1] # nested alternative

# alternatively: caching indices
ind <- 3:5
stack[ind]
```

What happenns if we try to access non-existing elements?

```{r}
#| label: Accessing non-existing elements

# Accessing non-existing elements
stack[c(2, 9, 2)] # NA (Not Available) appear
```

Omitting specific elements:

```{r}
#| label: Omitting specific elements

# Omitting specific elements via "-"
# selecting all elements except specified ones
stack[-c(3:5)]
stack[-c(3:5, 1)]
```

Dealing the cards equally:

```{r}
#| label: dealing cards

# Player 2 receives every second card
ind <- c(1, 3, 5)

hand1 <- stack[ind]
hand1

hand2 <- stack[-ind]
hand2
```

3.3 presents approach that is appliable to all lengths of vectors.

#### 3.2.2 Subsetting with Boolean Values

```{r}
#| label: Subsetting with boolean values

# Subsetting with boolean values
stack

# select alternating
bool1 <- c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE)
bool1

hand1 <- stack[bool1]
hand1

# Can reverse operation for second player via negation
hand2 <- stack[!bool1]
hand2


```

### 3.3 Recycling

```{r}
#| label: Recycling

# Selecting cards alternatung via recycling
stack

# subsetting vec recycled till reaching length of stack
stack[c(TRUE, FALSE)] # 6 divisible by 2, thus "complete recycling"

# incomplete recycling - subsetting vector cuts off stack
stack[c(TRUE, FALSE, FALSE, FALSE)] #  4 no divisor of 6
# "4" 7 6 2 "3" 5
```

Recycling also common in arithmetic operations:

```{r}
#| label: Recycling in arithmetic operations

# Recycling in arithmetic operations
y <- 1:5 * c(-1, 1)
y
```

### 3.4 Logical Conditions and Operators

#### 3.4.1 Comparing Elements – “==”, “!”, “!=”, “\<”, “\<=”, “\>”, “\>=”

|  Operator   | Meaning                                  |
|:-----------:|------------------------------------------|
|    `==`     | equal to                                 |
|     `!`     | not (negates a logical value)            |
|    `!=`     | not equal to                             |
| `<` or `<=` | less than or less than or equal to       |
| `>` or `>=` | greater than or greater than or equal to |

: Logical operators for comparisons and negation

```{r}
#| label: Logical conditions

stack

# elements equal 5
stack == 5
# elements greater than 5
stack > 5

# elements not equal 5
stack != 5

# selecting all cards at least 5
stack[stack >= 5] # original order of cards still retained

stack[! (stack < 5)]

stack[! stack < 5] # "<" takes priority

# logical conditions no element fulfills returns empty vec
stack[stack < 0]

length(stack[stack < 0]) # empty vec has lenght = 0

length(stack[stack < 0]) == 0 # checking if indeed empty
```

#### 3.4.2 Linking Conditions – “&”, “\|”

| Operator | Meaning       |
|:--------:|---------------|
|   `&`    | AND operation |
|   `\|`   | OR operation  |

: Operators for combining logical values

`AND operation` equals `TRUE` if both boolean values are `TRUE`. `OR operation` equals `TRUE` if at least one of the boolean values return `TRUE`.

```{r}
#| label: Linking Conditions - "&", "|"

# AND operations
TRUE & TRUE # TRUE

TRUE & FALSE # FALSE

FALSE & FALSE # FALSE

# OR operations
TRUE | TRUE # TRUE

TRUE | FALSE # TRUE

FALSE | FALSE # FALSE

# Ex: drawing cards greater than 2 and less or equal to 5
stack > 2
stack <= 5
stack > 2 & stack <= 5 

stack[stack > 2 & stack <= 5]

# Ex: drawing cards less than 4 or greater than 5
stack < 4
stack > 5
temp <- stack < 4 | stack > 5

stack[temp]

```

### 3.5 Calculating with Boolean Values

#### 3.5.1 Absolute and Relative Frequencies – sum(), mean()

```{r}
#| label: Example 3.5.1 absolute and rel freq

# Ex: how many cards with at least val 5
# cards p1
hand1 >= 5
sum(hand1 >= 5)

# cards p2
hand2 >= 5
sum(hand2 >= 5)
```

::: callout-important
R encodes `TRUE` as `1` and `FALSE` as `0` - thus, can use `sum()` to compute the number of `TRUE`.
:::

This also applies to other arithemtic operations:

```{r}
#| label: Encoding of boolean values

# 1 + 0 + 1
TRUE + FALSE + TRUE
# T + F + T; alternatively

# 2 * 1 + 1 * 0
2 * TRUE + 1 * FALSE

# Proportion of TRUE values - use mean()
mean(stack >= 5) # 50% of cards greater or equal to 5

# Ex: Did p1 receive more cards greater than or equal to 5 than p2?
sum(hand1 >= 5) > sum(hand2 >= 5) # FALSE
```

#### 3.5.2 Determining Index Values - which()

Used when interested in specific indices fulfilling some conditions. So far:

```         
(1:length(vec))[bool] # reveals at what index TRUE lies
```

```{r}
#| label: Det ind vals - which()

stack

# retrieve indices
1:length(stack)
stack == 7 

index <- (1:length(stack))[stack == 7]
index # 7 is at position 2

# alternatively, and more conveniently
which(stack == 7)

# Ex: positions with cards equal or greater than 5
which(stack >= 5)

# Ex: position of first card with val 5 or greater
ind1 <- which(stack >= 5)[1]
ind1
stack[ind1] # value of this card
```

What happens when calling `which()` on non-existing indices:

```{r}
#| label: which() on non-ex indices

# no element < 0 results in empty vec
which(stack < 0)
length(which(stack < 0))
length(which(stack < 0)) == 0

# trying to access non-ex elements
which(stack < 0)[1] # NA
stack[which(stack < 0)[1]] # NA

```

### 3.6 Replacing and Swapping Values

Elements of a vector can be replaced by selecting positions and assigning new values:

`vector[position] <- replacement`

```{r}
#| label: Example 3.6.1 replacing values

x <- c(2, -5, 3, 4, -1)

# double all positive values
x[x > 0] <- x[x > 0] * 2
x

# replace all negative values by 0
x[x < 0] <- 0
x
```

::: callout-important
When assigning a single value to multiple positions, **recycling** takes place.

For example, `0` is recycled to the length of `x[x < 0]`.
:::

#### 3.6.1 Swapping Values

```{r}
#| label: Example 3.6.2 swapping values

stack <- c(4, 7, 6, 2, 3, 5)

# swap first and last card - wrong!
stack_new <- stack
stack_new[1] <- stack_new[length(stack_new)]
stack_new[length(stack_new)] <- stack_new[1]
stack_new
```

The first value is overwritten before it can be assigned to the last position.

We can instead access the values of the original `stack`:

```{r}
#| label: Example 3.6.3 swapping values correctly

stack_new <- stack

stack_new[1] <- stack[length(stack)]
stack_new[length(stack_new)] <- stack[1]

stack_new
```

Alternatively, use a temporary object:

```{r}
#| label: Example 3.6.4 swapping with temporary object

stack_new <- stack

temp <- stack_new[1]
stack_new[1] <- stack_new[length(stack_new)]
stack_new[length(stack_new)] <- temp

stack_new
```

::: callout-important
When values are overwritten sequentially, store values that are still needed in a **temporary object**.
:::

### 3.7 Good Programming Practice

#### 3.7.1 Meaningful Object Names and Special Characters

Object names should be **short and meaningful**.

Examples of meaningful names:

```
stack
mixed_stack
hand_player1
```

Avoid names such as:

``` text
x
1.hand
hand 1
kartenstoß
blätter1
s
```

General rules:

- use short and descriptive names
- object names cannot start with a number
- avoid spaces and special characters
- avoid umlauts (`ä`, `ö`, `ü`) and `ß`
- use nouns for objects
- prefer lowercase names
- `.` and `_` can be used to separate words

::: callout-note
R is **case-sensitive**. Thus, `stack` and `Stack` refer to different objects.
:::

#### 3.7.2 General and Automated Programming

Code should be written such that it still works when the underlying data changes.

```{r}
#| label: Example 3.7.1 general programming

stack <- c(4, 7, 6, 2, 3, 5)

# last card - expandable
stack[6]

# last card - general
stack[length(stack)]

# every second card - expandable
stack[c(1, 3, 5)]

# every second card - general
stack[c(TRUE, FALSE)]
```

If the stack is extended:

```{r}
stack <- c(stack, 9, 8)
stack
```

the manually specified code no longer works correctly:

```{r}
stack[6]
stack[c(1, 3, 5)]
```

while the general code automatically adapts:

```{r}
stack[length(stack)]
stack[c(TRUE, FALSE)]
```

::: callout-important
Program as generally and automatically as possible. Let R determine information such as vector lengths or positions instead of **hard-coding** them.
:::

### 3.8 Exercises

#### Ex 1: Counting Boolean Values

```{r}
stack <- c(4, 7, 6, 2, 3, 5)

# number of cards with value >= 4
sum(stack >= 4)

# alternatively
length(stack[stack >= 4])

# using which()
length(which(stack >= 4))
```

`sum(stack >= 4)` works because R encodes `TRUE` as `1` and `FALSE` as `0`.

Using `which()` is unnecessary when we only want to select or count elements based on a condition.

#### Ex 2: Vector Operations

```{r}
x <- (1:6) * c(-1, 1)
x
```

`c(-1, 1)` is recycled:

``` text
1:6       1  2  3  4  5  6
c(-1, 1) -1  1 -1  1 -1  1
           ↓  ↓  ↓  ↓  ↓  ↓
x         -1  2 -3  4 -5  6
```

Replace all non-positive values by their negative value:

```{r}
x[x <= 0] <- -x[x <= 0]
x
```

Multiply using recycled Boolean values:

```{r}
x <- x * c(TRUE, FALSE)
x
```

Since `TRUE = 1` and `FALSE = 0`, every second value becomes `0`.

Select values different from `0` and smaller than `5`:

```{r}
x <- x[x != 0 & x < 5]
x
```

Select elements using numerical indices:

```{r}
x <- x[c(0:3, length(x), 2, 1)]
x
```

::: callout-note
When indexing, `0` is ignored. Repeated indices select the corresponding element multiple times.
:::

#### Ex 3: Select All Elements Except the Last $k$

```{r}
#| label: Ex 3 select all except last k

x <- 1:5
x
```

Approach 1:

```{r}
# brackets needed due to operator precedence
x[1:(length(x) - 2)]
```

Approach 2:

```{r}
# exclude last two positions
x[-c(length(x) - 1, length(x))]
```

Generalizing to the last `k` elements:

```{r}
k <- 2

# approach 1
x[1:(length(x) - k)]

# approach 2
x[-((length(x) - k + 1):length(x))]
```

##### Function

```{r}
#| label: Ex 3 function

omit_last_elmt <- function(vec, k) {
  if (length(vec) >= k + 1) {
    vec[1:(length(vec) - k)]
  } else NA
}

omit_last_elmt(x, 2)
```

Alternatively, using `seq_len()`:

```{r}
omit_last_elmt <- function(vec, k) {
  if (length(vec) >= k + 1) {
    vec[seq_len(length(vec) - k)]
  } else NA
}

omit_last_elmt(x, 2)
```

#### Ex 4: Logical Conditions

```{r}
age <- c(16, 18, 19, 24, 28, 32, 32, 45)

# number of entries
length(age)

# percentage between 18 and 27
mean(age >= 18 & age <= 27) * 100

# younger than 20 or older than 30
sum(age < 20 | age > 30)
```

#### Ex 5: Swapping Cards Between Two Players

```{r}
hand1 <- c(4, 6, 3)
hand2 <- c(7, 2, 5)
```

This approach does **not** work:

```{r}
hand1[1] <- hand2[1]
hand2[1] <- hand1[1]
```

`hand1[1]` is overwritten before its original value can be assigned to `hand2[1]`.

Use a temporary object instead:

```{r}
hand1 <- c(4, 6, 3)
hand2 <- c(7, 2, 5)

temp <- hand1[1]

hand1[1] <- hand2[1]
hand2[1] <- temp

hand1
hand2
```
