---
title: "Databases"
subtitle: "R for Data Science (2e), Chapter 21"
toc: true
theme: darkly
format: html
---

## Introduction

Basics of DBI package - using it to connect to databases and retrieving data with SQL (structured query language) query. 

__dbplyr__ - translates dplyr code to the SQL.


### Prerequisites
```{r}
#| label: Prereq

library(DBI)
library(dbplyr)
library(tidyverse)
```


## Database basics

Simplest level - Database as collection of data frames, .i.e., __tables__.
Database table is collection of named columns - every value in column is same type. 

Differences:

+ Databases stored on disk - can be arbitrarly large. Data frames stored in memory and are limited.

+ Database tables almost always have indexes unlike data frames and tibbles. 
+ Classical databases optimized for data collection - __row-orientend database__. __Column-orieneted__ databases make analyzing existing data much faster

Databases are run by database management systems( __DBMS__ ); Three forms:

+ __Client-server__: Powerful central server; Useful for haring within organization - e.g., PostgreSQL, SQL Server, Oracle, etc.

+ __Cloud__: e.g., Snowflake, Amazon's RedShift, Google's BigQuery run in cloud - useful for handling extremely large datasets/ more computing resources

+ __In-process__: e.g., SQLite, duckdb, run on computer - useful wehn working with large datasets where you're the primary user.


## Connecting to a database

Connecting from R requires pair of packages:

+ DBI (database interface) - provides generic functions that connect to database, uplaod data, runSQL queries, etc. 

+ Package for DBMS - translates generic DBI commands into specifics needed for given DBMS - usually one package for each DBMS, e.g., RPostgres for PostgreSQL, etc.

Creat database connection using `DBI::dbConnect()`:
+ first argument selects DBMS
+ second/subsequent argument describes how to connect to it

```{r}
#| label: connecting-to-database
#| eval: false


 con <- DBI::dbConnect(   
  RMariaDB::MariaDB(),
  username = "foo"
)

con <- DBI::dbConnect(
  RPostgresDB:: Postgres(),
  hostname = "database.mycompany.com",
  port = 1234
)
```

For specifics ask DBA (database administrator) in the team.


### In this book

Connect to duckdb - default creates temporary database that is deleted when quitting R:
```{r}
#| label: connect-to-duckdb

con <- DBI::dbConnect(
  duckdb::duckdb()
)
```



## Load some data

Add datasets from ggplot2 using `DBI::dbWriteTables()` - three arguments: 
+ database connection
+ name of table to create in database
+ data frame of data
```{r}
#| label: Loading data in db

 dbWriteTable(con, "mpg", ggplot2::mpg)
 dbWriteTable(con, "diamonds", ggplot2::diamonds)

```

Recommendation: Learn `duckdb_read_csv()`; `duckdb_register_arrow()`


### DBI basics

```{r}
#| label: DBI basics

 dbListTables(con)

 con |>
  dbReadTable("diamonds") |>
  as_tibble()
```


Can also use SQL query - specify SQL query and call `dbGetQuery()`:
```{r}
#| label: Get query results

sql <- "
SELECT carat, cut, clarity, color, price
FROM Diamonds
WHERE price > 15000"

 as_tibble(dbGetQuery(con, sql))


```


## dbplyr basics

dbplyr is dplyr __backend__ - i.e., we write dplyr code, but backend executes it differently - dbplyr translates to SQL (see different backends).

To use dbplyr must create object that represents database table using `tbl()`:
```{r}
#| label: dbplyr basics

 diamonds_db <- tbl(con, "diamonds")
 diamonds_db

```
This object is __lazy__, i.e., only records sequence of operations and only performs operations when needed (asked for): 
```{r}
#| label: lazy tbl() object

 big_diamonds_db <- diamonds_db |>
  filter(price > 15000) |>
  select(carat:clarity, price)

# big_diamonds_db
```

Can see SQL code generated by dplyr function using `show_query()`:
```{r}
big_diamonds_db|> show_query()
```
To get all data back into R - use `collect()`:
```{r}
#| label: Getting data into R

# after finishing with queries, use collect()

big_diamonds <- big_diamonds_db |>
  collect()
big_diamonds
```


## SQL

```{r}
#| label: Copying tables to database

dbplyr::copy_nycflights13(con) # in-built function of dbplyr for flights

# Create object representing database table:
flights <- tbl(con, "flights")
planes <- tbl(con, "planes")
```

### SQL Basics

Top-level components of SQL called __statements__:

+ `CREATE` for defining new tables
+ `INSERT` for adding data
+ `SELECT` for retrieving data - also called __queries__

Query is made up of clauses:
Five important clauses: `SELECT`, `FROM`, `WHERE`, `ORDER BY`, and `GROUP BY`.

Every query must have `SELECT` and `FROM` clauses - e.g., simple query SELECT * FROM table - selects allscolumns from table:
```{r}
#| label: Example Query

flights |> show_query()

planes |> show_query()

# WHERE controls which rows are included
# ORDER BY how they are ordered

flights |>
  filter(dest == "IAH") |>
  arrange(dep_delay) |>
  show_query()

# GROUP BY converts query to summary, causing aggregation:
flights |>
  group_by(dest) |>
  summarize(dep_delay = mean(dep_delay, na.rm = TRUE)) |>
  show_query()
```
:::{.callout_important}
In SQL order matters: SELECT, FROM, WHERE, GROUP BY, ORDER BY.
BUT order of evaluation: FROM, WHERE, GROUP BY, SELECT, ORDER BY.
:::


### SELECT

`SELECT` performs same job as `select()`, `mutate()`, `rename()`, `relocate()`, `summarize()`.

Renaming in SQL is called __aliasing__ and is done via `AS` - new name on the right NOT on the left.
```{r}
#| label: SQL SELECT

planes |>
  select(tailnum, type, manufacturer, model, year) |>
  show_query()

planes |>
  select(tailnum, type, manufacturer, model, year) |>
  rename(year_built = year) |>
  show_query()

planes |>
  select(tailnum, type, manufacturer, model, year) |>
  relocate(manufacturer, model, .before = type) |>
  show_query()

flights |>
  mutate(
    speed = distance / (air_time / 60)
  ) |>
  show_query()
```


### FROM

`FROM` clause defines the data source.


### GROUP BY

`group_by()` translates to `GROUP BY` clause;
`summarize()` translates to `SELECT`: 

```{r}
diamonds_db |>
  group_by(cut) |>
  summarize(
    n = n(),
    avg_price = mean(price, na.rm = TRUE)
  ) |>
  show_query()
```

### WHERE

`filter()` is translated to `WHERE` clause:
```{r}
flights |>
  filter(dest == "IAH" | dest == "HOU") |>
  show_query()

flights |>
  filter(arr_delay > 0 & arr_delay < 20) |>
  show_query()
```
+ `|` becomes `OR` and `&` becomes `AND`.
+ SQL uses `=` for comparison, not `==`. 
+ SQL uses only `''` for strings, not `""`.
+ SQL uses `IN` which is close to R's `%in%`

```{r}
#| label: SQL IN clause

flights |> 
  filter(dest %in% c("IAH", "HOU")) |>
  show_query()
```
SQL uses `NULL` instead off `NA`:
```{r}
#| label: SQL NULL

flights |>
  group_by(dest) |>
  summarize(delay = mean(arr_delay))
  

flights |>
  filter(!is.na(dep_delay)) |>
  show_query() # usually in SQL: WHERE dep_delay IS NOT NULL (equivalent)

```

Variables that are created using summarize, when filtering, will need `HAVING` clause in SQL, rather than `WHERE`:
```{r}
diamonds_db |>
  group_by(cut) |>
  summarize(n = n()) |>
  filter(n > 100) |>
  show_query()
```


### ORDER BY

`arrange()` is translated to `ORDER BY` clause:
```{r}
#| label: ORDER BY

flights |>
  arrange(year, month, day, desc(dep_delay)) |>
  show_query()
```


### Subqueries

Subsquery is a query used as a data source in the `FROM` clause - sometimes dbplyr uses subqueries working around limitations of SQL. 
```{r}
#| label: Subqueries

# Expressions in SELECT call cannot refer to column just created

flights |>
  mutate(
    year1 = year + 1,
    year2 = year1 + 1
  ) |>
  show_query()
```


### Joins

SQL are similar to dplyr joins: 
```{r}
#| label: Joins

flights |>
  left_join(planes |> rename(year_built = year), join_by(tailnum)) |>
  show_query()
```
__SQL Joins__:
SELECT flights.*, "type", manufacturer, model, engines, seats, speed
FROM flights
INNER JOIN planes ON (flights.tailnum = planes.tailnum)

SELECT flights.*, "type", manufacturer, model, engines, seats, speed
FROM flights
RIGHT JOIN planes ON (flights.tailnum = planes.tailnum)

SELECT flights.*, "type", manufacturer, model, engines, seats, speed
FROM flights
FULL JOIN planes ON (flights.tailnum = planes.tailnum)


:::{.callout-tip}
`dm package` for use when dealing with many tables - automatically determines the connections between tables, visualize them and generate the joins needed to connect one table to another.
:::

### Other verbs
Full set available at: https://dbplyr.tidyverse.org/reference/



