---
title: "Spreadsheets"
subtitle: "R for Data Science (2e), Chapter 20"
toc: true
theme: darkly
format: html
---

## Excel

Data organized in worksheets inside spreadsheet files.


### Prerequisites
```{r}
#| label: Prerequisites
#| output: false

library(readxl)
library(tidyverse)
library(writexl)

```


### Getting started

+ `read_xls()` - reads Excel files with `xls` format

+ `read_xlsx()` - reads Excel files with `xlsx` format

+ `read_excel()` - reads both `xls` and `xlsx` - guesses file type based in input


### Reading Excel spreadsheets
```{r}
#| label: Loading excel data


students <- read_excel(path = "/Users/filipkukic/Desktop/students.xlsx")

students

students <- read_excel(path = "/Users/filipkukic/Desktop/students.xlsx",
           col_names = c("student_id", "full_name", "favourite_food", "meal_plan", "age"),
           na = c("", "N/A"), # empty strings also recognized as NA 
           skip = 1,
           col_types = c("numeric", "text", "text", "text", "text")) # load age as text first, then parse as number
                                            
students <- students |>
  mutate(
    age = if_else(age == "five", "5", age),
    age = parse_number(age)
  )
```



### Reading worksheets

```{r}
#| label: Reading worksheets

penguins_torgersen <- read_excel(path = "/Users/filipkukic/desktop/penguins.xlsx",
                       sheet = "Torgersen Island",
                       na = c("", "NA"))

penguins

# Can use excel_sheets() to get information on all worksheets in Excel spreads.
# then read the ones interested in 
excel_sheets(path = "/Users/filipkukic/desktop/penguins.xlsx")

penguins_biscoe <- read_excel(path = "/Users/filipkukic/desktop/penguins.xlsx",
                              sheet = "Biscoe Island", 
                              na = c("", "NA"))

penguins_dream <- read_excel(path = "/Users/filipkukic/desktop/penguins.xlsx",
                             sheet = "Dream Island", 
                             na = c("", "NA"))
```

Penguins dataset spread across three worksheets in spreadsheet:
```{r}
#| label: Dimensions

dim(penguins_torgersen)
dim(penguins_biscoe)
dim(penguins_dream)

# Have same number of columns, can bind them via bind_rows()

# penguins_alt_bind <- rbind(penguins_torgersen, penguins_biscoe, penguins_dream)

penguins <- bind_rows(penguins_torgersen, penguins_biscoe, penguins_dream) # more flexible than base R rbind

```


### Reading part of a sheet

```{r}
#| label: Reading parts of a sheet

# can load example sheets via readxl_example()
deaths_path <- readxl_example("deaths.xlsx")
deaths_path

deaths <- read_excel(path = deaths_path)
deaths



deaths <- read_excel(path = deaths_path,
                     range = "A5:F15") # can also use skip, n_max arguments
deaths
```


### Data types

__CSV__: Values are strings - everything is a string.

__Excel__: Values (Cells) can be four things:
+ boolean - `TRUE`, `FALSE`, `NA`
+ number - "10", "10.5", etc. 
+ datetime - "11/1/21", "11/1/21 3:00 PM", etc.
+ text string - "ten", etc. 

__Recommendation__: Let readxl guess the column types - confirm column types - if necessary, go back and re-import specifying `col_types`. If column has mix of types - set type of column to `"list"`, which loads column as list of length 1, where type of each element of vector is guessed.


### Writing to Excel

```{r}
#| label: Writing to Excel

bake_sale <- tibble(
  item = factor(c("brownie", "cupcake", "cookie")),
  quantitiy = c(10, 5, 8)
)

bake_sale

# Can write data back to disk as Excel file using write_xlsx() function:
write_xlsx(bake_sale, path = "/Users/filipkukic/Desktop/bake-sale.xlsx")

# Excel unreliable - as we read back in data type will be lost:
read_excel(path = "/Users/filipkukic/Desktop/bake-sale.xlsx")
# Note: type of item is "character", though we specified it as "factor"

```


### Formatted output

For additional features - writing to sheets within spreadsheet and styling - use __openxlsx package__ - reading recommendation: https://ycphs.github.io/openxlsx/articles/Formatting.html.


### Exercises
```{r}
#| label: Exercise1

# Ex1
# Create data frame
survey <- tibble(
  survey_id = 1:6,
  n_pets = c(0, 1, "N/A", "two", 2, "")
)
# Write to excel
write_xlsx(survey, path = "/Users/filipkukic/Desktop/survey.xlsx")

# Read in excel
survey_loaded <- read_excel(path = "/Users/filipkukic/Desktop/survey.xlsx",
                            na = c("", "N/A"),
                            col_types = c("numeric", "text"))

survey_loaded |>
  mutate(
    n_pets = if_else(n_pets == "two", "2", n_pets),
    n_pets = parse_number(n_pets)
  )


# Ex2
roster <- read_excel(path = "/Users/filipkukic/Desktop/roster.xlsx") |>
  fill(group, subgroup) 

roster


# Ex3 

sales1 <- read_excel(path = "/Users/filipkukic/Desktop/sales.xlsx",
                    skip = 4) |>
  slice_head(n = 3) |> 
  rename("id" = "Brand 1") |>
  mutate("brand" = "Brand 1",
         n = parse_number(n),
         id = parse_number(id))
sales1

# |>
#  mutate("id" = "Brand 1")
         


sales2 <- read_excel(path = "/Users/filipkukic/Desktop/sales.xlsx",
                    skip = 8) |>
  rename("id" = "Brand 2") |>
  mutate("brand" = "Brand 2")

sales2

sales <- bind_rows(sales1, sales2) # binding both
sales <- sales |>
  relocate(brand, .before = id) # relocating brand
sales  

# Alternatively: 

sales <- read_excel(
  path = "~/Desktop/sales.xlsx",
  skip = 4,
  col_names = c("id", "n"),
  n_max = 9
)

sales <- sales |>
  mutate(
    brand = if_else(str_detect(id, "Brand"), id, NA_character_)
  ) |>
  fill(brand) |>
  filter(!str_detect(id, "Brand")) |>
  mutate(
    id = parse_number(id),
    n = parse_number(n)
  ) |>
  relocate(brand)

sales


# Ex4: 

read_excel(path = "~/Desktop/students.xlsx", 
           na = c("", "N/A")) |>
  janitor::clean_names() |>
  mutate(age = if_else(age == "five", "5", age),
         age = parse_number(age))
```


## Google Sheets

### Prerequisites
```{r}
#| label: Prereq - google sheets

library(googlesheets4)
library(tidyverse)
```


### Getting started

__Main fuction__: 

`read_sheet()` - reads Google Sheet from URL or file id (function also called `range_read()`).

Can create brand new sheet with `gs4_create()` or write to existing sheet with `sheet_write()`


### Reading Google Sheets

First argument to `read_sheet()` is URL of file to read:

https://docs.google.com/spreadsheets/d/1V1nPp1tzOuutXFLb3G9Eyxi3qxeEhnOXUzL5_BcCQ0w - (file from before, that we downloaded as xlsx)

Often useful to find file ID instead: 1V1nPp1tzOuutXFLb3G9Eyxi3qxeEhnOXUzL5_BcCQ0w
```{r}
gs4_deauth() # see ?gs4_deauth()

students_sheet_id <- "1V1nPp1tzOuutXFLb3G9Eyxi3qxeEhnOXUzL5_BcCQ0w"

students <- read_sheet(students_sheet_id)
students

# Similar to read_excel(), can supply column name and type, and na string
students <- read_sheet(students_sheet_id,
                       na = c("", "N/A"),
                       col_names =  c("student_id", "full_name", "favourite_food", "meal_plan", "age"),
                       skip = 1,
                       col_types = "dcccc") |>
  mutate(age = if_else(age == "five", "5", age), 
         age = parse_number(age))

students
```


Can also read individual sheets from Google Sheets:

Can obtain list of all sheets within Google Sheets with `sheet_names()`
and can read in portion of Google Sheets by defining `range` ind `read_sheet()`.
```{r}
#| label: Reading individual sheets from google sheets

# penguins_sheet_id <- "1aFu8lnD_g0yjF5O-K6SFgSEWiHPpgvFCF0NY9D6LXnY"
# read_sheet(penguins_sheet_id, sheet = "Torgersen Island")

```


### Writing to Google Sheets

`write_sheet()` - first argument is data frame, second argument is name (identifier) of Google Sheet to write to:
```{r}
#| label: write_sheet()

# write_sheet(bake_sale, ss = "bake-sale")

# Can also specify the sheet argument

# write_sheet(bake_sale, ss = "bake-sale", sheet = "Sales")
```


### Authentication

Reading a private sheet or writing sheet requires authentication - gs4 can view and manage your Google Sheets.

See https://googlesheets4.tidyverse.org/articles/auth.html.


