---
title: "dates-and-times"
subtitle: "R for Data Science (2e), Chapter 17"
toc: true
theme: darkly
format: html
---

### Prerequisites+

Chapter will focus on __lubridate__ package - part of core tidyverse:
```{r}
#| label: Prerequisites

library(tidyverse)
library(nycflights13)
```


## Creating date/times

Three types of date/time data refering to instant in time:

+ __date__: Tibbles print this as `<date>`

+ __time__ within day: Tibbles print this as `<time>`

+ __date-time__ date plus time: Uniquely identifies instant in time. Tibbles print this as <dttm> - Base R calls these POSIXct

:::{.callout-tip}
Use simplest possible data type - if can work with date instead of date-time, then do.
:::

```{r}
#| label: Current time and date

# Current time and date
today()
now()
```



### During import

CSV contains ISO8601 date or date-time - readr will automatically recognize it:
```{r}
#| label: readr for date, date-time

csv <- "
date,datetime
2022-01-02, 2022-01-02 05:12
"
read_csv(file = csv)

```

:::{.callout-note}
__ISO8601__ - international standard for writing dates - e.g., May 3 2022 is `2022-05-03`. Can include time separated by `:`. Date and time separated either by space or `T` - `2022-05-03 16:26` or `2022-05-03T16:26`.
:::

Other date-time formats require use of `col_types` plus `col_date()` or `col_datetime()` with date-time format. See `?col_date` for all format specifications! - e.g., `%Y-%m-%d` specifies a date that’s a year, `-`, month (as number) `-`, day:
```{r}
#| label: date-time formatting

csv <- "
  date
  01/02/15
"

read_csv(
  file = csv,
  col_types = cols(
    date = col_date(format = "%d/%m/%y")
  )
)

read_csv(
  file = csv,
  col_types = cols(
    date = col_date(format = "%m/%d/%y")
  )
)

read_csv(
  file = csv,
  col_types = cols(
    date = col_date(format = "%y/%m/%d")
  )
)

```

Using `%b` or `%B` needs `locale()` - see `date_names_langs()` or create via `date_names()`.


### From strings

Lubridate's helpers attempt to determine formate once order is specified - then lubridate function will parse the date:
```{r}
#| label: lubridate fuction for dates in strings

ymd("2017-01-31")
# str_view(ymd("2017-01-31"))
# class(ymd("2017-01-31"))

mdy("January 1st, 2017")

dmy("31-Jan-2017")
```

`ymd()` and friends create date. Creating date-time needs underscore and "h", "m", "s" to name of parsing function:
```{r}
#| label: lubridate fuction for dates-time in strings

ymd_hms("2017-01-31 20:11:59")

mdy_hm("01/31/2017 08:01")

# Can also force creatiom of date-time by spplying time zone:
ymd("2017-01-31", tz = "UTC")

```


### From individual components

Date-time components across multiple columns - use `make_date()` for date, `make_datetime()` for date-times:
```{r}
#| label: date-time components across multiple columns

flights |>
  select(year, month, day, hour, minute)

flights |>
  select(year, month, day, hour, minute) |>
  mutate(date = make_date(year = year,
                          month = month,
                          day = day),
         departure = make_datetime(year = year,
                                  month = month,
                                  day = day,
                                  hour = hour,
                                  min = minute)
         )



```

Can do same thing with the four time columns in `flights`: `dep_time`, `sched_dep_time`, `arr_time`, `sched_arr_time`:
```{r}
#| label: Modifying time columns

make_dttm <- function(year, month, day, time){
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights |>
  filter(!is.na(dep_time), !is.na(arr_time)) |>
  mutate(
    dep_time = make_dttm(year, month, day, dep_time),
    sched_dep_time = make_dttm(year, month, day, sched_dep_time),
    arr_time = make_dttm(year, month, day, arr_time),
    sched_arr_time = make_dttm(year, month, day, sched_arr_time)
  ) |>
  select(origin, dest, ends_with("delay"), ends_with("time"))

# Can visualize distribution of departure times across the year:

ggplot(
  data = flights_dt,
  mapping = aes(x = dep_time)
) +
  geom_freqpoly(binwidth = 86400) # 86400 seconds = 1 day

# within single day
flights_dt |>
  filter(dep_time < ymd(20130102)) |> # "2013-01-02" would work as well 
  ggplot(aes(x = dep_time)) +
  geom_freqpoly(binwidth = 600) # 600 s = 10 min
  
```
:::{.callout-note}
After call make_datetime(), the variable is no longer stored as hours and minutes—it becomes a full datetime measured internally in seconds. E.g., `dep_time` is now a POSIXct datetime. Internally, R stores a POSIXct values as: the number of seconds since 1970-01-01 00:00:00 UTC. See `as.numeric(make_dttm(2013, 1, 1, 5, 17))` will return large number which is seconds since 1970. For dates, 1 means 1 day; for date-time, 1 means 1 second.
:::


### From other types

Switching between date-time and date via `as_datetime()` and `as_date()`
```{r}
#| label: as_datetime(), as_date()
# from date to date-time
today()
as_datetime(today())
# from date-time to date
now()
as_date(now())

```


## Date-time components

You can pull out individual parts of the date with the accessor functions `year()`, `month()`, `mday(`) (day of the month), `yday()` (day of the year), `wday()` (day of the week), `hour()`, `minute()`, and `second()`. These are effectively the opposites of `make_datetime()`.
```{r}
#| label: Example date-time components

datetime <- ymd_hms("2026-07-08 12:34:56 ")

year(datetime)
month(datetime, label = TRUE, abbr = FALSE) # optional arguments displayed
day(datetime)

hour(datetime)
minute(datetime)
second(datetime)

mday(datetime)
yday(datetime)
wday(datetime, label = TRUE, abbr = FALSE) # optional arguments displayed


```

```{r}
#| label: Example - using wday() and minutes()

# Example: On what day most flights?
flights_dt |>
  mutate(wday = wday(dep_time, label =TRUE)) |>
  count(wday)

flights_dt |>
  mutate(wday = wday(dep_time, label =TRUE)) |>
  ggplot(mapping = aes(x = wday)) +
  geom_bar()

# Example: Average departure delay by minute
flights_dt |>
  mutate(minute = minute(dep_time)) |>
  group_by(minute) |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n = n()
  ) |>
  ggplot(
    mapping = aes(x = minute, y = avg_delay)
  ) +
  geom_line()


flights_dt |>
  mutate(minute = minute(sched_dep_time)) |>
  group_by(minute) |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n = n()
  ) |>
  ggplot(
    mapping = aes(x = minute, y = avg_delay)
  ) +
  geom_line()

```


### Rounding

Plotting individual components by rounding the date to nearby unit of time - `floor_date()`, `round_date()`, `ceiling_date()` - takes vector of dates to adjust and name of unit to round:
```{r}
#| label: Rounding
# Number of flights per week
flights_dt |>
  count(week = floor_date(dep_time, unit = "week")) |>
  ggplot(mapping = aes(x = week, y = n)) +
  geom_line() +
  geom_point()

# Distribution of flights across course of day; dep_time - earliest instance of thst day:

flights_dt |>
  mutate(dep_hour = dep_time - floor_date(dep_time, unit = "day")) |>
  ggplot(aes(x = dep_hour)) +
  geom_freqpoly(binwidth = 60 * 30)

flights_dt |>
  mutate(dep_hour = hms::as_hms(dep_time - floor_date(dep_time, unit = "day"))) |>
  ggplot(mapping = aes(x = dep_hour)) +
  geom_freqpoly(binwidth = 60 * 30)

```


### Modifying components

Can use accessor functions to modify components of a date/time. Useful when cleaning data that has incorrect dates:
```{r}
#| label: modifying components

(datetime <- ymd_hms("2026-07-08 12:34:56"))

year(datetime) <- 2030

month(datetime) <- 01

hour(datetime) <- hour(datetime) + 1
datetime

# Alterntively use update():

update(datetime, year = 2030, month = 01, mday = 2, hour = 2)

# Big values cause "roll-over"
update(ymd("2023-02-01"), mday = 30)

```

### Exercises

```{r}
#| label: Exercises

# Exercise1: How does distribution of flights within day change over course of year?
flights_dt |>
  filter(!is.na(dep_time)) |>
  mutate(dep_hour = update(dep_time, yday = 1)) |>
  mutate(month = factor(month(dep_time))) |>
  ggplot(aes(dep_hour, color = month)) +
  geom_freqpoly(binwidth = 60 * 60)

# Exercise2: Compare dep_time, sched_dep_time and dep_delay
flights_dt |>
  mutate(dep_delay0 = as.numeric(dep_time - sched_dep_time) / 60) |>
  filter(!is.na(dep_delay)) |>
  mutate(diff = dep_delay - dep_delay0) |>
  ggplot(aes(x = diff)) +
  geom_freqpoly()

# Exercise3: Comparing air time with duration between departure and arrival
flights_dt |>
  filter(!is.na(dep_time), !is.na(arr_time), !is.na(air_time)) |>
  mutate(
    duration = as.numeric(arr_time - dep_time) / 60,
    diff = duration - air_time
  ) |>
  group_by(origin, dest) |>
  summarize(
    avg_duration = mean(duration),
    avg_air_time = mean(air_time),
    avg_diff = mean(diff),
    n = n(),
    .groups = "drop"
  ) |>
  arrange(avg_diff)

# Exercise4: How does the average delay time change over the course of a day?

flights_dt |>
  mutate(sched_dep_hour = hour(sched_dep_time)) |>
  group_by(sched_dep_hour) |>
  summarize(dep_delay = mean(dep_delay, na.rm = TRUE)) |>
  ggplot(aes(x = sched_dep_hour, y = dep_delay)) +
  geom_point() +
  geom_smooth()

# Exercise5: On what day of the week should you leave if you want to minimize the chance of a delay?

flights_dt |>
  mutate(dow = wday(sched_dep_time)) |>
  group_by(dow) |>
  summarize(dep_delay = mean(dep_delay, na.rm = TRUE),
            arr_delay = mean(arr_delay, na.rm = TRUE)) |>
  ggplot(aes(x = dep_delay, y = arr_delay, color = dow, size = dow)) +
  geom_point() +
  scale_color_viridis_c() +
  guides(color = guide_legend(ncol = 2))

```


## Time spans

Arithmetic with dates - subtraction, addition, division.

Three important classes representing time spans:

+ __Durations__ - represent exact number of seconds
+ __Periods__ - represents human units like weeks and months
+ __Intervals__ - represent a starting and ending point


### Durations

In R when subtracting two dates, we get difftime object:
```{r}
#| label: Subtracting dates

h_age <- today() - ymd("2004-03-22")
h_age
```

`difftime` class object record a time span of seconds, minutes, hours, days, or weeks. Ambiguity problematic - lubridate provides alternative which always uses seconds: the duration:
```{r}
#| label: as.duration()

as.duration(h_age)
class(h_age)

```

```{r}
#| label: Different duration constructors
# Different duration constructors
dseconds(15)

dminutes(10)

dhours(c(12, 24))

ddays(0:5)

dweeks(3)

dyears(1)

```

Durations always convert time span in seconds.

```{r}
#| label: Add and multiply durations
# Add and multiple durations:
2 * dyears(1)

dyears(1) + dweeks(12) + dhours(15)

# Add and subtract durations to and from days:
tomorrow <- today() + ddays(1)
tomorrow

last_year <- today() - dyears(1)
last_year

```


### Periods

Are time spans but don't have fixed length in seconds - work with "human" times, like days and months:
```{r}
#| label: Periods

one_am <- ymd_hms("2026-03-08 01:00:00", tz = "America/New_York")

one_am

one_am + days(1)
```

Periods can be created with constructor functions:
```{r}
#| label: constructor functions for periods

hours(c(12, 24))

days(7)

months(1:6)

# Can add and multiply:
10 * (months(6) + days(1))

days(50) + hours(25) + minutes(2)

# Adding to dates

ymd("2024-01-01") + dyears(1)

ymd("2024-01-01") + years(1) # periods more likely to do what expected

# Daylight saving time
one_am + ddays(1)

one_am + days(1)

```

```{r}
#| label: Fixing flights dataset
# Overnight flights odd representation
flights_dt |>
  filter(arr_time < dep_time)

flights_dt |>
  mutate(
    overnight = arr_time < dep_time,
    arr_time = arr_time + days(overnight),
    sched_dep_time = sched_dep_time + days(overnight)
  ) |>
  filter(arr_time < dep_time)
```



### Intervals

Pair of starting and ending date times. or duration with starting point.

Create intervals by writing `start %--% end`:
```{r}
#| label: Creating intervals

y2023 <- ymd("2023-01-01") %--% ymd("2024-01-01")
y2024 <- ymd("2024-01-01") %--% ymd("2025-01-01")

y2023
y2024

# then can divide by days()
y2023 / days(1)
# Leap year:
y2024 / days(1)
# compare to:
years(1) / days(1) 


```

### Exercise

```{r}
#| label: Function returning age

# Example function returning age 

age_comp <- function(year, month, day){
  
  age_seconds <- today() - make_date(year, month, day)
  age_years <- as.numeric(age_seconds) / (years(1) / days(1))
    
  return(age_years)
}

age_comp(1998, 06, 10)


age_date <- function(birthdate_ymd){
  
  age_seconds <- today() - ymd(birthdate_ymd)
  age_years <- as.numeric(age_seconds) / (years(1) / days(1))
    
  return(age_years)
}

age_date("1998-06-10")

```


## Time zones

R uses international IANA time zones - form: `{area}/{location}`, typically as `{continent}/{city}` or `{ocean}/{city}`.

```{r}
#| label: Finding out current time zone

# Current time zone
Sys.timezone()

# Complete list of time zones:
length(OlsonNames())

head(OlsonNames())

```

Time zone is attribute of date-time controlling only printing:
```{r}
#| label: Time zone argument

x1 <- ymd_hms("2024-06-01 12:00:00", tz = "America/New_York")
x1

x2 <- ymd_hms("2024-06-01 18:00:00", tz = "Europe/Copenhagen")
x2

x3 <- ymd_hms("2024-06-02 04:00:00", tz = "Pacific/Auckland")
x3

# They're indeed the same time:

x1 - x2
x1 - x3
```

Lubridate always uses UTC (Coordinated Universal Time) - standard time zone, roughly equivalent to GMT (Greenwich Mean Time).
```{r}
#| label: Compuations woith time zones

# Operations like c() often drop the time zone
# Dates will be displayed with zone of first element

x4 <- c(x1, x2, x3)
x4
```


Can change time zone in two ways:
```{r}
#| label: Changing time zones

# Keep instant in time same, change display
x4a <- with_tz(x4, tzone = "Australia/Lord_Howe")
x4a
x4a - x4
# Change underlying instant in time. Use when instant lablled with incorrect tz
x4b <- force_tz(x4, tzone = "Australia/Lord_Howe")
x4b
x4b - x4
```

