---
title: "Joins"
subtitle: "R for Data Science (2e), Chapter 19"
toc: true
theme: darkly
format: html
---

## Introduction

Two important types of joins:

+ __Mutating joins__ - i.e., adding new variables to one data frame from matching observations in another

+ __Filtering joins__ - i.e., filtering observations from one data frame based on whether or not they match an observation in another.

+ __Keys__ - i.e., variables connecting pair of data frames in a join


### Prerequisites
```{r}
#| label: Prerequisites

library(tidyverse)
library(nycflights13)
```


## Keys

Connecting two tables through pair of keys, within each table. 

### Primary and foreign key

Every join involves pair of keys:

+ __primary keys__: variable or set of variables uniquely identifying each observation (more than one variable: __"compound key"__)
```{r}
#| label: Primary key

# Can identify airline with carrier code, thus carrier -> primary key
airlines 

# Can identify each airport by airport code, thus faa -> primary key
airports

# Can identify planes by tailnum, thus tailnum -> primary key
planes

# Can identify each observation by combination of location and time
# origin, time_hour -> primary key
weather
```

__foreign key__ - variable/set of variables corresponding to primary key in another table:

+ `flights$tailnum` foreign key to primary key `planes$tailnum`

+ `flights$carrier` foreign key to primary key `airlines$carrier`

+ `flights$origin` foreign key to primary key `airports$faa`

+ `flights$dest` foreign key to primary key `airports$faa`

+ `flights$dest` foreign key to primary key `airports$faa`

+ `flights$origin-flights$time_hour` is compound foreign key to primary key `weather$origin-weather$time_hour`


### Checking primary keys

How to verify that the "primary key" variable indeed uniquely identifies each observation?

Use `count()` on primary keys and and look for entries where `n` is greater than 1:
```{r}
#| label: Identifying and verifying primary key

planes |>
  count(tailnum) |>
  filter(n > 1)
# tailnum indeed primary key

weather |>
  count(origin, time_hour) |>
  filter(n > 1)
# origin and time-hour indeed compound key 
```

Also check for missing values - if there is a missing value it can NOT identify an observation!:
```{r}
#| label: Identifying and verifying primary key, check NA

planes |>
  filter(is.na(tailnum))

weather |>
  filter(is.na(origin) | is.na(time_hour))
```


### Surrogate key

Three variables in flight uniquely identify each flight:
```{r}
#| label: Identifying and verifying primary key flights df

flights |>
  count(time_hour, carrier, flight) |>
  filter(n > 1)
```


Might be better off introducing numeric surrogate key using row number:
```{r}
#| label: Surrogate key

flights |>
  mutate(id = row_number(), .before = 1)

```

Useful when communicating to ther humans: much easier saying to look at flight 2001 instead to look on flight UA430 which departet 9am 2013-01-03.


```{r}

weather

airports

```


## Basic joins

Six dplyr join functions:

+ `left_join()`
+ `right_join()`
+ `inner_join()`
+ `full_join()`
+ `semi_join()`
+ `anti_join()`

Syntax: Take pair of data frames (`x`and`y`) and return data frame.
Order of rows and columns in output primarly determined by `x`.


### Mutating joins

__Mutating joins__ allow combining variables from two data frames - first matches observations by keys, then copies across variables from one data frame to the other (like `mutate()` joins add variables to the right):
```{r}
#| label: Narrow dataset

# Narrow dataset for examples

flights2 <- flights |>
  select(year, time_hour, origin, dest, tailnum, carrier)

flights2
```

Four types of mutating joins. Almost always use `left_join()` - output will always have same rows as x, i.e., the data frame we're joining to. Primary use - add additional metadata:
```{r}
#| label: left_join()

# ?left_join: by = NULL; performs natural join, using all common variables across x and y.
flights2 |>
  left_join(airlines) 

# Alternatively, can specify 'by' argument
flights2 |>
  left_join(y = airlines, by = "carrier") 
# airlines$carrier as primary key to foreign key flights$carrier
```

Can also select within `left_join()` -  only adding colunms we want:
```{r}
#| label: left_join() multiple variables

flights |>
  left_join(weather)
# Note: ?left_join - if by argument NULL, *_join() performs natural join
# Message lists variables for us to check
# Here: "Joining with 'by = join_by(year, time_hour, origin)"
# join_by() used when joining multiple variables, e.g. join_by(a == b, c == d)
# If column names same between x and y, list only var name, e.g., join_by(a, c)

# Equivalent:
flights |>
  left_join(y = weather, by = join_by(year == year,
                                      time_hour == time_hour,
                                      origin == origin))
# Equivalent:
flights |>
  left_join(y = weather, by = join_by(year, time_hour, origin))
```

```{r}
#| label: left_join() selecting variables for join from y

# Selecting weather components:
flights2 |>
  left_join(weather |> select(origin, time_hour, temp, wind_speed))

# Equivalently
flights2 |>
  left_join(y = weather |>
              select(origin, time_hour, temp, wind_speed),
            by = join_by(
                         origin == origin,
                         time_hour == time_hour)
            )
# Note: our keys are now 'origin' and 'time_hour'
# Resulting in same data frame, as time_hour already 'contains' year
# Multiple keys possible, but not necessary

# Selecting plane components:
flights2 |>
  left_join(planes)
# Note: Without further selection, joins by tailnum and year

# Equivalently
flights2 |>
  left_join(y = planes |>
              select(tailnum, type, engine, seats), 
            by = join_by(tailnum == tailnum)
            )
# Note: with further selection joins by tailnum only

```

When `left_join()` fails to find match for a row in `x` - fills in new variable with missing values:
```{r}
#| label: left_join() fails finding matches for a row in x

flights2 |>
  filter(tailnum == "N3ALAA") |>
  left_join(planes |> select(tailnum, type, engine, seats))

```


### Specifying join keys

`left_join()` will uses all variables that appear in both data frames as the join key - __natural__ join.

Doesn't always work:
```{r}
#| label: Specifying join keys

flights2 |> 
  left_join(planes)
# Note: Joining with 'by = join_by(year, tailnum)'
# Lot of missing values as year and tailnum used as compound key

# flights$year = year flight occured
# planes$year = year plane was built
# NOT the same

# Thus need specification:
flights |>
  left_join(planes, join_by(tailnum))

```

`year` variables are disambiguated in the output with suffix(`year.x` and `year.y`) - tells us where variables comes from (`x` or `y`). Can override default with `suffix` argument.

`join_by(tailnum)` short for `join_by(tailnum == tailnum)` - Note: This is called an __equi join__ - keys must be equal.

`inner_join()`, `right_join()`, `full_join()` have same interface as `left_join()`. Difference in which row is kept:
+ left join keeps all rows in `x`
+ right join keeps all rows in `y`
+ full join keeps all rows in either `x` or `y`
+ inner join only keeps rows occurring in both `x` and `y`


### Filtering joins

Primary action - filtering rows.

Two types:

+ __Semi-joins__: keeps all rows in `x` that have match in `y` - do NOT add columns from y:
```{r}
#| label: semi_join()

# Filtering airports dataset showing just origin airports
airports |>
  semi_join(flights2, join_by(faa == origin))

# Filtering airports dataset showing just destination airports
airports |>
  semi_join(flights2, join_by(faa == dest))
```

+ __Anti-join__: return all rows in `x` that don't have match in `y` - useful for finding missing values implicit in data:
```{r}
#| label: anti_join()

# Finding airports that are missing from airports but are referenced as destinations in flights2
flights2 |>
  anti_join(airports, join_by(dest == faa)) |>
  distinct(dest)

# Which tailnums are missing from planes
flights2 |>
  anti_join(planes, join_by(tailnum)) |>
  distinct(tailnum)
```

```{r}
#| label: Exercises3

# Ex: Find the 48 hours (over the course of the whole year) that have the worst delays. Cross- reference with weather data

worst_hours <- flights |>
  group_by(origin, time_hour) |>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n_flights = n(),
    .groups = "drop"
  ) |>
  slice_max(avg_delay, n = 48)

worst_hours |>
  left_join(weather, join_by(origin, time_hour)
  ) |>
  select(
    origin,
    time_hour,
    avg_delay,
    n_flights,
    temp,
    wind_speed,
    wind_gust,
    precip,
    visib
  )

# Ex: How can you find all flights to those "top" destinations?
top_dest <- flights2 |>
  count(dest, sort = TRUE) |>
  head(10)

top_dest |> left_join(flights, join_by(dest == dest))
  # self-join to retrieve flight data


# Ex: Every departing flight have corresponding weather data for that hour?
flights |>
  anti_join(weather, join_by(origin, time_hour)) |>
  arrange(time_hour) |>
  relocate(time_hour, .before = dep_time)

# Ex: Tailnums that dont have a match in planes?
flights |>
  anti_join(planes, join_by(tailnum)) |>
  count(carrier, sort = TRUE) |>
  mutate(prop = n / sum(n))

# Ex: Add latitude, longitude of origin and destination airport to flights
flights |>
  left_join(airports |> select(lat_origin = lat, lon_origin = lon, faa), 
            join_by(origin == faa)) |>
  left_join(airports |> select(lat_dest = lat, lon_dest = lon, faa),
            join_by(dest == faa))

# Ex: Compute average delay by destination; join airports data frame; 
# show spatial distribution of delays:

flights |>
  filter(!is.na(dep_delay)) |>
  summarize(avg_delay = mean(dep_delay), .by = dest) |>
  left_join(airports, join_by(dest == faa)) |>
  ggplot(aes(x = lon, y = lat, color = avg_delay)) +
  borders("state") +
  geom_point(na.rm = TRUE) +
  coord_sf()



```


## How do joins work?

Focusing on how each row in `x` matches rows in `y`:
```{r}
#| label: Representative data frame

x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)
```
 
 __Inner-join__: Rows retained if and only if the keys are equal!
```{r}
#| label: How does inner_join() work

inner_join(x, y)
```
 
 __Outer-joins__ follow same principle:
 
 Keep observations that appear in at least one of the data frames.
 Note: Syntax - `*_join(x, y)`
 __Left-join__: Keeps all observations in `x` - "matches" NA row in `y`, when there is no match.
 
 __Right-join__: Keeps all observations in `y` - "matches" NA row in `x`, when there is no match.
 
 __Full join__: Keeps all observations that appear in `x` and `y`, every row will be included - observations with no match, "match" with NA value.
 
 Joins here are called __equi joins__, where rows match if the keys are equal.
 
 
 ### Row matching
 
 Three possible outcomes for a row in `x`:
 
 + Doesn't match anythingm it's dropped
 + Matches 1 row in `y`, it's preserved
 + Matches more than 1 row in `y`, it's duplicated once for each match
 
```{r}
#| label: Matches more than 1 row in y

# Example: matches more than 1 row in y
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     2, "y3"
)

inner_join(x, y) # row is duplicated once for each match
```
 
 
 ### Filtering joins
 
 __Semi join__ keeps rows in `x` that have one or more matches in `y`
 
 __Anti join__ keeps rows in `x` that match zero rows in `y`
 
 In both cases only existence of a match is important; doesn't matter how many times it matches. Filtering join never duplicate rows like mutating joins do.
 
 
 ## Non-equi joins
 
 __Equi joins__: Joins where the rows match if the `x` key equals the `y` key.
 As we move away from equi joins, will display keys as they will differ - dplyr keeps both keys with `keep = TRUE`.
```{r}
#| label: keep = TRUE argument in inner_join()

x |> inner_join(y, join_by(key == key), keep = TRUE)

# Example: Non-equi join
# Match whenever x$key greater than or equal to y$key:
x |> inner_join(y, join_by(key >= key))
```
 
 Four particular useful types of non-equi join:
 
 __Cross joins__ - match every pair of rows
 
 __Inequality joins__ - use `<`, `<=`, `>`, `>=` instead of `==`
 
 __Rolling joins__ - similar to inequality joins but only find closest match
 
 __Overlap joins__ - special inequality join designed to work with ranges
 
 
 ### Cross join
 
Matches everything - useful when generating Cartesian product of rows - ouput has nrow(x) * nrow(y) rows - useful when generating permutations:
```{r}
#| label: cross_join()

df <- tibble(name = c("John", "Simon", "Tracy", "Max"))
df |> cross_join(df) # self-join
```


### Inequality joins

Use `<`, `<=`, `>`, `>=` instead of `==` restricting set of possible matches:
```{r}
df <- tibble(id = 1:4, name = c("John", "Simon", "Tracy", "Max"))

df |> inner_join(df, join_by(id < id))

```


### Rolling joins

Getting closest row satisfying inequality - by adding `closest()`:
```{r}
#| label: Rolling joins 

parties <- tibble(
  q = 1:4,
  party = ymd(c("2022-01-10", "2022-04-04", "2022-07-11", "2022-10-03"))
)

set.seed(123)
employees <- tibble(
  name = sample(babynames::babynames$name, 100),
  birthday = ymd("2022-01-01") + (sample(365, 100, replace = TRUE) - 1)
)
employees

# Find the last party date that comes before (or on) their birthday.

employees |> 
  left_join(parties, join_by(closest(birthday >= party)))

# Problem with this approach: the folks with birthdays before January 10 don’t get a party

employees |> 
  anti_join(parties, join_by(closest(birthday >= party)))

```

Use Overlapping joins


### Overlap joins

Three helpers:

+ `between(x, y_lower, y_upper)` is short for `x >= y_lower`, `x <= y_upper`

+ `within(x_lower, x_upper, y_lower, y_upper)` is short for `x_lower >= y_lower,` `x_upper <= y_upper`.

+ `overlaps(x_lower, x_upper, y_lower, y_upper)` is short for `x_lower <= y_upper`, `x_upper >= y_lower`

```{r}
#| label: Overlapping joins

parties <- tibble(
  q = 1:4,
  party = ymd(c("2022-01-10", "2022-04-04", "2022-07-11", "2022-10-03")),
  start = ymd(c("2022-01-01", "2022-04-04", "2022-07-11", "2022-10-03")),
  end = ymd(c("2022-04-03", "2022-07-11", "2022-10-02", "2022-12-31"))
)
parties

parties |> 
  inner_join(parties, join_by(overlaps(start, end, start, end), q < q)) |> 
  select(start.x, end.x, start.y, end.y)

# fix problem

 parties <- tibble(
  q = 1:4,
  party = ymd(c("2022-01-10", "2022-04-04", "2022-07-11", "2022-10-03")),
  start = ymd(c("2022-01-01", "2022-04-04", "2022-07-11", "2022-10-03")),
  end = ymd(c("2022-04-03", "2022-07-10", "2022-10-02", "2022-12-31"))
)
 
 employees |> 
  inner_join(parties, join_by(between(birthday, start, end)), unmatched = "error")

```

