---
title: "Logical vectors"
subtitle: "R for Data Science (2e), Chapter 12"
author: "Filip Kukic"
toc: true
theme: darkly
format: html
---

## Logical Vectors

### Introduction

Logical vectors can only be: `TRUE`, `FALSE`, `NA`.

Simple way of creating logical vectors $\rightarrow$ numeric comparisons

Two useful functions $\rightarrow$ `ìf_else()` and `case_when()`.

### Prerequisites

:::{.callout-note}
Most function in chapter provided by base R, but will still load tidyverse to use `filter()` and `mutate()`.
:::

```{r}
#| label: Loading packages

library(tidyverse)
library(nycflights13) # will draw examples from 'flights' dataset

# Or will make up some dummy data
x <- c(1, 2, 3, 5, 7, 11, 13)
x * 2
```

:::{.callout-note}
Any manipulation to free-floating vector, can be done to variable in dataframe with `mutate()` and friends.
:::

```{r}
#| label: Manipulating variables in df via mutate()

df <- tibble(x)

df |>
  mutate(y = x * 2)

```


### Comparison

Creating logical vectors via numeric comparison with: `<`,`<=`,`>`,`>=`,`!=`,`==`.

So far created via `filter()` siletly.

```{r}
#| label: Example - filter()

# Example - Finding daytime departures arriving on time
flights |>
  filter(dep_time > 600 & dep_time < 2000 & abs(arr_delay) < 20)

# Can explicitely create these logical vectors with mutate()
flights |>
  mutate(
    daytime = (dep_time > 600) & (dep_time < 2000),
    approx_ontime = abs(arr_delay) < 20,
    .keep = "used"
  ) |>
  filter(daytime & approx_ontime)
```


### Floating point comparison

```{r}
#| label: Example - floating point comparison

x <- c(1 / 49 * 49, sqrt(2) ^ 2)
x

# Testing for equality results in FALSE:
x == c(1, 2)

# Reason:
print(x, digits = 16) # R is rounding these numbers by default!

# Alternatively use dyplr::near() ignoring small differences
near(x, c(1, 2))

# Base R: 
all.equal(x, c(1, 2))
```


### Missing values

:::{.callout-warning}
Almost any operation involving unknown value will be unknow.
:::

```{r}
#| label: Example - missing values NAs

NA > 5

10 == NA

NA == NA
# Why is this last one true?
# Example: 
age_mary <- NA
age_john <- NA
age_mary == age_john

# Thus following does not work:
flights |> 
  filter(dep_time == NA) # yields NA for every single row
# Solution. is.na()
```


### is.na()

`is.na(x)` works with any type of vector and returns `TRUE` for missing values and `FALSE` for everything else:

```{r}
#| label: Examples - is.na()

is.na(c(TRUE, NA, FALSE))

is.na(c(1, NA, 3))

is.na(c("a", NA, "b"))

# Can find rows with missing dep_time:
flights |>
  filter(is.na(dep_time))

# Useful with arrange(); default: puts NA values at end; can override:
flights |>
  filter(month == 1, day == 1) |>
  arrange(dep_time)

flights |>
  filter(month == 1, day == 1) |>
  arrange(desc(is.na(dep_time)), dep_time) # display NA first

flights |>
  mutate(
    missing_dep_time = is.na(dep_time),
    missing_sched_dep_time = is.na(sched_dep_time),
    missing_dep_delay = is.na(dep_delay),
    .keep = "all"
  ) |> 
  count(
    missing_dep_time,
    missing_sched_dep_time, 
    missing_dep_delay
    )

```


### Boolean algebra

Can combine multiple vectors using Boolean algebra:

`&`"and",`|`"or",`!`"not",`xor()`"exclusive or".

:::{.callout-note}
`xor()` is true if x is true, or if y is true, but not both.
:::

#### Missing values

```{r}
#| label: Example - missing values

df <- tibble(x = c(TRUE, FALSE, NA))

df |>
  mutate(
    and = x & NA,
    or = x | NA
  )
```

Missing value in logical vector can be `TRUE` or `FALSE` - we don't know.
+ `NA | TRUE`: Must be `TRUE`(either or is true)
+ `NA | FALSE`: Is `NA` - don't know if `NA` is `TRUE` or `FALSE`.
+ `NA | NA`: is `NA`

+ `NA & TRUE`: is `NA`- don't know if `NA` is `TRUE` or `FALSE`
+ `NA & FALSE`: is `FALSE` - at least one condition is `FALSE`
+ `NA & NA`: is `NA`


### Order of operations

```{r}
#| label: Example - order of operations

flights |>
  filter(month == 11 | month == 12)

# Does NOT work:
flights |>
  filter(month == 11 | 12)

# No error, as everything apart from 0 converted to TRUE
flights |> 
  mutate(
    nov = month == 11,
    final = nov | 12,
    .keep = "used"
  )
```


### %in%

Way to avoid `==` and `|` misordering by using `%in%`. `x %in% y` returns logical vector same length as `x` that is `TRUE` whenever a value in `x` si anywhere in `y`.

```{r}
#| label: Example - %in%

1:12 %in% c(1, 5, 11)

# Example: Finding all flights in November and December
flights |>
  filter(month %in% c(11, 12))

# Different rules for NA:
c(1, 2, NA) == NA # all NA

c(1, 2, NA) %in% NA

# Useful shortcut:
flights |> 
  filter(dep_time %in% c(NA, 0800))
```


## Summaries

### Logical summaries

Two main logical summaries: `any()` and `all()`.

`any(x)` is equivalent to `|` $\rightarrow$ returns `TRUE` if there any `TRUE`'s in `x`.

`all(x)` is equivalent to `&` $\rightarrow$ returns `TRUE` if all values of `x` are `TRUE`.

Most summary function include `na.rm = TRUE`.

```{r}
#| label: Example - all(), any()

flights |>
  group_by(year, month, day) |>
  summarize(
    all_delayed = all(dep_delay <= 60, na.rm = TRUE), 
    any_long_delay = any(arr_delay >= 300, na.rm = TRUE),
    .groups = "drop"
  )
  
```


### Numeric summaries of logical vectors

Logical vectors is numeric context, `TRUE` becomes 1, `FALSE` becomes 0
$\rightarrow$ `sum()` and `mean()` useful with logical vectors: 

+ `sum(x)`: gives number of `TRUE`s
+ `mean(x)`: gives __proportion__ of `TRUE`s

```{r}
#| label: Example - sum() and mean() with logical vectors

flights |> 
  group_by(year, month, day) |>
  summarize(
    proportion_delayed = mean(dep_delay <= 60, na.rm = TRUE),
    count_long_delay = sum(arr_delay >= 300, na.rm = TRUE), 
    .groups = "drop"
  )
```


### Logical subsetting

Can use logical vector to filter single variable to a subset of interest:

```{r}
#| label: Example - logical subsetting
# Avg. delay of flights, that were actually delayed
flights |>
  filter(arr_delay > 0) |>
  group_by(year, month, day) |>
  summarize(
    behind = mean(arr_delay),
    n = n(),
    .groups = "drop"
  )

# How to also compute avg. delay for flights arriving early?
# Use inline filtering with [
flights |>
  group_by(year, month, day) |>
  summarize(
    behind = mean(arr_delay[arr_delay > 0], na.rm = TRUE),
    ahead = mean(arr_delay[arr_delay < 0], na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
# Note: n() now return total number of flights
```


### Conditional transformation

Two important tools: `if_else()` and `case_when()`.

#### if_else()

:::{.callout-note}
Main advantage to base R's `ìfelse()` is that with `dyplr::if_else()` can choose what happens to missing values, and it's more likely to give meaningful error when variable types not compatible.
:::

Format: `if_else(condition, true, false, missing)`

```{r}
#| label: Example - if_else()

x <- c(-3:3, NA)
if_else(x > 0, "+ve", "-ve")

# How to deal with NA
if_else(x > 0, "+ve", "-ve", "???")

# can also use vectors for true and false
if_else(x > 0, -x, x)

# Can also implement simple coalesce() version
x1 <- c(NA, 1, 2, NA)
y1 <- c(3, NA, 4, 6)
if_else(is.na(x1), y1, x1) # replacing NA values

# Can also nest if_else() in if_else()
if_else(x == 0, "0", if_else(x < 0, "-ve", "+ve"), "?")
```

:::{.callout-tip}
With more conditions, switch to `dyplr::case_when()`.
:::


### case_when()

Different computations for different conditions. Syntax: `condition ~ output`.
`condition` must be a logical vector - when `TRUE`, `output` will be used:

```{r}
#| label: Example - nested if_else() with case_when()

case_when(
  x == 0 ~ "0",
  x > 0 ~ "+ve",
  x < 0 ~ "-ve",
  is.na(x) ~ "???"
)

# Can also specify default for all other values:
case_when(
  x < 0 ~ "-ve",
  x > 0 ~ "+ve",
  .default = "???"
)

# Note: if multiple conditions match, only first will be used:
case_when(
  x > 0 ~ "+ve",
  x > 2 ~ "big"
)
```

Can also use variables on both sides of `~`; can mix and match variables as needed.

```{r}
#| label: Example - case_when() and mutate()

flights |>
  mutate(
    status = case_when(
      is.na(arr_delay) ~ "cancelled",
      arr_delay < -30 ~ "very early",
      arr_delay < -15 ~ "early",
      abs(arr_delay) <= 15 ~ "on time",
      arr_delay < 60 ~ "late",
      arr_delay < Inf ~ "very late"
    ),
    .keep = "used"
  )
```


### Compatible types

`if_else()` and `case_when()` require __compatible__ types in output:

+ Numerical and logical vectors are compatible
+ strings and factors are compatible
+ Dates and date-times are compatible
+ NA is compatible with everything

