---
title: "Missing values"
subtitle: "R for Data Science (2e), Chapter 18"
toc: true
theme: darkly
format: html
---

### Prerequisites
```{r}
#| label: Prerequisites

library(tidyverse)
```


## Explicit missing values

Explicit values - i.e., cells with `NA`.


### Last observation carried forward

Common use for missing values is as data entry convenience. Missing values, when entry by hand, indicate that previous value should be carried forward, i.e., repeated:
```{r}
#| label: Carrying forward values by NA values

# Carrying forward values by NA values
treatment <- tribble(
  ~person, ~treatment, ~response,
  "Derrick Whitemore", 1, 7,
  NA, 2, 10, 
  NA, 3, NA, 
  "Katherine Burke", 1, 4
)

# treatment |>
#   fill(person, treatment, response)

treatment |>
  fill(everything())

# locf = "last observation carried forward"

```


### Fixed values

Sometime `NA` represent fixed, known values, most commonly 0. Can use `dplyr::coalesce()`:
```{r}
#| label: Fixed values

x <- c(1, 4, 5, 7, NA)
coalesce(x, 0)

# Can specify NA values when reading in CSV data
# read_csv(file = path, na = "99"), specify all "99" values as NA

# Alternatively use na_if():
x <- c(1, 4, 5, 7, -99)
na_if(x, -99)

```


### NaN

Special type of NA value: `NaN` (not a number). Behaves like `NA` values:
```{r}
#| label: NaN values

x <- c(NA, NaN)
x * 10 

x == 1

is.na(x)
is.nan(x)

na_if(c(TRUE, 99), x)
```

Encounter `NaN` when performing math:
```{r}
#| label: Encountering NaN

0 / 0

0 * Inf

Inf - Inf

sqrt(-1)
```


## Implicit missing values

`NA` values that cannot be seen:
```{r}
#| label: Implicit missing values

stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```

+ `price` in 4th quarter of 2020 - `NA` - __explicit__
+ `price` 1st quarter of 2021 - does not appear in dataset- __implicit__


### Pivoting

Making implicit missing values explicit (vice versa) by pivoting:
```{r}
#| label: Pivoting and missing values

stocks |>
  pivot_wider(names_from = qtr,
              values_from = price
              )

stocks |>
  pivot_wider(names_from = qtr,
              values_from = price
              )

```


### Complete

tidyr::complete() allows generating missing values by providing set of variables defining combination of rows that should exist:
```{r}
#| label: complete()

stocks |>
  complete(year, qtr)

# ?complete: ... Specification of columns to expand or complete; 
# Finds all unique combinations of x, y, and z, including those not
# present in the data.

# Can also provide own data, when variables incomplete themselves

stocks |>
  complete(year = 2019:2021, qtr)
```

Range of variable correct, but not all values present, use `full_seq(x, 1)` generating all values from `min(x)` to `max(x)` spaced by 1.

Alternatively create dataframe containig all rows that should exist, then combine with original dataset with `dplyr::full_join`.


### Joins

Revealing implicitely missing observations via __joins__. Can often only know what values are missing from dataset when comparing to another.

`dplyr::anti_join(x, y)` useful here - selects rows in `x` that dont have match in `y`:
```{r}
#| labe: anti_join()

library(nycflights13)

flights |>
  distinct(faa = dest) |>
  anti_join(airports) # missing information for four airports in flights

flights |>
  distinct(tailnum) |>
  anti_join(planes) # missing information for 722 planes in flights

```

### Exercise

```{r}
#| label: Exercise1

# Relationship between carrier and rows missing from planes?
flights |>
    anti_join(planes, by = "tailnum") |>
    count(carrier, sort = TRUE)

```



## Factors and empty groups

Group not containing any observations - can arise when working with factors:
```{r}
#| label: Empty groups

health <- tibble(
  name   = c("Ikaia", "Oletta", "Leriah", "Dashay", "Tresaun"),
  smoker = factor(c("no", "no", "no", "no", "no"), levels = c("yes", "no")),
  age    = c(34, 88, 75, 47, 56),
)

health |>
  count(smoker) # smoker groups is empty, but we know they exist

levels(health$smoker)

# Can request count() to keep all groups using .drop = FALSE

health |>
  count(smoker, .drop = FALSE) # now count for smoker is included
  

```

Same applies to ggplot2's discrete axis - drop levels that don't have any values - can force them to display supplying `drop = FALSE`:
```{r}
#| label: Keeping all levels

ggplot(health, aes(x = smoker)) + 
  geom_bar() +
  scale_x_discrete()

ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

Same problem and solution with `group_by()`:
```{r}
#| label: Missing levels when grouping

health |>
  group_by(smoker, .drop = FALSE) |>
  summarize(,
            n = n(),
            mean_age = mean(age),
            min_age = min(age),
            max_age = max(age),
            sd_age = sd(age)
            ) # But do have implicit missing values
# can make them explicit
health |>
  group_by(smoker) |>
  summarize(,
            n = n(),
            mean_age = mean(age),
            min_age = min(age),
            max_age = max(age),
            sd_age = sd(age)
            ) |> # But do have implicit missing values
  complete(smoker) # But count is now NA instead of 0 (drawback of approach)

```

