---
title: "Numbers"
subtitle: "R for Data Science (2e), Chapter 12"
author: "Filip Kukic"
toc: true
theme: darkly
format: html
---

### Prerequisites

Mostly functions from base R; will still need tidyverse, as we'll use base functions inside functions like `mutate()` anf `filter()`.

```{r}
#| label: Prerequisites

library(tidyverse)
library(nycflights13)

```


## Making numbers

Most cases, numbers will be recorded in R's numeric type: integer or double. Sometimes as strings. readr provides useful functions parsing strings into numbers: `parse_double()` and `parse_number()`.

Use `parse_double()` when numbers have been written as strings:
```{r}
#| label: Example - parse_double()

x <- c("1.2", "5.6", "1e3")
class(x)


parse_double(x)
class(parse_double(x))

```

Use `parse_number()` when strings contain non-numeric text that we want to ignore. Useful for currency data and percentages:
```{r}
#| label: Example - parse_number()

x <- c("$1,234", "USD 3,513", "59%")
class(x)

parse_number(x)
class(parse_number(x))

```


## Counts

`count()` is great for quick exploration and checks during analysis:
```{r}
#| label: Example - count()

flights |> count(dest)

# Most common values can be retrieved by sort = TRUE:
flights |> count(dest, sort = TRUE)

# Can perform same computation by hand:
flights |>
  group_by(dest) |>
  summarize(
    n = n(),
    delay = mean(arr_delay, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  arrange(desc(n))

# n() takes no argument, accesses information about current 

# If we want to see all values: |> print(n = Inf)
```

Couple variants of `n()` and `count()`:
+ `n_distinct()`: counts number of distinct (unique) values of one or more variables
```{r}
#| label: Example - n_distinct()

flights |>
  group_by(dest) |>
  summarize(
    carriers = n_distinct(carrier)
  ) |>
  arrange(desc(carriers))
```

+ A weigthed count is a sum
```{r}
#| label: Example - weighted count

flights |>
  group_by(tailnum) |>
  summarize(miles = sum(distance))

# Thus count has wt argument:
flights |>
  count(tailnum, wt = distance)
```

+ Can count missing values - combining `sum()` and `is.na()`
```{r}
#| label: Example - counting missing values

flights |>
  group_by(dest) |>
  summarize(n_cancelled = sum(is.na(dep_time))) |>
  arrange(desc(n_cancelled))
```


## Numeric transformations

Transformation function work well with `mutate()` - as output is same length as input. Most transformation functions built in base R - will list only most practical one.

### Arithmetic and recycling rules

Basics of arithmetic (`x`, `-`, `*`, `/`, `^`). __Recycling rules__ determine what happens with different size sides.

$\rightarrow$ R handles mismatched lengths by recycling (repeating) the short vector:
```{r}
#| label: Example - Recycling

x <- c(1, 2, 10, 20)
x / 5
# is shorthand for:
x / c(5, 5, 5, 5)

```

Want to recycle single numbers - BUT R will recylce any shorter length vector:
```{r}
#| label: Example2 - Recycling

x * c(1, 2)

x * c(1, 2, 3)

```
Recycling rules apply to logical comparisons (`==`, `<`, `<=`, `>`, `>=`, `!=`):
```{r}
#| label: Example - Recycling logical vector

flights |>
  filter(month == c(1, 2)) # not correct; Use %in%!

flights |>
  filter(month %in% c(1, 2))
```


### Minimum and Maximum

Arithmetic functions work onm pairs, e.g., `pmin()` and `pmax()` - when given two or more variables, returns smallest or largest value in each row:
```{r}
#| label: Example - pmin() and pmax()

df <- tribble(
  ~x, ~y,
  1, 3, 
  5, 2, 
  7, NA,
)

df |>
  mutate(
    min = pmin(x, y, na.rm = TRUE),
    max = pmax(x, y, na.rm = TRUE)
  )

# Note: summary functions min() and max() return single value from multiple observations

df |> 
  mutate(
    min = min(x, y, na.rm = TRUE),
    max = max(x, y, na.rm = TRUE)
  )
```


### Modular arithmetic

I.e. division that yields a whole number and a remainder. 

`%/%` = integer division
`%%` = computes remainder
```{r}
#| label: Example - Modular arithmetic

1:10 %/% 3 

1:10 %% 3

flights |>
  mutate(
    hour = sched_dep_time %/% 100,
    minute = sched_dep_time %% 100,
    .keep = "used"
  )

flights |>
  group_by(hour = sched_dep_time %/% 100) |>
  summarize(prop_cancelled = mean(is.na(dep_time)),
            n = n()) |> 
  filter(hour > 1) |>
  ggplot(
    mapping = aes(x = hour, y = prop_cancelled)
  ) +
  geom_line(color = "gray50") +
  geom_point(aes(size = n))
```


### Logarithms

Used when dealing with data ranging across multiple orders and converting exponetial growth to linear growth.

Three choices: `log()`, `log2()`(base 2), `log10()`(base 10).

Inverse of `log()` is `exp()`.


### Rounding

`round(x)`round number to nearest integer.
```{r}
#| label: Example - round()

round(123.456)

```

Can control precision of rounding with second argument, `digits`, i.e. `round(x, digits)`.

```{r}
#| label: Example - precision of rounding

round(123.456, digits = 2)

round(123.456, digits = 1)

round(123.456, digits = -1) # rounds to nearest ten

round(123.456, digits = -2) # rounds to nearest hundred

round(123.456, digit = -3) # rounds to nearest thousand

# Banker's rounding, i.e. "round half to even":
round(c(1.5, 2.5))

# floor() and ceiling()
x <- 123.456

floor(x)

ceiling(x)

# Rounding down to nearest two digits
floor(x / 0.01) * 0.01

# Rounding up to nearest two digits
ceiling(x / 0.01) * 0.01

# Round to nearest multiple of 4
round(x / 4) * 4

round(x / 0.25) * 0.25
```


### Cutting numbers into ranges

Use `cut()` to break up (aka bin) numeric vector into discrete buckets:

```{r}
#| label: Example - cut()

x <- c(1, 2, 5, 10, 15, 20)

cut(x, breaks = c(0, 5, 10, 15, 20))

# breaks don't need evenly spacing
cut(x, breaks = c(0, 5, 10, 100))

# can supply labels
cut(x, 
  breaks = c(0, 5, 10, 15, 20), 
  labels = c("sm", "md", "lg", "xl")
)

# Values outside the range:
y <- c(NA, -10, 5, 10, 30)
cut(y, breaks = c(0, 5, 10, 15, 20))

```


### Cummulative and rolling aggregates

Base R: `cumsum()`, `cumprod()`, `cummin()`, `cummax()`
`dplyr::cummean()` provides cumulative means.

```{r}
#| label: Example - cumulative functions

x <- 1:10
cumsum(x)

```


## General transformations

### Ranks

See documentation for: `dplyr::min_rank()`, `dplyr::row_number()`, `dplyr::dense_rank()`

```{r}
#| label: Example - *_rank() functions

x <- c(1, 5, 5, 17, 22, NA)

min_rank(x) # gives tie same smallest value

dense_rank(x) # works like min_rank() but leaves no gap

row_number(x) # gives every input unique rank
# equivalent to
rank(x, ties.method = "first", na.last = "keep") # see documentation


min_rank(desc(x)) # largest values get smallest rank

# Example with df
df <- tibble(x = x)

df |>
  mutate(
    row_number = row_number(x),
    dense_rank = dense_rank(x),
    percent_rank = percent_rank(x),
    cume_dist = cume_dist(x)
  )

```

:::{.callout-note}
Many of these results can be achieved by picking appropriate `ties.method` argument for base R `rank()`. Set `na.last = "keep"` to keep `NA` as `NA`.
:::

`row_number()` can also be used without any arguments when inside dplyr verb. Useful with %% and %/% dividing in similar sized groups:
```{r}
#| label: Example - row_number()

df <- tibble(id = 1:10)

df |>
  mutate(
    row0 = row_number() - 1,
    three_groups = row0 %% 3, 
    three_in_each_group = row0 %/% 3
  )

```


### Offsets

`dplyr::lag()` and `dplyr::lead()` allow to refer to values before or after "current" value - returns vector of same length:
```{r}
#| label: Example - lag() and lead()

x <- c(2, 5, 11, 11, 19, 35)

lag(x)

lead(x)

# Difference between current and previous value
x - lag(x)

# When does current value change?
x == lag(x)

```

:::{.callout-note}
Call lead or lag by more than one position by using second argument, `n`.
:::


### Consecutive identifiers

E.g., breaking events into sessions - begin new session after some gap
```{r}
#| label: Example - cons identifiers

events <- tibble(
  time = c(0, 1, 2, 3, 5, 10, 12, 15, 17, 19, 20, 27, 28, 30)
)

events <- events |>
  mutate(
    diff = time - lag(time, default = first(time)),
    has_gap = diff >= 5,
    group = cumsum(has_gap) # allows us now to group_by
  )

events

# Another grouping approach
# df with repeating values
df <- tibble(
  x = c("a", "a", "a", "b", "c", "c", "d", "e", "a", "a", "b", "b"),
  y = c(1, 2, 3, 2, 4, 1, 3, 9, 4, 8, 10, 199)
)

df |> 
  group_by(id = consecutive_id(x)) |>
  slice_head(n = 1)
```

```{r}
#| label: Useful Example
# Useful Example
# All destinations that are flown by at least two carriers
# Relative ranking of the carriers based on their performance for the same destination

flights |>
  filter(n_distinct(carrier) > 2, .by = dest) |>
  summarize(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    n = n(),
    .by = c(dest, carrier)
  ) |>
  mutate(
    rank = min_rank(avg_dep_delay),
    .by = dest) |>
  arrange(dest, rank)
  
  
```


## Numeric summaries

### Center
Alternativ to `mean()` is function `median()` - finding values in middle of vector. 

For symmetric distribution, rather report mean, for skewed distributions usually report median.

```{r}
#| label: Example - median

flights |>
  group_by(year, month, day) |>
  summarize(
    mean = mean(dep_delay, na.rm = TRUE),
    median = median(dep_delay, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(
    mapping = aes(x = mean, y = median)
  ) +
  geom_abline(slope = 1, intercept = 0, color = "white", linewidth = 2) +
  geom_point()
```


### Minimum, maximum, and quantiles

Other locations:
+ `min()`, `max()` - giving largest and smallest values.
+ `quantile()` - Syntax: `quantile(x, 0.25)`, finds `x` greater than 25% of the values. Note: `median(x) == quantile(x, 0.5)`.

```{r}
#| label: Example - quantile(

flights |>
  group_by(year, month, day) |>
  summarize(
    max = max(dep_delay, na.rm = TRUE),
    q95 = quantile(dep_delay, 0.95, na.rm = TRUE),
    .groups = "drop"
  )
```


### Spread

Common summaries: 
+ `sd()` - standard deviation 
+ `IQR()` - inter-quartile range, i.e., `quantile(x, 0.75) - quantile(x, 0.25)`

```{r}
#| label: Example - spread

flights |>
  group_by(origin, dest) |>
  summarize(
    distance_iqr = IQR(distance), # Every distance the same for every flight thus IQR = 0? Not quite!
    n = n(),
    .groups = "drop"
  ) |>
  filter(distance_iqr > 0)
```


### Distributions

Visualize distribution before comitting to summary statistics - avoid missing important insights by choosing wrong summary

E.g., plot shows 365 frequency polygons of dep_delay, one for each day, are overlaid. Distribution seems to follow common pattern - it is fine to use same summary for each day:
```{r}
#| label: Example - visualizing distribution

flights |>
  filter(dep_delay < 120) |> 
  ggplot(aes(x = dep_delay, group = interaction(day, month))) + 
  geom_freqpoly(binwidth = 5, alpha = 1/5)
```


### Positions

Final type of summary - Extracting values at specific positions:
`first(x)`, `last(x)`, `nth(x, n)`.

```{r}
#| label: Example - positions

flights |>
  group_by(year, month, day) |>
  summarize(
    first_dep = first(dep_time, na_rm = TRUE),
    fifth_dep = nth(dep_time, 5, na_rm = TRUE),
    last_dep = last(dep_time, na_rm = TRUE)
  )

# Note: dplyr functions use _ to separate components of functions
# Ever need [ function?

flights |> 
  group_by(year, month, day) |>
  mutate(r = min_rank(sched_dep_time)) |>
  filter(r %in% c(1, max(r)))
```


### With mutate()

Summary functions typically paired with `summarize()`. Because of recycling, can be paired with `mutate()`.

+ `x / sum(x)`: computes proportion of total
+ `(x - mean(x)) / sd(x)`: computes Z-score
+ `(x - min(x)) / (max(x) - min(x))`: standardize to range [0, 1]
+ `x / first(x)`: compute index based on first observation.


