---
title: "Strings"
subtitle: "R for Data Science (2e), Chapter 14"
toc: true
theme: darkly
format: html
---


### Prerequisites
```{r}
#| label: Prerequisites

library(tidyverse)
library(babynames)

```


:::{.callout-tip}
__stringr__ functions start with `str_` - will trigger autocomplete, can be useful.
:::


## Creating a string

```{r}
#| label: Creating strings

string1 <- "This is a sting"

string2 <- 'Include "quote" with single quotes'
```


### Escapes

```{r}
#| label: Escaping special characters

double_quote <- "\"" # or '"'
single_quote <- "\'" # or "'"
backslash <- "\\"

# can see content with str_view()
x <- c(double_quote, single_quote, backslash)
str_view(x)

```


### Raw strings

```{r}
#| label: Raw strings


tricky <- "double_quote <- \"\\\"\" # or '\"' single_quote <- '\\'' # or \"'\""
str_view(tricky)

# Use raw string 
tricky <- r"(double_quote <- "\"" # or '"' single_quote <- '\'' # or "'")"
str_view(tricky)

```


### Other special characters

Most common: `\n` (new line) and `\t` (tab). See complete list under `?Quotes`.

```{r}
#| label: Special characters

x <- c("one\ntwo", "one\ttwo", "\u00b5", "\U0001f604")
x
str_view(x)

```


## Creating many strings from data 

### str_c()

Takes any number of vectors as arguments and returns a character vector:
```{r}
#| label: str_c()

str_c("x", "y")

str_c("x", "y", "z", sep = " ")

str_c("Hello ", c("John", "Susan"))

```

Documentation: `str_c()` similar to `paste0()` - BUT designed to be used with `mutate()`:
```{r}
#| label: str_c() and mutate()

df <- tibble(name = c("Flora", "David", "Terra", NA))

df |> mutate(greeting = str_c("Hi ", name, "!"))

```

Missing values can be displayed in another way using `coalesce()` to replace them: 
```{r}
#| label: Using coalesce()
# Documentation 'Value' - Vector with same type and size
df |>
  mutate(
    greeting1 = str_c("Hi ", coalesce(name, "you"), "!"),
    greeting2 = coalesce(str_c("Hi ", name, "!"), "Hi!")
  )
```


### str_glue()

Mixing many fixed and variable strings with str_c() requires typing a lot of `"`s. Alternative: Use __glue package__ via `str_glue()` - special feature: Anything inside `{}` will be evaluated like it's outside of quotes (see `?str_glue`):
```{r}
#| label: str_glue

df |>
  mutate(greeting = str_glue("Hi {name}"))


```


### str_flatten()

`str_flatten()` works with `summarize()` - takes character vector and combines each element of the vector into single string:
```{r}
#| label: str_flatten()

str_flatten(c("x", "y", "z"))

str_flatten(c("x", "y", "z"), collapse = ", ")

str_flatten(c("x", "y", "z"), collapse = ", ", last = ", and ")


# Works well with summarize:

df <- tribble(
  ~ name, ~ fruit,
  "Carmen", "banana",
  "Carmen", "apple",
  "Marvin", "nectarine",
  "Terence", "cantaloupe",
  "Terence", "papaya",
  "Terence", "mandarin"
)

df |> 
  group_by(name) |>
  summarize(fruits = str_flatten(fruit, collapse = ", "))

```


## Extracting data from strings

Common that multiple variables getting crammed together into a single string.

Four different tidyr functions for extraction:
+ `df |> separate_longer_delim(col, delim)`
+ `df |> separate_longer_position(col, width)`
+ `df |> separate_wider_delim(col, delim, names)`
+ `df |> separate_wider_position(col, widths)```

Similar with `pivot_longer()` and `pivot_wider()` - make input data frame longer creating new rows and wider by generating new columns.

`delim` splits up string with delimiter like `", "` or `" "`.; `position` splits at specific widths, like `c(3, 5, 2)`.


### Separating into rows

Useful when number of components varies from row to row. Most common case is splitting based on delimiter:
```{r}
#| label: separate_longer_delim(col, delim)

df1 <- tibble(x = c("a,b,c", "d,e", "f"))

df1 |>
  separate_longer_delim(x, delim = ",")

df2 <- tibble(x = c("1211", "131", "21"))

df2 |>
  separate_longer_position(x, width = 1)

```

### Separating into columns

Useful when fixed numbers of components in each string, and want to spread into columns:
```{r}
#| label: separate_wider_delim(col, delim, names)

df3 <- tibble(x = c("a10.1.2022", "b10.2.2011", "e15.1.2015"))

df3 |>
  separate_wider_delim(x,
                       delim = ".",
                       names = c("code", "edition", "year")
                       )

# Can also omit specific piece by NA

df3 |>
  separate_wider_delim(x,
                       delim = ".",
                       names = c("code", NA, "year")
                       )

# separate_wider_position() needs specifying width of each column

df4 <- tibble(x = c("202215TX", "202122LA", "202325CA"))

df4 |>
  separate_wider_position(
    x,
    widths = c(year = 4, age = 2, state = 2)
  )

```


### Diagnosing widening problems

```{r}
#| label: separate_wider_delim() arguments

df <- tibble(a = c("1-1-1", "1-1-2", "1-3", "1-3-2", "1"))

#df |>
#  separate_wider_delim(
#    a,
#    delim = "-",
#    names = c("x", "y", "z")
#  )

 df |>
  separate_wider_delim(
    a, delim = "-", names = c("x","y","z"),
    too_few = "debug"
  )
```
 
```{r}
#| label: Too few
# Too few pieces
df |>
  separate_wider_delim(
    a, delim = "-", names = c("x","y","z"),
    too_few = "align_start"
  )

df |>
  separate_wider_delim(
    a, delim = "-", names = c("x","y","z"),
    too_few = "align_end"
  )

df |>
  separate_wider_delim(
    a, delim = "-", names = c("x","y","z"),
    too_few = "debug"
  )
```

```{r}
#| label: Too many
# Too many pieces

df <- tibble(a = c("1-1-1", "1-1-2", "1-3-5-6", "1-3-2", "1-3-5-7-9"))

df |>
  separate_wider_delim(a, delim = "-",
                       names = c("x", "y", "z"),
                       too_many = "drop")

df |>
  separate_wider_delim(a, delim = "-",
                       names = c("x", "y", "z"),
                       too_many = "merge")

```


## Letters

`str_length()` tells number of letters in strings
```{r}
#| label: Number of strings

str_length(c("a", "R for data science", NA))
```

```{r}
#| label: using count() and str_length()

babynames |>
  count(length = str_length(name), wt = n)

# Note: equivalent to
babynames |>
  mutate(length = str_length(name)) |>
  group_by(length) |>
  summarize(n = sum(n))

# Can use filter() to look at longest name
babynames |>
  filter(str_length(name) == 15) |>
  count(name, wt = n, sort = TRUE)

```


### Subsetting

Extract parts of string via `str_sub(string, start, end)`.
+ `start` and `end` are position where substring starts and ends (inclusive arguments). Length of returned string: `end - start + 1`:
```{r}
#| label: str_sub(string, start, end)

x <- c("Apple", "Banana", "Pear")

str_sub(x, start = 1, end = 3)

# Can use negative values to count back from the end of the string:
# -1 is last character, -2 is second to last character

str_sub(x, start = -3, end = -1)

# Too short string; str_sub returns as much as possible

str_sub("a", start = -3, end = -1)

babynames |>
  mutate(
    first = str_sub(name, start = 1, end = 1),
    last = str_sub(name, start = -1, end = -1)
  )

# Example: Extracting middle letter; when even, extract left letter
babynames |>
      mutate(
        length = str_length(name),
        middle_length = if_else(
          length %% 2 == 0,
          length / 2,
          round(length / 2)),
        middle_letter = str_sub(
          name, start = middle_length, end = middle_length
        )
      )

```


## Non-English text

### Encoding 

Underlying representation of strings in R accessed via `charToRaw()`:
```{r}
#| label: charToRaw()

charToRaw("Hadley")

# Each represents one letter
# mapping from hexadecimal numbers to character is called encoding
# Encoding is ASCII

```
For more read http://kunststube.net/encoding/.

### Letter variations

```{r}
#| label: Letter variations

u <- c("\u00fc", "u\u0308")
str_view(u)

str_length(u)

str_sub(u, 1, 1)


u[[1]] == u[[2]]


str_equal(u[[1]], u[[2]])


```

