---
title: "base-r"
subtitle: "R for Data Science (2e), Chapter 26"
toc: true
theme: darkly
format: html
---

## Introduction

tidyverse is not only way to solve data science problems - it is not possible to use tidyverse without base R.

### Prerequisites

```{r}
#| label: Prereq

# Will load tidyverse for the sake of comparing approaches to base R
library(tidyverse)
```

## Selecting mutliple elements with \[

`[` - extracts sub-components from vectors and data frames - e.g., `x[i]`, `y[i, j]`.

### Subsetting vectors

Five main things that can subset a vector with - i.e., can be the `i` in `x[i]`

1.  **vector of positive integers**

```{r}
#| label: Subsetting vector with positive integers

x <- c("one", "two", "three", "four", "five")
x[c(3, 2, 5)]

# can make longer output by repeating positions
x[c(3, 2, 5, 3, 2, 4)]
```

2.  **vector of negative integers**

```{r}
#| label: Subsetting vector with negative integers

x[c(-3,-2,-5)]
```

3.  **logical vector**

```{r}
#| label: Subsetting with logical vector

x[c(TRUE, TRUE, FALSE)] # recycled the vector as length of logical vector ≠ length of vector x

x[!is.na(x)] # subsetting all non-missing values


y <- c(10, 3, NA, 5, 8, 1, NA)
y[y %% 2 == 0]
```

Unlike `filter()`, `NA` values will be included in the output.

4.  **character vector**

```{r}
#| label: Subsetting with character vector

x <- c(abc = 1, def = 2, xyz = 5)

x[c("abc", "xyz")]

# can also use character vectors to duplicate individual entires
```

5.  **nothing**

```{r}
#| label: Subseeting with nothing

# return complete x
x[]

```

### Subsetting data frames

`df[rows, cols]` - can only select specific rows or cols while keeping all other cols and rows `df[, cols]` and `df[rows, ]`.

```{r}
#| label: Subsetting data frames - examples

df <- tibble(
  x = 1:3,
  y = c("a", "b", "c"),
  z = runif(3)
)

df

# Selecting first row and second column
df[1, 2]

# Selecting all rows and columns x and y
df[, c("x", "y")]

# Selecting rows where "x" is greater than 1
df[df$x > 1, ]
```

Base R `data.frame` (similar to tidyverse `tibble`- though some behavior makes life easier) - returns vector if df\[, cols\] selects single single column, returns data frame when selecting multiple columns.

Note: if `df` is a tibble, then `[` will return a tibble.

```{r}
#| label: Subsetting data frames in base R

df1 <- data.frame(x = 1:3)
df1
df1[, "x"] # returns vector as single column selected

# Can avoid ambiguity by specifying drop = FALSE

df1 <- data.frame(x = 1:3)
df[, "x", drop = FALSE] # now return data frame instead of vector

# comparing to tidyverse "tibble"
df2 <- tibble(x = 1:3)
df[, "x"] # tibble returns tibble

```

### dplyr equivalents

Several dplyr verbs are special cases of `[`:

- `filter()`:

```{r}
#| label: filter() verb as [

df <- tibble(
  x = c(2, 3, 1, 1, NA),
  y = letters[1:5],
  z = runif(5)
)

df |> 
  filter(x > 1)

# equivalent to base R:

df[!is.na(df$x) & df$x > 1, ]

# alternatively use which() - dropping the missing values

df[which(df$x > 1), ]

```

- `arrange()`:

```{r}
#| label: Subsetting rows with integer vector - arrange()

# arrange() is subsetting with integer vector - usually using order()

df |>
  arrange(x, y)

# equivalent to base R:

df[order(df$x, df$y), ]
```

- `select()`, `relocate()`:

```{r}
#| label: Base R version of select() and relocate()

# Similar to subsetting the columns with a character vector
df |> 
  select(x, y)

# base R alternative:
df[c("x", "y")]
```

Base R also combines futures of `filter()` and `select()` - `subset()`:

```{r}
#| label: Using subset()

df |>
  filter(x > 1) |>
  select(y, z)

# Base R alternative:
df |>
  subset(subset = x > 1, select = c(y, z))
```

### Exercises

```{r}
#| label: exercises

# ex1: 
# a. function taking a vector as input and returning elements at even-numbered positions
even_elmts <- function(vec) {
  vec[1:length(vec) %% 2 == 0]
}

vec <- sample(x = 1:10, size = 13, replace = TRUE)
vec

even_elmts(vec = vec)

# b. every element except the last one
all_elmts_exlo <- function(vec) {
  vec[1:(length(vec)-1)]
}

vec
all_elmts_exlo(vec)

# c. only even values (and no missing values)
even_vals <- function(vec) {
  vec[which(vec %% 2 == 0)]
}

vec <- c(vec, NA)
vec

even_vals(vec)


# ex2: Read ?which; Why is not x[-which(x>0)] same as x[x <= 0]

```

## Selecting single element with \$ and \[\[

Used to pull columns out of data frames.

### Data frames

\[\[ can access by position or by name; \$ accesse by name.

```{r}
#| label: extracting columns by position, by name

tb <- tibble(
  x = 1:4, 
  y = c(10, 4, 1, 21)
)

tb

# by position
tb[[1]]
# by name
tb[["x"]]

# by name ($)
tb$x

# can also be used to create new columns
tb$z <- tb$y + tb$x
tb

```

Several other approaches in base R to create new columns: `transform()`, `with()`, `within()`.

https://gist.github.com/hadley/1986a273e384fb2d4d752c18ed71bedf

`$` convenient when performing quick summaries:

```{r}
#| label: Using $ when performing quick summaries

max(diamonds$carat)

levels(diamonds$cut)

# dplyr alternative to [[/$:
diamonds |>
  pull(carat) |>
  max()

diamonds |>
  pull(cut) |>
  levels()
```

### Tibbles

Important differences between tibbles and base `data.frame` when it comes to `$`. Data frames match prefix of any variable name (**partial matching**) - does not complain if column does not exist:

```{r}
#| label: Partial matching

df <- data.frame(x1 = 1)

df$x # partially matches the prefix

df$z
```

Tibbles more strict - only match variable names exactly - generate a warning if column trying to access does not exist:

```{r}
#| label: Partial matching - cont.

tb <- tibble(x1 = 1)

tb$x # no "partial matching"; looks for exact match

tb$x1 # only works with exact match

```

### Lists

```{r}
#| label: Differences btw [ and [[ w lists

l <- list(
  a = 1:3,
  b = "a string",
  c = pi,
  d = list(-1, -5)
)

```

- `[` extracts sub-list - result will always be a list:

```{r}
#| label: Extracting a sub-list

str(l[1:2])

str(l[1])

str(l["d"]) # can subset with logical, integer, or character

```

- `[[` and `$` extracts single component from a list - remove a level of hierarchy from the list:

```{r}
#| label: Extracting a component

str(l[[1]])

str(l[[4]])

str(l$a)
```

::: callout-important
`[` retruns a new, smaller list.

`[[` drills down into the list.
:::

## Apply family

Base equivalents for tidyverse techniques for iteration like `dplyr::across()` and map family of functions.

Most important member of apply family - `lapply()` - similar to `purrr::map()`.

`dplyr::across()` can be achieved using `lapply()` with `[`:

```{r}
#| label: apply family

df <- tibble(
  a = 1,
  b = 2,
  c = "a",
  d = "b",
  e = 4
)

df

# example apply family function
# sapply(), simplifies results, thus "s", (compare to map_vec())
num_cols <- sapply(df, is.numeric)

# transform each column with lapply() then replace original values

df[, num_cols] <- lapply(df[, num_cols, drop = FALSE], \(x) x * 2)

df
```

Stricter version of `sapply()` is `vapply()` - vector apply - using additional argument specifying expected type, ensuring simplification occurs same way regardless of input:

```{r}
#| label: apply family - vapply()

# Call above can be done by specifying type:
vapply(df, is.numeric, logical(1))
# we expect is.numeric() to return a logical vector of length 1, thus logical(1).

```

Another important member: `tapply()` - computes single grouped summary:

```{r}
#| label: tapply()


diamonds |>
  group_by(cut) |>
  summarize(price = mean(price))

tapply(diamonds$price, diamonds$cut, mean) # see ?tapply documentation
# tapply returns results in named vector
```

`apply()` works with matrices and arrays.

## for **loops**

Building block of iteration used by apply and map families under the hood:

```         
for (element in vector) {
  # do something with element
}
```

Use of `for` loops is to achieve same effect as `walk()` - call some function with side-effect on each element of a list:

```         
paths |> walk(append_file)

Could use for loop:

for(path in paths) {
  append_file(path)
}
```

Slightly harder to save output of the `for` loop:

```{r}
#| label: Example for loops
#| eval: false


# Example:
paths <- dir("data/gapminder", pattern = "\\.xlsx$", full.names = TRUE)

files <- map(paths, readxl::read_excel)

# Alternatively using for loop iteration:

# Define upfront what output is going to look like
files <- vector("list", length(paths))

# Iterate over indices, instead over the elements
for (i in seq_along(paths)) {
  files[[i]] <- readxl::read_excel(paths[[i]])
}

# Combining list of tibbles into single tibble using do.call() + rbind()
do.call(rbind, files)
```


## Plots

```{r}
#| label: Examples of base plot functions

# Examples of base plot functions
hist(diamonds$carat)

plot(diamonds$carat, diamonds$price)

```

