---
title: "iteration"
subtitle: "R for Data Science (2e), Chapter 26"
toc: true
theme: darkly
format: html
---

### Prerequisites

```{r}
#| label: Prereq

library(tidyverse)
```

## Modifying multiple columns

```{r}
#| label: Modifying multiple columns

set.seed(1950)
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df |> 
  summarize(
    n = n(),
    a = median(a),
    b = median(b),
    c = median(c),
    d = median(d)
  )

# Instead of copying, can use across():

df |> 
  summarize(
    n = n(),
    across(.cols = a:d, .fns = median)
  )

```

`across()` has three arguments:

### .cols

First argument in `across()` - `.cols` selects columns to transform. Same specification as `select()`, i.e., can use `starts_with()` and `ends_with()`.

Two additional selection techniques: `everything()` and `where()` -

```{r}
#| label: across() - .cols

set.seed(1014)

df <- tibble(
  grp = sample(2, 10, replace = TRUE),
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df |>
  group_by(grp) |>
  summarize(across(everything(), median))

```

`where()` - allows to select columns based on type:

- `where(is.numeric)` - selects all numeric columns
- `where(is.character)` - selects all string columns
- `where(is.Date)` - selects all date columns
- `where(is.POSIXct)` - selects all date-time columns
- `where(is.logical)`- selects all logical columns

Can combine these with Boolean algebra, e.g., `!where(is.numeric)` or `starts_with("a") & where(is.logical)`.

### Calling a single function

Second argument defines how each column will be transformed:

```{r}
#| label: across() calling a function - second arg

df |>
  group_by(grp) |>
  summarize(across(everything(), median))
```

### Calling multiple functions

```{r}
#| label: Example - calling multiple functions

set.seed(1950)

# Random permutation of NA values
rnorm_na <- function(n, n_na, mean = 0, sd = 1) {
  sample(x = c(
    rnorm(n - n_na, mean = mean, sd = sd),
    rep(x = NA, times = n_na)
  ))
}

df_miss <- tibble(
  a = rnorm_na(5, 1),
  b = rnorm_na(5, 1),
  c = rnorm_na(5, 1), 
  d = rnorm(5)
)

df_miss |>
  summarize(
    across(.cols = a:d, .fns = median), # returns NA
    n = n() 
  )

# create function with desired argument na.rn = TRUE to median():
df_miss |>
  summarize(
    across(.cols = a:d, .fns = function(x) median(x, na.rm = TRUE)),
    n = n()
  )

# Alternatively use shortcut for anonymous function replacing function with \:
df_miss |>
  summarize(
    across(.cols = a:d, .fns = \(x) median(x, na.rm = TRUE))
  )

# across() effectively expands to the following code:
df_miss |> 
  summarize(
    a = median(a, na.rm = TRUE),
    b = median(b, na.rm = TRUE),
    c = median(c, na.rm = TRUE),
    d = median(d, na.rm = TRUE),
    n = n()
  )

# Can also supply another function by using named list:

df_miss |>
  summarize(
    across(
      .cols = a:d,
      .fns = list(
        median = \(x) median(x, na.rm = TRUE),
        n_miss = \(x) sum(is.na(x))
      )
    ),
    n = n()
  )
```

Columns are named using glue specification like `{.col}_{.fn}` - can usen `.names` argument to supply own glue spec.

### Column names

```{r}
#| label: Specifying names in across()

df_miss |>
  summarize(
    across(
      .cols = a:d,
      .fns = list(
        median = \(x) median(x, na.rm = TRUE),
        n_miss = \(x) sum(is.na(x))
      ),
      .names = "{.col}_{.fn}"
    ),
    n = n()
  )
```

`.names`argument important when using `across()` with `mutate()`. Output of `across()` is given same name as the inputs - `across()` inside `mutate()` will replace existing columns:

```{r}
#| label: Example across() with mutate()

df_miss
# Example, using coealesce to replace NA values
df_miss |>
  mutate(across(
    .cols = a:d, 
    .fns = \(x) coalesce(x, 0) # mutate() w. across() overrides names
  ))

# If want new columns, use .names argument to give output new names
df_miss |>
  mutate(
    across(
      .cols = a:d,
      .fns = \(x) coalesce(x, 0),
      .names = "{.col}_na_zero"
    )
  )
```

### Filtering

`across()` - great match for `summarize()` and `mutate()`.

`across()` - awkward for `filter()` - dplyr provides two variants of `across()`: `if_any()` and `if_all()`:

```{r}
#| label: Using if_any() and if_all()

# same as df_miss |> filter(is.na(a) | is.na(b) | is.na(c) | is.na(d))
df_miss |> filter(if_any(a:d, is.na))

# same as df_miss |> filter(is.na(a) & is.na(b) & is.na(c) & is.na(d))
df_miss |> filter(if_all(
  .cols = a:d, 
  .fns = is.na
))
```

### across() in functions

```{r}
#| label: across() in functions

# Examples

# expand_dates <- function(df) {
#   df |>
#     mutate(
#       across(
#         .cols = where(is.Date),
#         .fns = list(year = \(x) year(x), month = \(x) month(x), day = \(x) day(x)))
#     )
# }

expand_dates <- function(df) {
  df |>
    mutate(
      across(
        .cols = where(is.Date),
        .fns = list(year = year, month = month, day = day))
    )
}

df_date <- tibble(
  name = c("Amy", "Bob"),
  date = ymd(c("2009-08-03", "2010-01-16"))
)

df_date |>
  expand_dates()
```

Can also supply multiple columns in a single argument in `function()` because first argument uses tidy-select - just need to embrace argument:

```{r}
#| label: Multiple columns in function() with across() inside

# Exmaple: Function computes means of numeric columns by default - but by supplying second argument can choose to summarize just selected columns

summarize_mean <- function(df, summary_vars = where(is.numeric)) {
  df |>
    summarize(
      across(
        .cols =  {{ summary_vars }}, 
        .fns = \(x) mean(x, na.rm = TRUE)
      ),
      n = n(),
      .groups = "drop"
    )
}

diamonds |>
  group_by(cut) |>
  summarize_mean()

diamonds |>
  group_by(cut) |>
  summarize_mean(c(carat, x:z))
```

### Compare with pivot_longer()

Same calculations performed:

```{r}
#| label: pivot_longer() vs. across()

df |>
  summarize(
    across(a:d, list(median = median, mean = mean))
  )

long <- df |>
  pivot_longer(cols = a:d) |>
  group_by(name) |>
  summarize(
    median = median(value),
    mean = mean(value)
  )

long

# Can form it to across() , by using pivot_wider():
long |>
  pivot_wider(names_from = name,
              values_from = c(median, mean),
              names_vary = "slowest",
              names_glue = "{name}_{.value}") # see '?pivot_wider' for args
```

Useful when problem not solvable with `across()`:

```{r}
#| label: Switching from across() to pivot_*() example

set.seed(1950)
df_paired <- tibble(
  a_val = rnorm(10),
  a_wts = runif(10),
  b_val = rnorm(10),
  b_wts = runif(10),
  c_val = rnorm(10),
  c_wts = runif(10),
  d_val = rnorm(10),
  d_wts = runif(10)
)

df_paired |>
  pivot_longer(everything(),
               names_to = c("groups", ".value"),
               names_pattern = "(.)_(.*)"
  )

# Alternatively:
df_long <- df_paired |>
  pivot_longer(everything(),
               names_to = c("groups", ".value"),
               names_sep = "_"
  )

df_long

result <- df_long |>
  group_by(groups) |>
  summarize(mean = weighted.mean(val, wts))

result

# Can call pivot_wider() to bring into original form:
result |>
  pivot_wider(names_from = "groups", values_from = "mean",
              names_glue = "{groups}_{.value}",
              names_sep = "_")

```

### Exercises

```{r}
#| label: Exercises - modifying columns

# Ex1.1
# Computing the number of unique values in each column of palmerpenguins::penguins
palmerpenguins::penguins |>
    summarize(across(everything(), n_distinct))

# Ex1.2
# Computing the mean of every column in mtcars
mtcars |>
  summarize(across(everything(), \(x) mean(x, na.rm = TRUE)))
# Alternatively, after checking mtcars |> filter(if_any(everything(), is.na))
mtcars |>
  summarize(across(everything(), mean))

# Ex1.3
# Grouping diamonds by cut, clarity, and color then counting the number of observations and computing the mean of each numeric column.
diamonds |>
  group_by(cut, clarity, color) |>
  summarize(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)),
            n = n(),
            .groups = "drop")


# Ex2: What happens if you use a list of functions in across(), but don’t name them? How is the output named?
mtcars |>
  summarize(across(everything(), list(\(x) mean(x, na.rm = TRUE),
                                      \(x) median(x, na.rm = TRUE))))
# Chooses to split the variable name

# Ex3: Adjust expand_dates() to automatically remove the date columns after they’ve been expanded. Do you need to embrace any arguments?
expand_dates <- function(df) {
  df |> 
    mutate(
      across(where(is.Date), list(year = year, month = month, day = mday))
    ) |>
    select(-date) # no embracing needed
}


df_date <- tibble(
  name = c("Amy", "Bob"),
  date = ymd(c("2009-08-03", "2010-01-16"))
)


df_date |> 
  expand_dates()

# Ex4: Explain what each step of the pipeline in this function does. What special feature of where() are we taking advantage of?
show_missing <- function(df, group_vars, summary_vars = everything()) {
  df |> 
    group_by(pick({{ group_vars }})) |> 
    summarize(
      across({{ summary_vars }}, \(x) sum(is.na(x))),
      .groups = "drop"
    ) |>
    select(where(\(x) any(x > 0)))
}

# group_by(pick({{ group_vars }})): groups the data by the specified variables.
# summarize(across(...)): for each group, counts the number of missing (NA) values in each selected summary column.
# select(where(\(x) any(x > 0))): removes columns that contain only zeros, keeping only columns that have at least one missing value in some group.
# Special feature of where() being used here is:
# where() allows to select columns based on a condition evaluated on the entire column, not just by their names or types.

```

## Reading multiple files

`purrr::map()` - function for manipulating all files in a directory.

```{r}
#| label: Example - purrr::map() 
#| eval: false

# Example - reading in directory full of excel spreadsheets
# Option1: read in individually:
data2019 <- readxl::read_excel("data/y2019.xlsx")
data2020 <- readxl::read_excel("data/y2020.xlsx")
data2021 <- readxl::read_excel("data/y2021.xlsx")
data2022 <- readxl::read_excel("data/y2022.xlsx")
# Then combine vie dplyr::bind_rows():
data <- bind_rows(data2019, data2020, data2021, data2022)

```

These kind of tasks require automation - three steps:

1.  `list.files()` - lists all files in a directory.
2.  Use `purrr::map()` - reads each of them into a list
3.  Use `purrr::list_rbind()` - combining them into single data frame

(Note: Approach not always possible, especially with increasing heterogeneity)

### Listing files in directory

`list.files()` - three arguments: + `path` - directory to look in + `pattern` - regular expression used to filter file names - most common pattern: `[.]xlsx$` or `[.]csv$` to find all files with specific extensions + `full.names` determines whether or not directory name should be included in output - almost always want `TRUE`.

```{r}
#| label: Using list.files()

# Example: Using list.files()
paths <- list.files("data/gapminder", pattern = "[.]xlsx$", full.names = TRUE)
paths
```

### Lists

```{r}
#| label: Reading lists
#| eval: false

# reading lists individually:
# gapminder_1952 <- readxl::read_excel("data/gapminder/1952.xlsx")
# gapminder_1957 <- readxl::read_excel("data/gapminder/1957.xlsx")
# gapminder_1962 <- readxl::read_excel("data/gapminder/1962.xlsx")
#  ...,
# gapminder_2007 <- readxl::read_excel("data/gapminder/2007.xlsx")

# Instead: Put all in one obejct - a list:
files <- list(
  readxl::read_excel("data/gapminder/1952.xlsx"),
  readxl::read_excel("data/gapminder/1957.xlsx"),
  readxl::read_excel("data/gapminder/1962.xlsx"),
  readxl::read_excel("data/gapminder/1972.xlsx"),
  readxl::read_excel("data/gapminder/1967.xlsx")
)

# Can extract the ith element by files[[i]]:
files[[3]]
```

### purrr::map() and list_rbind()

`map()` similar to `across()` - difference, it does something to each element of a vector (Note: `across()` does something to each column in a data frame).

`map(x, f)` - shorthand for:

```         
list(
 f(x[[1]]),
 f(x[[2]]),
 ...,
 f(x[[n]])
)
```

Can use `map()` to get list of 12 data frames:

```{r}
#| label: Using map()

files <- map(paths, readxl::read_excel)
length(files)

files[[1]]
```

Note: Easier with `View()` than `str()` to inspect.

Can use `purrr:::list_rbind()` to combine list of data frames into a single data frame:

```{r}
#| label: Using list_rbind()

list_rbind(files)

# Can do both steps in one:
paths |>
   map(readxl::read_excel) |>
   list_rbind()
   
# For extra arguments use same technique as with across:
# Example:
paths |>
   map(\(path) readxl::read_excel(path, n_max = 1)) |>
   list_rbind()
```

### Data in the path

File name contains year - not recorded in individual files. Two things to get year column into the data frame:

1. Name vector of paths - using `set_names()` (can take function) - use `basename()` extracting file name from the full path:
```{r}
#| label: Extracting file name

paths |> set_names(basename)

# names carried along by map functions - list of data frames will have those same names:
files <- paths |> set_names(basename) |>
  map(readxl::read_excel)

# Can use [[]] to extract elemnts by name
files[["1952.xlsx"]]
```

Shorthand for:
```
files <- list(
  "1952.xlsx" = readxl::read_excel("data/gapminder/1952.xlsx"),
  "1957.xlsx" = readxl::read_excel("data/gapminder/1957.xlsx"),
  "1962.xlsx" = readxl::read_excel("data/gapminder/1962.xlsx"),
  ...,
  "2007.xlsx" = readxl::read_excel("data/gapminder/2007.xlsx")
)

# Can also extract names via [[:
files[["1962.xlsx"]]
```

```{r}
#| label: Extracting names of files

paths |>
  set_names(basename) |>
  map(readxl::read_excel) |>
  list_rbind(names_to = "year") |>
  mutate(year = parse_number(year))


# Alternatively, if other variables stored in directory name, use set_names() without argument for full path name, and separate_wider_delim() and friends

paths |>
  set_names() |>
  map(readxl::read_excel) |>
  list_rbind(names_to = "dir") |>
  separate_wider_delim(cols = "dir", delim = "/", names = c(NA, "dir", "file")) |>
  separate_wider_delim(cols = "file", delim = ".", names = c("file", "ext"))
```

### Saving the work

```{r}
#| label: Saving the work

gapminder <- paths |>
  set_names(basename) |>
  map(readxl::read_excel) |>
  list_rbind(names_to = "year") |>
  mutate(year = parse_number(year))

write.csv(gapminder, file = "gapminder.csv")
  
```


### Many simple iterations

Options: doing one single round of iterations, or, multiple rounds of iterations.

```
process_file <- function(path){
    df <- read_csv(path)
    
    df |>
       filter(!is.na(id)) |>
       mutate(id = tolower(id)) |>
       pivot_longer(jan:dec, names_to = "month")
}

paths |>
    map(process_file) |>
    list_rbind()
```

Alternatively, perform each step of `process_file()` to every file:

```
paths |>
    map(read_csv) |>
    map(\(df) df |> filter(!is.na(id))) |>
    map(\(df) df |> mutate(id = tolower(id))) |>
    map(\(df) df |> pivot_longer(jan:dec, names_to = "month")) |>
    list_rbind()
```

Or binding all together earlier:

```
paths |>
   map(read_csv) |>
   list_rbind() |>
   filter(!is.na(id)) |>
   mutate(id = tolower(id)) |>
   pivot_longer(jan:dec, names_to = "month")
```

### Heterogeneous data

Hetergeneous data might to lead to `list_rbind()` to fail, previously calling `map()`. Still useful to load all files:
```{r}
#| label: heterogeneous data

files <- paths |>
  map(readxl::read_excel)

# Handy df_types function, exploring df:

df_types <- function(df) {
  tibble(
    col_name = names(df),
    col_type = map_chr(df, vctrs::vec_ptype_full),
    n_miss = map_int(df, \(x) sum(is.na(x)))
  )
}

df_types(gapminder)
```

Can apply this function to all files, and do some pivoting to see the differnces between files:
```{r}
#| label: example df_types

files |>
  map(df_types) |>
  list_rbind(names_to = "file_name") |>
  select(-n_miss) |>
  pivot_wider(names_from = col_name, values_from = col_type)
```
Read about `map_if()`, `map_at()` - allows modifying elements of list based on their values, and based on their names.

### Handling failures

`map()` either successfully reads all files in a directory or fails with an error. Solution to this is using `purrr:possibly()` - function operator: takes function and returns function with modified behavior:
```{r}
#| label: handling failures

files <- paths |>
  map(possibly(\(path) readxl::read_excel(path), otherwise = NULL))

data <- files |>
  list_rbind() # automatically ignores NULL's 

data

# Can also inspect the files that failed - starting by getting paths that failed

failed <- map_vec(files, is.null)
failed
paths[failed]

```

### Writing to a database

Sometimes not possible to fit in data into memory - when working with a lot of files - cannot do `map(files, read_csv)`. One approach - load data into database, access only bits needed with **dbplyr**.

Some data packages will contain a function that takes a vector of paths and loads them all into the database - case with `duckdb_read_csv()`:
```
con <- DBI::dbConnect(duckdb::duckdb())
duckdb::duckdb_read_csv(conn = con, name = "gapminder", paths)
```

Helpful if we work with csv files - if we work with excel spreadsheets, need to do it by hand.

Start by creating a table that we fill in with data - create a template - a dummy data frame containing all columns we want, but only sampling of data:
```{r}
#| label: Creating a template

template <- readxl::read_excel(paths[[1]])

# adding year
template$year <- 1952
```

Now use `DBI::dbCreateTable()` to turn template into database table:
```{r}
#| label: Connecting to db and creating table 

# connecting to databse
con <- DBI::dbConnect(duckdb::duckdb())

DBI::dbCreateTable(conn = con, name = "gapminder", fields = template)
```
`dbCreateTable()` does not use data in `template`, only uses variable names and types - `gapminder` will be empty but it has he variables needed with expected types:
```{r}
#| label: Inspecting gapminder

con |>
  tbl("gapminder")
```

Next - function that takes single file path, reads into R, adds result to `gapminder` table -  do that by combining `read_excel()` with `DBI::dbAppendTable()`: 
```{r}
#| label: Function for reading in files

append_file <- function(path){
  df <- readxl::read_excel(path)
  df$year <- parse_number(basename(path))
  
  DBI::dbAppendTable(con, "gapminder", df)
}
```


Now call `append_file()` once for each element of `paths` - use `map()`:
```{r}
#| label: Apply map() with append_file

paths|>
  map(append_file)

# walk() - does exactly the same thing as map() but throws the output away:

paths |>
  walk(append_file)
```

can see if we have all the data in our table:
```{r}
#| label: Apply walk() with append_file

# checking output
con |>
  tbl("gapminder") |>
  count(year)

```

### Writing csv files

Same principles apply when writing to mulitple csv files, one for each group. 

Example: Want to take `ggplot2::diamonds` data and save one csv file for each `clarity` - first make individual datasets: 
```{r}
#| label: Using group_nest()

by_clarity <- diamonds |>
  group_nest(clarity)

by_clarity

by_clarity$data[[1]]

# creating column giving name of output file - using mutate() and str_glue():

by_clarity <- by_clarity |>
  mutate(path = str_glue("diamonds-{clarity}.csv"))
  
```

now data contains one tibble for each unique value of `clarity` and we have created a column with name of output file.

Saving data frames by hand:
```
write.csv(by_clarity$data[[1]], by_clarity$path[[1]])
write.csv(by_clarity$data[[2]], by_clarity$path[[2]])
...
write.csv(by_clarity$data[[8]], by_clarity$path[[8]])

```

Note: `map()` only allows the first argument to change - can use `map2()` for first two arguments to change - or, if we do not care about ouput, can use `walk2()`:
```{r}
#| label: Having two arguments to map() and walk()

# map2(by_clarity$data, by_clarity$path, write_csv)

walk2(by_clarity$data, by_clarity$path, write_csv)
```


### Saving plots

Function drawuing the plot:
```{r}
#| label: Creating a function for drawing plots

carat_histogram <- function(df) {
  ggplot(df, aes(x = carat)) + geom_histogram(binwidth = 0.1)
}

carat_histogram(by_clarity$data[[1]])
```

can use `map()` to create list of many plots and eventual file paths:
```{r}
#| label: Creating many files

by_clarity <- by_clarity |>
  mutate(
    plot = map(data, carat_histogram),
    path = str_glue("clarity-{clarity}.png")
         )

by_clarity

# now use walk() and ggsave() to save each plot

walk2(
  by_clarity$path,
  by_clarity$plot,
  \(path, plot) ggsave(path, plot, width = 6, height = 6)
)
```

This is shorthand for:
```
ggsave(by_clarity$path[[1]], by_clarity$plot[[1]], width = 6, height = 6)
ggsave(by_clarity$path[[2]], by_clarity$plot[[2]], width = 6, height = 6)
...
ggsave(by_clarity$path[[8]], by_clarity$plot[[8]], width = 6, height = 6)

```

