---
title: "communication"
subtitle: "R for Data Science (2e), Chapter 10"
author: "Filip Kukic"
toc: true
theme: darkly
format: html
---

### Introduction

Goal: Communicating understandings to others

$\rightarrow$ make plots as self-explanatory as possible

### Prerequisites

```{r}
#| label: prerequisities
# Note: Install first
library(tidyverse)
library(scales)
library(ggrepel)
library(patchwork)

```

### Labels

Add labels with `labs()` function:

```{r}
#| label: Adding labels

ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point(
    mapping = aes(color = class)
  ) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Car type",
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```
:::{.callout-tip}
Avoid titles that just describe what the plot is, e.g., "A scatterplot of engine displacement vs. fuel economy". `caption` often used to describe source of data.
:::

Also can use mathematical equations instead of text strings by replacing `""` with `quote` (see also `?plotmath`). 

```{r}
#| label: Using math equations as labs

df <- tibble(
  x = 1:10,
  y = cumsum(x^2)
)

ggplot(
  data = df,
  mapping = aes(x = x, y = y)
) +
  geom_point() +
  labs(
    x = quote(x[i]),
    y = quote(sum(x[i]^2, i == 1, n))
  )
```

### Annotations

Often useful to label individual observations or groups of observations.

$\rightarrow$ `geom_text()` is similar to `geom_point()`, BUT has additional aesthetic: `label`.

```{r}
#| label: Adding labels to plot via geom_text()

# First pull out cars with highest engine size per drive type
label_info <- mpg |>
  group_by(drv) |>
  arrange(desc(displ)) |>
  slice_head(n = 1) |>
  mutate(
    drive_type = case_when(
      drv == "f" ~ "front-wheel drive",
      drv == "r" ~ "rear-wheel drive",
      drv == "4" ~ "4-wheel drive"
    )
  ) |>
  select(displ, hwy, drv, drive_type)

label_info

# Use new data frame to label the three groups
ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy, color = drv)
) + 
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_text(
    data = label_info, 
    mapping = aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", # size, fontface customizes look of text labels
    size = 5,
    hjust = "right", # horizontal justification
    vjust = "bottom" # vertical justification, controlling alignment of labels
  ) +
  theme(legend.position = "none") # turning all legends off

```

Can use `geom_label_repel()` function from ggrepel package to address labels overlapping with each other and with point: 

```{r}
#| label: Adressing overlapping label with ggrepel

ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy, color = drv)
) + 
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_label_repel(
    data = label_info,
    mapping = aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, nudge_x = 1, nudge_y = 3 # Adjust looks and position
  ) +
  theme(legend.position = "none")

# Can also use to highlight certain points on a plot
potential_outliers <- mpg |>
  filter(hwy > 40 | (hwy > 20 & displ > 5))

ggplot(
  data = mpg,
  mapping = aes(x = displ, hwy)
) + 
  geom_point() +
  geom_text_repel(
    data = potential_outliers,
    mapping = aes(label = model)
  ) +
   geom_point(
     data = potential_outliers,
     color = "red"
   ) +
  geom_point(
    data = potential_outliers,
    color = "red",
    size = 3,
    shape = "circle open"
  )

```

:::{.callout-tip}
There are many more geoms available in ggplot2 for annotations, besides `geom_text()` and `geom_label()`:

+ `geom_hline()` and `geom_vline()` adds reference lines. (Usually `linewidth = 2` and `color = "white"` to not draw attention away from data)

+ `geom_rect()` draws rectangles around points of interest. Boundaries defined by `xmin`,`xmax`,`ymin`,`ymax`.

+ `geom_segment()` with `arrow` arguments drwas attention to a point with an arrow. Aesthetics `x`, `y`, `xend`, `yend`to define start and end locations.
:::

Another function for adding annotations is `annotate()`

:::{.callout-tip}
Rule of thumb:

$\rightarrow$ Use geoms when highlighting subset of data.

$\rightarrow$ Use `annotate()` when adding one of few annotation elements to a plot
:::

```{r}
#| label: Example - annotate()

trend_text <- "Larger engine sizes tend to have lower fuel economy." |>
  str_wrap(width = 30) #stringr::str_wrap adds line breaks given number of character per line wanted.

ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) +
  geom_point() + 
  annotate(
    geom = "label",
    x = 3.5,
    y = 38, 
    label = trend_text,
    hjust = "left",
    color = "red"
   ) +
  annotate(
    geom = "segment",
    x = 3.5,
    y = 35,
    xend = 5,
    yend = 25,
    color = "red",
    arrow = arrow(type = "closed")
  )
```

### Exercises

Ex1: Putting different labelsin each facet:
```{r}
#| label: Ex1 - different labels, different facets using geom_text()

Labs <- mpg |>
  mutate(
    labs = case_when(
      class == unique(class)[1] ~ "A",
      class == unique(class)[2] ~ "B",
      class == unique(class)[3] ~ "C",
      class == unique(class)[4] ~ "D",
      class == unique(class)[5] ~ "E",
      class == unique(class)[6] ~ "F",
      class == unique(class)[7] ~ "G",
    )
  ) |>
  distinct(class, labs)

print(Labs)

ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point() + 
  geom_text(
    data = Labs,
    mapping = aes(x = 6, y = 40, label = labs)
  ) + 
  facet_wrap(~class)
```
Ex2: What arguments to geom_label() control appearance of background box?

$\rightarrow$ See documentation under `?geom_label`, section "geom_label()": ...the `fill` aesthetic controls the background color of the label:

```{r}
#| label: Ex2 - geom_label()

ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point() + 
  geom_label(
    mapping = aes(x = 6, y = 40, label = "This is a text"),
    fill = "yellow"
  )

```

Ex3: What are the four arguments to `arrow()`?

```{r}
#| label: Ex3 - arrow() function

ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point() +
  annotate(
    geom = "segment",
    x = 3,
    y = 35,
    xend = 5, 
    yend = 25, 
    color = "red",
    arrow = arrow(angle = 90, # Angle of arrow head
                  length = unit(0.25, "cm"), # unit specifying length of head
                  ends = "both", # Which ends to draw ("first", "last")
                  type = "closed") # type of arrow head ("open")
  )

```

## Scales

Scales control how aesthetic mappings will be displayed

### Default scales

ggplot2 automatically adds scales:

```{r}
#| label: Example - Default scales

ggplot(
  data = mpg, 
  mapping = aes(x = displ, y = hwy)
) +
  geom_point(
    mapping = aes(color = class)
  )

# Behind the scenes
ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point(
    mapping = aes(color = class)
  ) +
  scale_x_continuous() + # puts values of displ on continuous line on x-axis
  scale_y_continuous() +
  scale_color_discrete() # chooses colors for each of the class of car
```
Scales are named according to type of variable:

+ continuous
+ discrete 
+ datetime 
+ date

Reasons to override defaults:

+ changing paramters of default scale, allowing e.g. to change breaks on axes, or key labels on the legend
+ replacing scales, and using different algorithm


### Axis ticks and legend keys

Axis and legends are calles __guides__. Axis are used for x/y aesthetics; legends for everything else.

Two arguments affecting appearance of ticks on axis and keys on legend: `breaks` and `labels`:

+ `break`: controls position of ticks; - values associated with keys
+ `labels`: controls text labels

```{r}
#| label: Example - common use of breaks

ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy, color = drv)
) + 
  geom_point() +
  scale_y_continuous(breaks = seq(from = 15, to = 40, by = 5))

```

Can use labels same way, or set `NULL` to suppress. For discrete scales for categorical variables, `labels` can be named list of existing level names and desired labels for them:

```{r}
#| label: Example - labels

# Example: Rescaling and relabeling x-axis
ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point() +
  scale_x_continuous(
    breaks = seq(from = 2, to = 7, length.out = 3),
    labels = c("A", "B", "C"))

# Example: labels = NULL
ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy, color = drv)
) + 
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL) +
  scale_color_discrete(labels = c(
    "4" = "4-wheel",
    "f" = "front",
    "r" = "rear"
  ))

```

:::{.callout-tip}
`labels` coupled with labelling functions from _scales package_ useful for formatting numbers as currency, percent, etc.
:::

```{r}
#| label: Example - formatting numbers, currencies, ect.

# Left
ggplot(
  data = diamonds,
  mapping = aes(x = price, y = cut)
) +
  geom_boxplot(alpha = 0.05) +
  scale_x_continuous(labels = label_dollar())

# Right
ggplot(
  data = diamonds,
  mapping = aes(x = price, y = cut)
) +
  geom_boxplot(alpha = 0.05) + 
  scale_x_continuous(
    labels = label_dollar(scale = 1/1000, # divide dollar values by 1000
                          suffix = "K"), # add suffix
    breaks = seq(from = 1000, to = 19000, by = 6000) # add custom breaks
  )
```

Another label function is `label_percent()`:

```{r}
#| label: Example - label_percent()

ggplot(
  data = diamonds,
  mapping = aes(x = cut, fill = clarity)
) +
   geom_bar(position = "fill") + # position = "fill" return proportions at every x-position; most useful for geoms anchored to x-axis, like geom_bar()
   scale_y_continuous(
    labels = label_percent(),
    name = "Percentage"
   )
```

Another use of `breaks` when having few data points, wanting to highlight where observations occur:

```{r}
#| label: Example - using breaks with few data points
# Presidential dataset
glimpse(presidential)

presidential |>
  mutate(id = 33 + row_number()) |> # add id corresponding to number of pres.
  ggplot(
    mapping = aes(x = start, y = id)
  ) +
  geom_point() +
  geom_segment(
    mapping = aes(xend = end, yend = id)
  ) +
  scale_x_date(name = NULL,
               breaks = presidential$start,
               date_labels = "'%y"
  )


```


### Legend layout

Controls overall position of the legend using `theme()` setting - they control non-data parts of the plot. Theme setting `legend.position` controls where the legend is drawn:

```{r}
#| label: Example - theme()

base <- ggplot(
  data = mpg, 
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point(
    mapping = aes(color = class)
  )

base + theme(legend.position = "right") # default
base + theme(legend.position = "left")
base +
  theme(legend.position = "top") +
  guides(color = guide_legend(nrow = 3))
base +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3))
```

:::{.callout-note}
`guides()` along with `guides_legend()` or `guide_colorbar()` controls the display of individual legends. `nrow`controls number of rows the legend uses; `override.aes` overrides one of the aesthetics, e.g., making points bigger. Useful when used a low `alpha`to display many points on a plot.
:::

```{r}
#| label: Example - controlling individual legends

ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point(
    mapping = aes(color = class)
  ) + 
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3, override.aes = list(size= 4)))
```

:::{.callout-note}
Argument in `guides()` matches name of aesthetic, just like in `labs()`.
:::


### Replacing a scale

Two types of scales likely to switch out: continuous position scales and color scales. 

```{r}
#| label: Example - replacing scales
# Example - log transforming to better see relationship
# Left
ggplot(
  data = diamonds,
  mapping = aes(x = carat, y = price)
) + 
  geom_bin2d()

# Right
# Note: scales are labelled with transformed values
ggplot(
  data = diamonds,
  mapping = aes(x = log10(carat), y = log10(price))
) + 
  geom_bin2d()

# Instead doing transformation in aesthetic mapping, transform with scales
# Note: Scales are labelled with original values
ggplot(
  data = diamonds,
  mapping = aes(x = carat, y = price)
) + 
  geom_bin2d() +
  scale_x_log10() +
  scale_y_log10()
```

Another scale frequently customized is color. Default categorical scale picks colors that are evenly spaced around the color wheel.

Useful alternative: ColorBrewer - tuned colors for color blindness.

```{r}
#| label: Example - categorical scale - color

ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point(
    mapping = aes(color = drv)
  )

# For people with red-green color blindness: 
ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) +
  geom_point(
    mapping = aes(color = drv, shape = drv)
  ) +
  scale_color_brewer(palette = "Set1") # See documentation for palettes
```

:::{.callout-tip}
For _ColorBrewer_ scales see online documentation at <https://colorbrewer2.org/>.
:::

For predefined mapping, use `scale_color_manual()`:

```{r}
#| label: Example - scale_color_manual()

presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(
    mapping = aes(x = start, y = id, color = party)
  ) +
  geom_point() +
  geom_segment(
    mapping = aes(yend = id, xend = end)
  ) +
  scale_y_continuous(
    name = "Name of president",
    breaks = seq(34, 45, by = 1),
    labels = presidential$name
    ) +
  scale_x_date(
    name = "Start of presidency",
    date_breaks = "4 years", # see documentation
    date_labels = "'%y"
    ) +
  scale_color_manual(
    values = c(Republican = "red", Democratic = "blue")
  ) +
  theme(
    legend.position = "top"
    ) +
  labs(
    title = "Presidential terms",
     color = "Party"
  )
```

:::{.callout-tip}
For continuous color, use `scale_color_gradient()` or `scale_fill_gradient()`. For diverging scale use `scale_color_gradient2()` - allows to give negative and positive values different color.
:::

One very useful color scale - viridis color scales:

```{r}
#| label: Viridis color scales

df <- tibble(
  x = rnorm(10000),
  y =  rnorm(10000)
)

ggplot(
  data = df,
  mapping = aes(x = x, y = y)
) +
  geom_hex() +
  coord_fixed() +
  labs(
    title = "Default, continous",
    x = NULL, 
    y = NULL
  )

ggplot(
  data = df, 
  mapping = aes(x = x, y = y)
) + 
  geom_hex() + 
  coord_fixed() +
  scale_fill_viridis_c() +
  labs(
    title = "Viridius, continuous",
    x = NULL, 
    y = NULL
  )

ggplot(
  data = df, 
  mapping = aes(x = x, y = y)
) + 
  geom_hex() +
  coord_fixed() +
  scale_fill_viridis_b() +
  labs(
    title = "Viridius, binned",
    x = NULL,
    y = NULL
  )

ggplot(df, aes(x, y)) +
  geom_hex() +
  scale_fill_gradient(low = "white", high = "red") +
  coord_fixed()
```

:::{.callout-note}
Color scales come in two variables: `scale_color_*()` and `scale_fill_*()`.
:::

### Zooming

Three ways to control plot limits:.
1. Adjust what data are plotted.
2. Set limits in each scale.
3. Set `xlim` and `ylim` in `coord_cartesian()`

```{r}
#| label: Example - Zooming
# Left
ggplot(
  data = mpg, 
  mapping = aes(x = displ, y = hwy)
) +
  geom_point(
    mapping = aes(color = drv)
  ) +
  geom_smooth()

# Right
mpg |> 
  filter(displ >= 5 & displ <= 6 & hwy >= 10 & hwy <= 25) |>
  ggplot(
    mapping = aes(x = displ, y = hwy)
  ) +
  geom_point(
    mapping = aes(color = drv)
  ) + 
  geom_smooth() + 
  scale_color_discrete(
    labels = c(
      "4" = "4-wheel",
      "f" = " front",
      "r" = "rear"
    )
  )
  
# compare 'Right' to: 
ggplot(
  data = mpg, 
  mapping = aes(x = displ, y = hwy)
) +
  geom_point(
    mapping = aes(color = drv)
  ) +
  geom_smooth() +
  coord_cartesian(xlim = c(5, 6), ylim = c(10, 25))

```


## Themes

Customizing non-data elements of the plot with a theme:

```{r}
#| label: Example - themes

ggplot(
  data = mpg, 
  mapping = aes(x = displ, y = hwy)
) +
  geom_point(
    mapping = aes(color = drv)
  ) +
  geom_smooth(se = FALSE) +
  theme_bw()
```

:::{.callout-note}
ggplot includes 12 themes: `*_bw()`, `*_light()`, `*_classic()`, `*_linedraw()`, `*_dark()`, `*_minimal()`, `*_gray()`, `*_void()`;
Many more in add-on packages like __ggthemes__.
:::

```{r}
#| label: Example - customizing themes

ggplot(
  data = mpg, 
  mapping = aes(x= displ, y = hwy, color = drv)
) +
  geom_point() +
  labs(
    title = "Larger engine sizes tend to have lower fuel economy",
    caption = "Source: https://fueleconomy.gov."
  ) +
  theme(
    legend.position = c(0.6, 0.7), # position of legend in plot
    legend.direction = "horizontal", # layout of items in legend
    legend.box.background = element_rect(color = "red"),
    plot.title = element_text(face = "bold", hjust = 1),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0),
    axis.title = element_text(face = "bold", color = "blue")
  )
```


## Layout

__Patchwork__ package allows combining separate plots into same graphic:

```{r}
#| label: Example - Plotting two plots

p1 <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  labs(title = "Plot 1")
p2 <- ggplot(mpg, aes(x = drv, y = hwy)) +
  geom_boxplot() +
  labs(title = "Plot 2")

p1 + p2 # First create and save both plots, then place next to each other with +

# Or more complex plot layout:
p3 <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point() +
  labs(title = "Plot 3")

(p1 | p3) / p2
```

Patchwork allows collecting legends from multiple plots into one common legend - customizing placement, dimension, add title, subtitle, caption, etc. to the plots $\rightarrow$ look through the guides on the package website: https://patchwork.data-imaginist.com.




