---
title: "Facets"
subtitle: "R for Data Science (2e), Chapter 9.5"
author: "Filip Kukic"
toc: true
theme: darkly
format: html
---

## Faceting with facet_wrap()

Splitting plots into subplots, each displaying one subset of the data.

::: {.callout-note}
Use categorical variable!
:::

### Prerequisites

```{r}
#| label: load-packages
#| warning: false

library(tidyverse)
library(ggthemes)

```

### Dataset

```{r}
#| label: mpg dataset
mpg
glimpse(mpg)
# ?mpg, to get access to variables used in ds
# str(mpg), to access structure of object
```

### Example

```{r}
#| label: Example 1
ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy) ) + 
  geom_point() + 
  facet_wrap(~cyl)

```

### Faceting plot with combination of two variables, use facet_grid()

```{r}
#| label: facet_grid()- two variables

ggplot(data = mpg,
       mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv ~ cyl) # double sided formula: rows ~ cols

ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy) ) + 
  geom_point(mapping = aes(color = class)) + 
  scale_color_colorblind() + # by default, each facet shares same scale
  # set scales = "free", to allow different scales for both axis # other option: "free_x", "free_y" 
  facet_grid(drv ~ cyl, scales = "free")
  
  
```

### Exercises

#### Q1: What happens if we facet on continuous variable?

```{r}
#| label: Q1

ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_wrap(~displ)  

```

A: Creates facet for every unique value of `displ`.

#### Q2: What do empty cells in the plot mean with facet_grid(drv ~ cyl)

```{r}
#| label: Q2

ggplot(
  data = mpg,
  mapping = aes(x = drv, y = cyl)) + 
  geom_point() + # compare to: ggplot( data = mpg, mapping = aes(x = displ, y = hwy))
  geom_point() + facet_grid(drv ~ cyl) # 

```

A: No corresponding values for some combinations.

#### Q3: What plots does following code make? What does . do?

```{r}
#| label: Q3

ggplot( data = mpg, mapping = aes(x = displ, y = hwy) ) + geom_point() + facet_grid(drv ~ .)

ggplot( data = mpg ) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_grid(. ~ cyl)
```

A: See ?facet_grid,'. dot in the formula used to indicate there should be no faceting on this dimension (either row or column)'

#### Q4: Faceting vs color aesthetic; Advantages? Larger data set influence?

```{r}
#|label: Q4

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~cyl, nrow = 2)

```

A: Faceting challenging with large data set; Distinguish characteristics through colors might work better.

#### Q6: Which plot easier for comparing engine size (displ)?

```{r}


ggplot(
  data = mpg,
  mapping = aes(x = displ) ) + 
  geom_histogram() + 
  facet_grid(drv ~.) # => best representation of engine size, when faceting rowwise 

ggplot(data = mpg,
       mapping = aes(x = displ) ) + 
  geom_histogram() + 
  facet_grid(.~ drv)

# Note: previous plot equivalent to see below 
ggplot(data = mpg,
       mapping = aes(x = displ) ) + 
  geom_histogram() + 
  facet_wrap(~drv)

```

#### Q7: Recreating plot using facet_wrap() instead of facet_grid().

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~drv, nrow = 3)
```
