---
title: "Position adjustment"
subtitle: "R for Data Science (2e), Chapter 9.6"
author: "Filip Kukic"
toc: true
theme: darkly
format: html
---

### Prerequisites

```{r} 
#| label: load-packages
#| warning: false

library(tidyverse)
library(ggthemes)

```

### MPG Dataset

```{r}
#| label: dataset
#| output: false

# ?mpg
# str(mpg)
mpg
glimpse(mpg)

```

### Additional visualization tools for bar charts

One can color a bar chart using the `color` or `fill` aesthetic: 

```{r}
#| label: Example color/fill aesthetic

ggplot(
  data = mpg, 
  mapping = aes(x = drv, color = drv)
) + 
  geom_bar()

ggplot(
  data = mpg,
  mapping = aes(x = drv, fill = drv)
) + 
  geom_bar()

```

::: {.callout-note}
Mapping the fill aesthetic to another variable, like `class` will result in bars automatically being stacked. Each colored rectangle represents a combination of `drv` and `class`.
:::

```{r}
#| label: Mapping fill to another variable

ggplot(
  data = mpg,
  mapping = aes(x = drv, fill = class)
) +
  geom_bar()

```

Stacking is performed using `position` argument (if you look up `?geom_bar`, youl will see that the default is set to `position = "stack"`).

### Other position options
+ `"identity"`: Places each object exactly where it falls in context of the graph - resulting in overlapping

```{r}
#| label: position - identity
# Left
ggplot(
  data = mpg,
  mapping = aes(x = drv, fill = class)
) +
  geom_bar(position = "identity", alpha = 1/5) # need to specify transparency with alpha or alternatively use fill = NA
# Right´
ggplot(
  data = mpg,
  mapping = aes(x = drv, color = class)
) + 
  geom_bar(fill = NA, position = "identity")
```

+ `"dodge"`: Places overlapping object direclty besides on another

```{r}
#| label: position - dodge

ggplot(
  data = mpg, 
  mapping = aes(x = drv, fill = class)
) +
  geom_bar(position = "dodge")

```

+ `"jitter"`: Very useful for scatterplots with lots of overlapping (overplotting)

::: {.callout-note}
ggplot2 has shorthand for `geom_point(position = "jitter"): geom_jitter()``
:::

```{r}
#| label: position - jitter

ggplot(
  data = mpg, 
  mapping = aes(x = displ, y = hwy)
) + 
  geom_point(position = "jitter")
```

## Miscellaneous

Comparing `geom_jitter` with `geom_count`:

```{r}
#| label: geom_count()

ggplot(
  data = mpg, 
  mapping = aes(x = displ, y = hwy)
) + 
  geom_count(position = "jitter")

```

Default position for `geom_boxplot`:

```{r}
#| label: Default position adjustement for boxplots

# Right
ggplot(
  data = mpg,
  mapping = aes(x = class, y = hwy)) +
  geom_boxplot()

# Left
ggplot(data = mpg,
       mapping = aes(x = class, y = hwy)) +
  geom_boxplot(position = "dodge2")

```

