---
title: "Statistical Transformation"
subtitle: "R for Data Science (2e), Chapter 9.5"
author: "Filip Kukic"
toc: true
theme: darkly
format: html
---

### Prerequisites

```{r} 
#| label: load-packages
#| warning: false

library(tidyverse)
library(ggthemes)

```

### Diamonds Dataset

The `diamonds` data set contains information on ~54,000 diamonds, including price, carat, color, clarity, and cut of each diamond.
```{r}
#| label: dataset
#| output: false

# ?diamonds
# str(diamonds)
diamonds
glimpse(diamonds)

```

Chart shows, that more diamonds are available with high quality cuts than low quality cuts:

```{r}
#| label: bar-plot diamond cuts

ggplot(
  data = diamonds,
  mapping = aes(x = cut)
) + 
  geom_bar()
```

### What is a stat?

::: {.callout-note}
On the y-axis, count is displayed - __NOT__ a variable in `diamonds`.
Many graphs, like scatterplots, plot raw values of dataset. Other graphs, like bar charts, calculate new values to the plot:

+ __Bar charts__, __histograms__, __frequency polygons__ bin data and then plot bin counts - i.e. the number of points falling in each bin.

+ __Smoothers__ fit a model to data and then plot predictions from the model.

+ __Boxplots__ compute the five-number summary of the distribution and then display that summary
:::

Algorithm used to calculate new values is called __stat__ - i.e. statistical transformation.


### How to find out which stat a geom uses?

Inspect default value for `stat` argument with `?geom_`.
E.g. `geom_bar()` uses `stat_count()` - computes two new variables: _count_ and _prop_.

Every geom has default stat; every stat has default geom.


### Why use stat explicitely then?

1. Might want to override default stat, e.g.:
```{r}
#| label: Example 1 of overriding default stat

diamonds |> 
  count(cut) |>
  ggplot(mapping = aes(x = cut, y = n)) +
  geom_bar(stat = "identity") # leaves data as is
```

2. Want to override the default mapping from transformed variables in aesthetics, e.g.:
```{r}
#| label: Example 2 of overriding default stat

ggplot(
  data = diamonds, 
  mapping = aes(x = cut, y = after_stat(prop), group = 1)) +
    geom_bar()


ggplot(
  data = diamonds, 
  mapping = aes(x = cut, y = after_stat(count))) +
    geom_bar()

```

::: {.callout-note}
Need to specify group argument, otherwise ggplot computes proportions within each x category, so every bar equals 1. With group = 1, ggplot computes proportions across the entire data set, so the bars represent the fraction of all diamonds in each category.
:::

3. Want to draw greater attention to statistical transformation in code, e.g.:
```{r}
#| label: Example 3 of overriding default stat

ggplot(data = diamonds) +
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )

```

```{r}
#| label: Example 3 - Using default geom_ command

diamonds_summary <-
  diamonds |>
  summarize(
    ymin = min(depth),
    y = median(depth),
    ymax = max(depth),
    .by = cut
  )

ggplot(
  data = diamonds_summary,
  mapping = aes(x = cut,
                y = y,
                ymin = ymin,
                ymax = ymax)) +
  geom_pointrange()
```

## Miscellaneous

Difference between `geom_bar()` and `geom_col()`:
```{r}
#| label: geom_bar() vs. geom_col()

ggplot(diamonds, aes(x = cut)) +
  geom_bar() # here we use default stat_count()

ggplot(diamonds, mapping = aes(x = cut, y = depth)) + 
  geom_col() # here geom_bar() stacks the "depth" variable

diamonds |> 
  summarize(mean_depth = mean(depth, na.rm = TRUE), .by = cut) |>
  ggplot(mapping = aes(x = cut, y = mean_depth)) + 
  geom_col() # here we compute the average depth and then plot using geom_col()
```
::: {.callout-important}
__Documentation:__ `geom_bar()` makes the height of the bar proportional to the number of cases in each group. If you want the heights of the bars to represent values in the data, use `geom_col()` instead. `geom_bar()` uses `stat_count()` by default: it counts the number of cases at each x position. `geom_col()` uses `stat_identity()`: it leaves the data as is.
:::