---
title: "factors"
subtitle: "R for Data Science (2e), Chapter 16"
toc: true
theme: darkly
format: html
---

:::{.callout-caution}
Factors are used for categorical variables - known set of possible values. 

Also useful when displaying character vectors in non-alphabetical order.
:::

### Prerequisites
```{r}
#| label: Prerequisites
# Base R and forcats package included in tidyverse will be used
library(tidyverse)
library(patchwork)
```


## Factor basics
```{r}
#| label: Basic factors

x1 <- c("Dec", "Apr", "Jan", "Mar")

# Issues: Only twelve possible months and danger of typos; doesn't sort in useful way

x2 <- c("Dec", "Apr", "Jam", "Mar")
sort(x1)

# Solution: Factors
# create list of valid levels

month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
```

```{r}

# Specifiying Levels
y <- factor(x = x1) # omitting levels results in levels being taken from data in alphabetical order
y
sort(y)

# Not specifying levels
y1 <- factor(x = x1, levels = month_levels) 
y1
sort(y1) # with levels added, can ensure correct alphabetical order
```

```{r}

# What happens to typos after adding levels?
y2 <- factor(x = x2, levels = month_levels)
y2 # can see that typo will be spottet by predefining levels
sort(y2) # NA will be omitted in sorting
```

```{r}

# Can use forcats::fct() - will not convert faulty values silently to NA
# y3 <- fct(x2, levels = month_levels)

# forcats::fct orders by first appearance:
fct(x1)


# Can access levels via levels():
levels(y2)
```


Can also create factors when reading in data with readr with `col_factor()`:
```{r}
#| label: col_factor()

csv <- c("
month,value
Jan,12
Feb,56
Mar,12"
)

df <- read_csv(csv, col_types = cols(month = col_factor(month_levels)))
# See ?read_csv <- cols() create columns specification; must contain column specification

df$month

```



## General Social Survey

```{r}
#| label: ?gss_cat
# Data set
gss_cat

# For tibble, see levels of factors with count()
gss_cat |>
  count(race)

```

Two most common operations when working with factors: 
+ Changing order of the levels
+ Changing values of the levels


## Modifying factor order

Often useful to change the order of the factor level in a visualization:
```{r}
#| label: Changing order of factors

# Example: Exploring average number of hours spend watching TV per day across religions

relig_summary <- gss_cat |>
  group_by(relig) |>
  summarize(
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(
    data = relig_summary,
    mapping = aes(x = tvhours, y = relig)
  ) +
  geom_point()
# Hard reading this plot due to no overall pattern

# Can use fct_reorder()
# Three arguments:
# .f -> factor (levels) to be modified; 
# .x -> numeric vector used to reorder levels; 
# Optionally: .fun -> function used if multiple values of .x for each value of .f; default is median

ggplot(
  data = relig_summary,
  mapping = aes(x = tvhours, y = fct_reorder(relig, tvhours))
) +
  geom_point()

# Recommendation: Move out of aes into mutate()
# relig_summary1 <- relig_summary |>
#  mutate(
#    relig = fct_reorder(relig, tvhours)
#  ) 

relig_summary |>
  mutate(
    relig = fct_reorder(relig, tvhours)
  ) |>
  ggplot(
    mapping = aes(x = tvhours, y = relig)
  ) +
  geom_point()

# Check first: factor(relig_summary$relig), note the order of levels
# Then check:  factor(relig_summary1$relig), notice reordering of levels in correspondance to values of tvhours

```

```{r}
#| label: Example - fct_reporder() 

# Example fct_reorder()
# Average age varying across reported inc
rincome_summary <- gss_cat |>
  group_by(rincome) |>
  summarize(
    avg_age = mean(age, na.rm = TRUE),
    n = n()
  )

ggplot(
  data = rincome_summary,
  mapping = aes(x = avg_age, y = rincome) 
) +
  geom_point()


ggplot(
  data = rincome_summary,
  mapping = aes(x = avg_age, y = fct_reorder(rincome, avg_age)) 
) +
  geom_point()

ggplot(
  data = rincome_summary,
  mapping = aes(x = avg_age, y = fct_relevel(rincome, "Not applicable") # can also be character string of levels
  )
) +
  geom_point()

```

`fct_reorder()` mostly makes sense for arbitrarily ordered levels of factors. However, can pull "Not applicable" to the other special levels to the front using `fct_relevel()`: Takes factor `.f` and any number of levels that we want to move to the front of the line (see `?fct_relevel```: ... = function or character levels, default is after = 0...any levels not mentioned will be left in existing order).

Another type of ordering useful when coloring the lines on a plot. `fct_reorder2(.f, .x, .y)`: reorders factor `.f` by `.y` values associated with largest `.x` values.
```{r}
#| label: fct_reorder2(.f, .x, .y)
by_age <- gss_cat |>
  filter(!is.na(age)) |> 
  count(age, marital) |>
  group_by(age) |>
  mutate(prop = n / sum(n))

by_age_plot <- ggplot(
  data = by_age,
  mapping = aes(x = age, y = prop, color = marital)
) + 
  geom_line(linewidth = 1) +
  scale_color_brewer(palette = "Set1")

by_age_plot_fctreord <- ggplot(
  data = by_age,
  mapping = aes(x = age, y = prop, color = fct_reorder2(marital, prop, age))
) + 
  geom_line(linewidth = 1) +
  scale_color_brewer(palette = "Set1")

by_age_plot + by_age_plot_fctreord

```

For bar plots can `fct_infreq()` to order levels in decreasing frequency (or use `fct_rev()` for reversed ordering):
```{r}
#| label: fct_infreq() and fct_rev()

gss_cat |>
  mutate(marital = marital |> fct_infreq() |> fct_rev()) |>
  ggplot(aes(x = marital)) + 
  geom_bar()
```


## Modifying factor levels

Most common tool is `fct_recode()` - allows changing value for each level:
```{r}
#| label: fct_recode()

gss_cat |>
  count(partyid)

# See ?fct_recode: Levels not mentioned, will be left as is;
# Levels can be removed by aneming them NULL
gss_cat |>
  mutate(
    partyid = fct_recode(partyid,
      "Republican, strong"    = "Strong republican",
      "Republican, weak"      = "Not str republican",
      "Independent, near rep" = "Ind,near rep",
      "Independent, near dem" = "Ind,near dem",
      "Democrat, weak"        = "Not str democrat",
      "Democrat, strong"      = "Strong democrat"
    )
  ) |>
  count(partyid)

# Can also combine groups:
gss_cat |>
  mutate(
    partyid = fct_recode(partyid,
      "Republican, strong"    = "Strong republican",
      "Republican, weak"      = "Not str republican",
      "Independent, near rep" = "Ind,near rep",
      "Independent, near dem" = "Ind,near dem",
      "Democrat, weak"        = "Not str democrat",
      "Democrat, strong"      = "Strong democrat",
      "Other"                 = "No answer",
      "Other"                 = "Don't know",
      "Other"                 = "Other party"
    )
  )
```


Can also collapse a lot of level using `fct_collapse()` - useful variant of `fct_recode()`:
```{r}
#| label: fct_collapse()

gss_cat |>
  mutate(
    partyid = fct_collapse(partyid,
      "other" = c("No answer", "Don't know", "Other party"),
      "rep" = c("Strong republican", "Not str republican"),
      "ind" = c("Ind,near rep", "Independent", "Ind,near dem"),
      "dem" = c("Not str democrat", "Strong democrat")
    )
  ) |>
  count(partyid)

```

Can also use `fct_lump_*()` - lumping together small groups. E.g., `fct_lump_lowfreq()` lumps smallest group categories into "Other, anmd keeping it the smallest catgeory:
```{r}
#| label: fct_lum_*()

gss_cat |>
  mutate(relig = fct_lump_lowfreq(relig)) |> count(relig)

# Can specify lump-groups by fct_lump_n():

gss_cat |>
  mutate(relig = fct_lump_n(relig, n = 10)) |> count(relig)

```

```{r}
#| label: Exercises

# Exercise 1: Change of proportion of Democrats, Republican, Independent changes over year.

Ex1 <- gss_cat |>
  mutate(
         partyid = fct_collapse(partyid,
          "other" = c("No answer", "Don't know", "Other party"),
          "rep" = c("Strong republican", "Not str republican"),
          "ind" = c("Ind,near rep", "Independent", "Ind,near dem"),
         "dem" = c("Not str democrat", "Strong democrat") 
         )
  ) |> 
  count(partyid, year) |>
  mutate(prop = n / sum(n), .by = year) |>
  ggplot(
    aes(x = year, y = prop, color = fct_reorder2(partyid, year, prop))
  ) + geom_line(linewidth = 1)
   
# Exercise 2: How can we collapse rincome into small sets of categories?
# Note: Approach "fast and easy", though fct_collapse can be used for manually adjsuting the levels
Ex2 <- gss_cat |>
    mutate(
        rincome = fct_lump_n(rincome, n = 4) # Keeps 4 most frequent groups, creates 5th group by default for "others", thus 5 groups in total
    ) |>
    count(rincome) |>
    mutate(
        prop = n / sum(n)
    ) |> 
    ggplot(aes(x = prop, y = rincome)) +
    geom_point()

Ex1 / Ex2 # patchwork package
```


## Ordered factors

Can be created using `ordered()` function:
```{r}
#| label: Idemtifying ordered factors

# Can identify ordered factors via ordering of levels in output

ordered(c("a", "b", "c"))


```

Different behavior in two places, compared to regular factors: 
+ mapping ordered factor to color in ggplot2 defaults to `scale_color_viridis()` - color scale implying ranking
+ in linear models, when using ordered predictor, will use "polynomial contrasts". 

