---
title: "Regular expressions"
subtitle: "R for Data Science (2e), Chapter 15"
toc: true
theme: darkly
format: html
---

## Introduction

__regular expressions__ $\rightarrow$ language for describing patterns within strings.

### Prerequisites
```{r}
#| label: Prerequisites

library(tidyverse)
library(babynames)

# plus character vectors from stringr
```

## Pattern basics

`str_view()` for regex patterns underlying mechanism - will show for regex arguments elements of string that match, surrounding each match with <>. 

```{r}
#| label: Example regex and str_view()

str_view(fruit, pattern = "berry")

str_view(fruit, pattern = "berry", match = TRUE) # Default

# Show all that don't match pattern -> See documentation ?str_view
str_view(fruit, pattern = "berry", match = FALSE)
```

:::{.callout-note}
Letters and number are called __literal characters__ - they match exactly.
Punctuation characters `.`,`+`, `*`, `[`, `]`, are __metacharacters__ - have special meaning.
:::

E.g., `.` will match any character - `a.` matches string containing "a" followed by any character:
```{r}
#| label: Metacharacters

str_view(c("a", "ab", "ae", "bd", "ea", "eab"), pattern = "a.")

str_view(fruit, pattern = "a...e")
```

Quantifiers control the times a pattern can match:
+ `?` makes pattern optional - either matches 0 or 1 time
+ `+` lets pattern repeat - matches at least once
+ `*` lets pattern be optional or repeat
```{r}
#| label: Quantifiers
# ? matches "a" optionally followed by "b"
str_view(c("a", "ab", "abb"), pattern = "ab?")

# + matches "a" followed by at least one "b"
str_view(c("a", "ab", "abb"), pattern = "ab+")

# * matches "a" followed by any number of b's
str_view(c("a", "ab", "abb"), pattern = "ab*")

```

Character classes defined by `[]` - matches set of characters, e.g., [abcd] matches "a", "b", "c", or "d" - ^[abdc] matches anything except "a", "b", "c", or "d":
```{r}
#| label: Character classes for matching

# Example: Finding words where x is surrounded by vowels

str_view(words, pattern = "[aeiou]x[aeiou]")

```

Can also use __alternations__ `|` - choosing one or more alternative patterns:
```{r}
#| label: Alternations

str_view(fruit, pattern = "apple|melon|nut")

str_view(words, pattern = "aa|ee|ii|oo|uu")
```


## Key functions

`str_detect()` returns logical vector - `TRUE` pattern matches element or character vector, `FALSE` otherwise. Pairs well with `filter()` and `summarize()`:
```{r}
#| label: str_detect()

str_detect(c("a", "b", "c"), pattern = "[aeiou]")

# filter()
babynames |>
  filter(str_detect(name, pattern = "M.")) |>
  count(name, wt = n, sort = TRUE)

# summarize()
babynames |>
  group_by(year) |>
  summarize(prop_x = mean(str_detect(name, pattern = "x"))) |>
  ggplot(mapping = aes(x = year, y = prop_x)) +
  geom_line()

```

Closely related functions: 

`str_subset()` - returns character vector containing only strings matching. See ?str_subset: It's a wrapper around `x[str_detect(x, pattern)]` (equivalent to grep(pattern, x, value = TRUE))

`str_which()` - return indices of string where there's at least one match (see ?strt_which) - wrapper around `which(str_detect(x, pattern))` (equivalent to grep(pattern, x))


### Count matches

`str_count()` - tells how many matches are in each string:
```{r}
#| label: str_count()
# Example 1:
x <- c("apple", "banane", "pear")

str_count(x, pattern = "p")

# Example 2:
str_view("abababa", pattern = "aba")

str_count("abababa", pattern = "aba")
```

Use `str_count()` with `mutate()`:
```{r}
#| label: mutate() and str_count()

babynames |>
  count(name) |>
  mutate(
    vowels = str_count(name, pattern = "[aeiou]"),
    consonants = str_count(name, pattern = "[^aeiou]")
  )
  
# Case sensitivity
  
babynames |>
  count(name) |>
  mutate(
    vowels = str_count(name, pattern = "[aeiouAEIOU]"),
    consonants = str_count(name, pattern = "[^aeiouAEIOU]")
  )
  # Or alternatively
  # str_count(name, regex([aeiou"], ignore.case = TRUE))
  # str_count(str_to_lower(name), "[aeiou]")

babynames |>
  count(name) |>
  mutate(
    name = str_to_lower(name),
    vowels = str_count(name, pattern = "[aeiou]"),
    consonants = str_count(name, pattern = "[^aeiou]")
  )
```


### Replace values

`str_replace()` - replaces first match 

`str_replace_all()` - replaces all matches

Can be paired with `mutate()` function.
```{r}
#| label: str_replace() and str_replace_all()

x <- c("apple", "pear", "banana")

str_replace_all(x, pattern = "[aeiou]", replacement = "")

str_replace_all(x, pattern = "[aeiou]", replacement = "_")

# Tricky example: Replacing forward-slah with back-slash
str_view(str_replace_all("a/b/c/d/e", pattern = "/", replacement = "\\\\"))

# Alternative to str_to_lower():
x <- c("Apple", "Banana", "Berry")
str_replace_all(x , pattern = "[A-Z]", replacement = tolower)

```


### Extract variables

Functions extracting data out of one column into one or more new columns: `separate_wider_regex()`:
```{r}
#| label: Extracting variables 

df <- tribble(
  ~str,
  "<Sheryl>-F_34",
  "<Kisha>-F_45", 
  "<Brandon>-N_33",
  "<Sharon>-F_38", 
  "<Penny>-F_58",
  "<Justin>-M_41", 
  "<Patricia>-F_84", 
)

df |>
  separate_wider_regex(cols = str,
                       patterns = c("<", name = "[A-Za-z]+",
                                    ">-", gender = ".",
                                    "_", age = "[0-9]+"),
                       too_few = "debug"
  )

```


## Pattern details

### Escaping

Matching matcharacters, like `.`, requires escaping. Regexps use backslash for escaping - so need to `\.` - BUT `\` also special character, thus `\\.` is required.
```{r}
#| label: Escaping

dot <- "\\."

str_view(dot)

str_view(c("abc", "a.c", "bef"), pattern = "\\.")

str_view(c("abc", "a.c", "bef"), pattern = "a\\.c")

# To match literal \ need "\\\\"

x <- "a\\b"

str_view(x)

str_view(x, pattern = "\\\\")

# Can use raw strings alternatively:

str_view(x, pattern = r"{\\}")

# Matching literal ., $, |, *, +, ?, {, }, (, ) via []:

str_view(c("abc", "a.b", "a*c", "a c"), pattern = "[.]")

str_view(c("abc", "a.b", "a*c", "a c"), pattern = "[*]c")

```


### Anchors

Use __anchors__ to match at start or end of string - use `^` for start, use `$` for end.
```{r}
#| label: anchors

str_view(fruit, pattern = "^a")

str_view(fruit, pattern = "a$")

str_view(fruit, pattern = "^apple$")

```

Can also match boundaries between words:
```{r}
#| label: matching boundaries

x <- c("summary(x)", "summarize(df)", "rowsum(x)", "sum(x)")

str_view(x, pattern = "sum")

str_view(x, pattern = "\\bsum\\b")

# produce alone a zero-width match
str_view("abc", pattern = c("^", "$", "\\b"))

str_replace_all("abc", pattern = c("^", "$", "\\b"), replacement = "__")

```


### Character classes

Character class, character set - allows matching any character in a set. Can construct sets via `[]`:
+ `[^abc]`: Matches any character except "a", "b", "c".
+ `[a-z]`: matches any lower case letter
+ `[0-9]`: matches any number
+ `\`: escapes special characters, e.g., [\^\-\]] matches ^, -, ]
```{r}
#| label: Character classes

x <- "abc ABCD 12345 -!@#%."

str_view(x, pattern = "[abc]")
str_view(x, pattern = "[a-z]")

str_view(x, pattern = "[abc]+")

str_view(x, pattern = "[^a-z0-9]+")

# You need an escape to match characters that are special inside []
str_view("a-b-c", pattern = "[a-c]")

str_view("a-b-c", pattern = "[a\\-c]")

str_view(x, pattern = "\\w+")

str_view(x, pattern = "\\d+")
```

Shortcut:
+ `\d`: matches any digit
+ `\D`: matches anything other than digit
+ `\s`: matches whitespace
+ `\S`: matches anything other than whitespace
+ `\w`: matches any "word" character
+ `\W`: matches any "non-word" character


### Quantifiers

Control how many times pattern matches:
+ `?`: 0 or 1 matches
+ `+`: 1 or more matches
+ `*`: 0 or more matches
+ `{n}`: matches exactly n times
+ `{n,}`: matches at least n times
+ `{n,m}`: matches between n and m times


### Operator precedence and parentheses

Quantifiers have high precedence and alternation has low precedence which means that ab+ is equivalent to `a(b+)`, and `^a|b$` is equivalent to `(^a)|(b$)`. Just like with algebra, you can use parentheses to override the usual order.


### Grouping and capturing

Capturing groups allowing to use sub-components of the match via __back reference__: `\1` refers to match contained in first parenthesis, `\2` in the second parenthesis, and so on.
```{r}
#| label: Capturing groups

# Finding repeated pair of letters
str_view(fruit, pattern = "(..)\\1")

# Finding word starting and ending with same pair of letters
str_view(words, pattern = "^(..).*\\1$")

# Example: Switching first word with second word
sentences |>
  str_replace(pattern = "(\\w+) (\\w+) (\\w+)", replacement = "\\1 \\3 \\2")

```

Can extract matches for each group via `str_match()`:
```{r}
#| label: str_match()

str_view(fruit, pattern = "(..)\\1")



sentences |>
  str_match(pattern = "the (\\w+) (\\w+)") # Note: returns a matrix

# Easier when converted to tibble
sentences |> 
  str_match("the (\\w+) (\\w+)") |> 
  as_tibble(.name_repair = "minimal") |> 
  set_names("match", "word1", "word2")

```



### Regex flags

Settings to control details of regexp - called __flags__ - i.e., wrapping pattern in a call to `regex()`. Most useful flag is `ìgnore_case = TRUE` - allowing matching upper- or lowercase forms:
```{r}
#| label: flags

bananas <- c("banana", "Banana", "BANANA")
str_view(bananas)

str_view(bananas, pattern = regex("banana", ignore_case = TRUE))

```

+ `dotall = TRUE` lets `.` match everything including `\n`:
```{r}
#| label: dotall = TRUE

x <- "Line 1\nLine 2\nLine 3"

str_view(x)

str_view(x, pattern = ".Line")

str_view(x, pattern = regex(pattern = ".Line", dotall = TRUE))
```

+ `multiline = TRUE` makes `^` and `$` match start and end of each line rather than start and end of complete string:
```{r}
#| label: multiline = TRUE

x <- "Line 1\nLine 2\nLine 3"
str_view(x, "^Line")

str_view(x, regex("^Line", multiline = TRUE))

```



### Fixed matches

Can opt-out of using regular expression rules using `fixed()`:
```{r}
#| label: fixed()

str_view(c("","a", "."), pattern = fixed("."))

str_view(c("x", "X"), pattern = fixed("x", ignore_case = TRUE))
```


## Practice

1. check work by creating simple positive and negative controls. 
2. combine regexps with Boolean algebra
3. Create complex patterns using string manipulation

### Check work
```{r}
#| label: Example - checking work

str_view(sentences, pattern = "^The")
# matches all patterns
str_view(sentences, pattern = "The\\b")
# now "e" is last letter

# Finding all sentences beginning with pronoun
str_view(sentences, pattern = "^She|He|It|They\\b")
# But sgsin finds all pattern see 118
# Solution: anchor
str_view(sentences, pattern = "^(She|He|It|They)\\b")

# Spotting these mistakes by testing own pattern

pos <- c("He is a boy", "She had a good time")
neg <- c("Shells come from the sea", "Hadley said 'It's a great day'")

pattern <- "^(She|He|It|They)\\b"
str_detect(pos, pattern)

str_detect(neg, pattern)


```


### Boolean operations

```{r}
#| label: Boolean operations

str_view(words, pattern =  "^[^aeiou]+$")

# Alternatively
str_view(words[!str_detect(words, pattern = "[aeiou]")])
```
&\rightarrow$ Useful when dealing with logial combinations like "and", "or".

```{r}
#| label: Boolean operations - 1

str_view(words, pattern = "a.*b")

# Alternatively:

words[str_detect(words, pattern = "a") & str_detect(words, pattern = "b")]

# Word that contains all vowels?

words[
  str_detect(words, "a") &
  str_detect(words, "e") &
  str_detect(words, "i") &
  str_detect(words, "o") &
  str_detect(words, "u")
]


```


### Creating pattern with code

```{r}
#| label: creating pattern with code

# Find all sentences metioning a color
str_view(sentences, pattern = "\\b(red|green|blue)\\b")

# How to use many colors
cols <- colors()
cols <- cols[!str_detect(cols, pattern = "\\d")] # eliminate numbered variants
pattern <- str_c("\\b(", str_flatten(cols, collapse = "|"), ")\\b")
str_view(sentences, pattern = pattern)

```


## Regular expressions in other places

### tidyverse

Other useful functions
+ `matches(pattern)` - selects all variables whose name matches pattern - can use in `select()`, `rename_with()` and `across()`

+ `pivot_longer()` and `separate_wider_regex()` has `names_pattern` argument - takes vector of regular expressions - useul when extracting data out of variable names with complex structure.

+ `separate_longer_delim()` and `separate_wider_delim()` match fixed string - can use `regex()` to match a pattern - useful when trying to match comma optionally followed by space, i.e., regex(", ?").

### Base R

`apropos(pattern)` searches  all objects from global environment matching given pattern - useful when looking for function
```{r}
#| label: Using apropos()

apropos("replace")

list.files(pattern = "\\.qmd$")
```

